---
title: "GlobalGeoTree Germany, Berlin and NRW visualisation report"
author: "GlobalGeoTree team"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: false
execute:
  echo: false
  warning: false
  message: false
---

## Purpose

This is one cleaned and unified QMD for the Germany, Berlin, and Nordrhein-Westfalen (NRW) parts of the GlobalGeoTree analysis.
The rendered HTML hides the code (`echo: false`) so the report reads like a normal analysis report. The code in the QMD source is still commented and readable.

The dataset used here is always called `GlobalGeoTree.csv`.

## What I changed after checking the older QMDs and PDFs

| Old visualisation / section | Decision | Reason |
|---|---|---|
| Germany overview table | Keep | It gives the two most important checks before any maps: the number of German observations and the number of recorded species. |
| Germany distribution map | Improve | The old grid map was too pale and hard to interpret. The default map is now a smoother hotspot map, and two alternative map types are included in the setup. |
| Observations by year | Keep | It shows whether the German subset is dominated by a few recent years. This is important context for all later maps. |
| Observations by source: Germany map and bar chart | Keep | The map shows where the main sources are active. The bar chart shows which source groups dominate the dataset. |
| Observations mapped to Bundesländer, plus Bundesländer table, bar chart and map | Keep | The table checks the spatial join, the bar chart ranks states, and the map shows the regional pattern. |
| Berlin district boundary loading and Berlin observations table | Keep | These are essential checks before any Berlin district or density map. |
| Berlin observation/species bar chart | Remove | Observations and species are different measures on very different scales, so the table is clearer than a two-bar plot. |
| Berlin source table and source bar chart | Keep | Berlin is strongly shaped by a few sources, so source composition should be shown. |
| Spatial distribution of Berlin observations | Keep | This was the strongest Berlin visual because it clearly shows observation hotspots inside the city. |
| Berlin local-area table, bar chart, species chart and map | Keep, but rename carefully | The downloaded Berlin file has about 97 polygons, so these are better described as Berlin local areas / district polygons rather than the 12 official Bezirke. |
| Berlin source-by-district heatmap | Remove from the main report | With many local areas and long source names, it becomes visually crowded and hard to read. |
| Berlin top recorded species table and bar chart | Keep | It adds useful biological interpretation, as long as the caption says it is based on recorded observations, not true abundance. |
| NRW boundary check and summary table | Add | The NRW PDF already had a useful boundary workflow and summary, so it belongs in this unified report. |
| NRW source table and source bar chart | Add | NRW is strongly dominated by a few sources, so the source chart is needed. |
| NRW observation map | Add, but improve | A raw point map can overplot, so the default NRW map uses a density/hotspot layer with the NRW boundary. |
| NRW district table, district bar chart, species bar chart and district map | Add | These are useful for comparing NUTS3 districts / Kreise and kreisfreie Städte. |
| NRW source map | Add as a faceted source map | It helps check whether the largest sources have different spatial footprints. |
| NRW common-name / GBIF enrichment | Remove | It depends on internet access and extra packages. Scientific names are more reproducible. |
| NRW category composition | Remove | The user confirmed that category should be ignored because it is not available in the full data file. |
| Germany category and climate-zone sections | Remove | Category is not available in the full file, and the climate-zone section needs an extra polygon file. |
| Very wide first-10-row preview tables | Remove | They were not readable in the rendered report and did not directly answer the visualisation questions. |

## Density map options used in this report

The report now keeps several density-style maps so you can compare them directly in the rendered HTML.
This is better than forcing one single map style too early.

The main comparison styles are:

1. **Points + contours**: good when you want to see both the individual recorded points and the broader hotspot shapes.
2. **Smooth filled density**: good for a clean presentation view of hotspots.
3. **Gridded heatmap**: good when you want counts in visible cells rather than a smoothed surface.
4. **Sampled point map**: good as an honest reference map showing raw record locations.

My recommendation is to keep **points + contours** and **smooth filled density** in the final report, and use the **gridded heatmap** as a comparison/checking visual.

## Setup

```{r setup}
# I only load packages that are used later in this report.
# Keeping the package list short makes the script easier to run on another computer.
required_packages <- c(
  "data.table",  # fast reading of the large CSV file
  "dplyr",       # data cleaning and summaries
  "stringr",     # simple text cleaning
  "ggplot2",     # plots
  "scales",      # nicer number labels
  "sf",          # spatial data and spatial joins
  "knitr",       # tables
  "tibble"       # small readable data frames
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Please install these packages before rendering this report: ",
    paste(missing_packages, collapse = ", ")
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

# A seed makes all sampled point maps reproducible.
# The counts always use all rows; only some displayed point layers are sampled.
set.seed(42)

# Turning s2 off often makes point-in-polygon joins easier for these administrative layers.
sf::sf_use_s2(FALSE)

# The user confirmed the full dataset name.
observations_path <- "GlobalGeoTree.csv"

if (!file.exists(observations_path)) {
  stop("GlobalGeoTree.csv was not found. Put it in the same folder as this QMD before rendering.")
}

# Keep downloaded boundary files in one place.
dir.create("data/external", recursive = TRUE, showWarnings = FALSE)

bundeslaender_local_path <- "data/external/germany_nuts1_2021.geojson"
nuts3_local_path <- "data/external/germany_nuts3_2021.geojson"
berlin_local_path <- "data/external/berlin_districts.geojson"

gisco_nuts1_url <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_4326_LEVL_1.geojson"
gisco_nuts3_url <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_4326_LEVL_3.geojson"
berlin_district_url <- "https://raw.githubusercontent.com/codeforgermany/click_that_hood/main/public/data/berlin.geojson"

# Map limits used for Germany. These keep the view focused on Germany.
germany_xlim <- c(5, 16)
germany_ylim <- c(47, 56)

# Approximate fallback boxes, used only if polygon boundaries are unavailable.
berlin_bbox <- list(lon_min = 13.088, lon_max = 13.761, lat_min = 52.338, lat_max = 52.675)
nrw_bbox <- list(lon_min = 5.80, lon_max = 9.55, lat_min = 50.25, lat_max = 52.65)

# The report now shows several density-map styles for comparison.
# This variable picks which style should be treated as the main Germany map in the text.
germany_distribution_map_type <- "contours_points"

# Point limits keep the HTML file readable and fast to render.
max_points_per_source_group <- 15000
max_berlin_points <- 8000
max_nrw_points <- 12000
```

```{r helper-functions}
# Clean column names in the same way every time.
# Example: "Country Code" becomes "country_code".
clean_colnames <- function(x) {
  x <- tolower(gsub("[^A-Za-z0-9]+", "_", trimws(x)))
  x <- gsub("^_+|_+$", "", x)
  make.unique(x, sep = "_")
}

# Read only the columns needed for this analysis.
# This is much faster than loading every column from a large CSV file.
read_selected_columns <- function(path, columns_wanted) {
  raw_names <- names(data.table::fread(path, nrows = 0, showProgress = FALSE))
  clean_names <- clean_colnames(raw_names)

  keep <- clean_names %in% columns_wanted
  data <- data.table::fread(path, select = raw_names[keep], showProgress = FALSE)
  names(data) <- clean_names[keep]

  tibble::as_tibble(data)
}

nice_table <- function(data, caption = NULL) {
  knitr::kable(data, caption = caption, format.args = list(big.mark = ","))
}

format_number <- function(x) {
  scales::comma(x, accuracy = 1)
}

format_percent <- function(x) {
  scales::percent(x, accuracy = 0.1)
}

sample_up_to <- function(data, n_max) {
  if (nrow(data) <= n_max) {
    data
  } else {
    dplyr::slice_sample(data, n = n_max)
  }
}

empty_plot <- function(message) {
  ggplot() +
    annotate("text", x = 0, y = 0, label = message, size = 4) +
    theme_void()
}

theme_gtree <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(size = base_size, color = "gray35"),
      plot.caption = element_text(size = base_size - 2, color = "gray45"),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")
    )
}

theme_map_gtree <- function(base_size = 12) {
  theme_void(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(size = base_size, color = "gray35"),
      plot.caption = element_text(size = base_size - 2, color = "gray45"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

make_source_label <- function(source) {
  source <- stringr::str_squish(as.character(source))

  dplyr::case_when(
    stringr::str_detect(source, stringr::regex("Pl@ntNet automatically", ignore_case = TRUE)) ~ "Pl@ntNet auto ID",
    stringr::str_detect(source, stringr::regex("Pl@ntNet observations", ignore_case = TRUE)) ~ "Pl@ntNet observations",
    stringr::str_detect(source, stringr::regex("iNaturalist", ignore_case = TRUE)) ~ "iNaturalist research-grade",
    stringr::str_detect(source, stringr::regex("Observation\\.org", ignore_case = TRUE)) ~ "Observation.org",
    stringr::str_detect(source, stringr::regex("naturgucker", ignore_case = TRUE)) ~ "naturgucker",
    stringr::str_detect(source, stringr::regex("ArtenFinder", ignore_case = TRUE)) ~ "ArtenFinder",
    is.na(source) | source == "" ~ "Unknown source",
    TRUE ~ stringr::str_trunc(source, width = 35)
  )
}

read_geojson_local_or_url <- function(local_path, url, label) {
  source_note <- paste("No", label, "boundary file was loaded.")
  x <- NULL

  if (file.exists(local_path)) {
    x <- try(sf::st_read(local_path, quiet = TRUE), silent = TRUE)
    source_note <- paste("Loaded", label, "boundaries from", local_path)
  } else {
    download_result <- try(
      utils::download.file(url, local_path, quiet = TRUE, mode = "wb"),
      silent = TRUE
    )

    if (!inherits(download_result, "try-error") && file.exists(local_path)) {
      x <- try(sf::st_read(local_path, quiet = TRUE), silent = TRUE)
      source_note <- paste("Downloaded", label, "boundaries from public source")
    }
  }

  if (inherits(x, "try-error") || is.null(x)) {
    return(list(data = NULL, note = source_note))
  }

  x <- sf::st_make_valid(x) |>
    sf::st_transform(4326)

  names(x) <- clean_colnames(names(x))
  list(data = x, note = source_note)
}

# These helper functions build different density-map styles.
# I kept them separate so the code is easier to read and easier to edit later.
plot_points_contours_map <- function(points_data, boundary_data = NULL,
                                     title, subtitle, caption = NULL,
                                     xlim = NULL, ylim = NULL,
                                     point_alpha = 0.18, point_size = 0.45,
                                     bins = 9) {
  p <- ggplot()

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = "gray98", color = "gray75", linewidth = 0.25)
  }

  p <- p +
    geom_point(
      data = points_data,
      aes(x = longitude, y = latitude),
      color = "gray20",
      alpha = point_alpha,
      size = point_size
    ) +
    stat_density_2d(
      data = points_data,
      aes(x = longitude, y = latitude, color = after_stat(level)),
      bins = bins,
      linewidth = 0.8,
      alpha = 0.95
    )

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = NA, color = "gray35", linewidth = 0.35)
  }

  if (!is.null(xlim) && !is.null(ylim)) {
    p <- p + coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)
  } else {
    p <- p + coord_sf(expand = FALSE)
  }

  p +
    scale_color_viridis_c(option = "D", end = 0.95, name = "Contour
level") +
    labs(
      title = title,
      subtitle = subtitle,
      caption = caption
    ) +
    theme_map_gtree()
}

plot_filled_density_map <- function(points_data, boundary_data = NULL,
                                    title, subtitle, caption = NULL,
                                    xlim = NULL, ylim = NULL,
                                    bins = 9, alpha = 0.88) {
  p <- ggplot()

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = "gray98", color = "gray75", linewidth = 0.25)
  }

  p <- p +
    stat_density_2d_filled(
      data = points_data,
      aes(x = longitude, y = latitude, fill = after_stat(level)),
      contour_var = "density",
      bins = bins,
      alpha = alpha
    )

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = NA, color = "gray35", linewidth = 0.35)
  }

  if (!is.null(xlim) && !is.null(ylim)) {
    p <- p + coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)
  } else {
    p <- p + coord_sf(expand = FALSE)
  }

  p +
    scale_fill_viridis_d(option = "D", direction = -1, name = "Density") +
    labs(
      title = title,
      subtitle = subtitle,
      caption = caption
    ) +
    theme_map_gtree()
}

plot_grid_density_map <- function(points_data, boundary_data = NULL,
                                  title, subtitle, caption = NULL,
                                  xlim = NULL, ylim = NULL,
                                  bins = 65) {
  p <- ggplot()

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = "gray98", color = "gray75", linewidth = 0.25)
  }

  p <- p +
    stat_bin_2d(
      data = points_data,
      aes(x = longitude, y = latitude, fill = after_stat(count)),
      bins = bins,
      alpha = 0.92
    )

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = NA, color = "gray35", linewidth = 0.30)
  }

  if (!is.null(xlim) && !is.null(ylim)) {
    p <- p + coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)
  } else {
    p <- p + coord_sf(expand = FALSE)
  }

  p +
    scale_fill_viridis_c(option = "D", trans = "sqrt", labels = scales::comma_format(), name = "Observations") +
    labs(
      title = title,
      subtitle = subtitle,
      caption = caption
    ) +
    theme_map_gtree()
}

plot_points_only_map <- function(points_data, boundary_data = NULL,
                                 title, subtitle, caption = NULL,
                                 xlim = NULL, ylim = NULL,
                                 point_alpha = 0.12, point_size = 0.25) {
  p <- ggplot()

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = "gray98", color = "gray75", linewidth = 0.25)
  }

  p <- p +
    geom_point(
      data = points_data,
      aes(x = longitude, y = latitude),
      color = "gray15",
      alpha = point_alpha,
      size = point_size
    )

  if (!is.null(boundary_data)) {
    p <- p + geom_sf(data = boundary_data, fill = NA, color = "gray35", linewidth = 0.30)
  }

  if (!is.null(xlim) && !is.null(ylim)) {
    p <- p + coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)
  } else {
    p <- p + coord_sf(expand = FALSE)
  }

  p +
    labs(
      title = title,
      subtitle = subtitle,
      caption = caption
    ) +
    theme_map_gtree()
}
```

## Load and clean the observation data

The data file is large, so the script reads only the columns needed for this report.
This keeps the report faster and makes the workflow easier to understand.

```{r load-observations}
columns_wanted <- c(
  "sample_id", "country_code", "level0", "level1_family", "level2_genus",
  "level3_species", "location", "source", "species_key", "year",
  "longitude", "latitude"
)

obs <- read_selected_columns(observations_path, columns_wanted)

required_cols <- c("country_code", "longitude", "latitude", "level3_species")
missing_required <- setdiff(required_cols, names(obs))

if (length(missing_required) > 0) {
  stop("The data file is missing required columns: ", paste(missing_required, collapse = ", "))
}

# Add optional columns as NA if they are missing.
# This avoids many if-statements later in the script.
for (col in setdiff(columns_wanted, names(obs))) {
  obs[[col]] <- NA
}

obs <- obs |>
  mutate(
    country_code = stringr::str_to_upper(stringr::str_squish(as.character(country_code))),
    level3_species = stringr::str_squish(as.character(level3_species)),
    longitude = suppressWarnings(as.numeric(longitude)),
    latitude = suppressWarnings(as.numeric(latitude)),
    year = suppressWarnings(as.integer(year)),
    source = stringr::str_squish(as.character(source)),
    source = if_else(is.na(source) | source == "", "Unknown source", source),
    source_short = make_source_label(source)
  ) |>
  filter(
    !is.na(longitude), !is.na(latitude),
    dplyr::between(longitude, -180, 180),
    dplyr::between(latitude, -90, 90),
    !is.na(level3_species), level3_species != ""
  )

germany <- obs |>
  filter(country_code == "DE")

germany_points <- sf::st_as_sf(
  germany,
  coords = c("longitude", "latitude"),
  crs = 4326,
  remove = FALSE
)
```

```{r input-summary}
input_summary <- tibble::tibble(
  item = c(
    "Input file",
    "Rows after coordinate/species cleaning",
    "Countries after cleaning",
    "Germany rows used in this report"
  ),
  value = c(
    observations_path,
    format_number(nrow(obs)),
    format_number(dplyr::n_distinct(obs$country_code)),
    format_number(nrow(germany))
  )
)

nice_table(input_summary, "Input data check")
```

## Germany overview table

**Why this table is needed:** before drawing any maps, I want a quick check of how much German data is available.
This also keeps the requested metric table in the report.

```{r germany-overview-table}
germany_overview <- tibble::tibble(
  metric = c("Germany observations", "Germany recorded species"),
  value = c(
    format_number(nrow(germany)),
    format_number(dplyr::n_distinct(germany$level3_species))
  )
)

nice_table(germany_overview, "Germany overview")
```

**Interpretation.** The observations value counts records, while the recorded species value counts unique scientific species names. These two numbers answer different questions, so they should stay separate.

## Load Bundesländer boundaries

The Germany state maps need Bundesländer polygons. The code first looks for a local GeoJSON file and then tries the public GISCO NUTS1 file.

```{r load-bundeslaender}
bundeslaender_result <- read_geojson_local_or_url(
  bundeslaender_local_path,
  gisco_nuts1_url,
  "Bundesländer / NUTS1"
)

bundeslaender <- bundeslaender_result$data

if (!is.null(bundeslaender)) {
  if ("cntr_code" %in% names(bundeslaender)) {
    bundeslaender <- bundeslaender |> filter(as.character(cntr_code) == "DE")
  }

  if ("levl_code" %in% names(bundeslaender)) {
    bundeslaender <- bundeslaender |> filter(as.character(levl_code) == "1")
  }

  name_col <- intersect(c("name_latn", "nuts_name", "name", "nuts_id"), names(bundeslaender))[1]

  if (is.na(name_col)) {
    bundeslaender$bundesland <- paste("Bundesland", seq_len(nrow(bundeslaender)))
  } else {
    bundeslaender$bundesland <- as.character(bundeslaender[[name_col]])
  }

  bundeslaender <- bundeslaender |>
    select(bundesland, geometry) |>
    filter(!is.na(bundesland), bundesland != "")
}

has_bundeslaender <- !is.null(bundeslaender) && nrow(bundeslaender) > 0

bundeslaender_status <- tibble::tibble(
  item = c("Boundary status", "Number of Bundesländer polygons"),
  value = c(
    bundeslaender_result$note,
    ifelse(has_bundeslaender, format_number(nrow(bundeslaender)), "0")
  )
)

nice_table(bundeslaender_status, "Bundesländer boundary check")
```

## Distribution of observations in Germany

**Why this visual is needed:** the dataset contains coordinates, so a map is the fastest way to see where records are concentrated.
Because one map type does not tell the whole story, the report now keeps several Germany density visualisations for comparison.

**Suggested reading order:** start with the **points + contours** map, check the **smooth filled density** map for a cleaner hotspot view, and use the **grid map** as a non-smoothed count-based comparison.

```{r germany-distribution-main, fig.width=8.5, fig.height=7}
germany_points_for_maps <- sample_up_to(germany, 70000)
germany_boundary_for_maps <- if (has_bundeslaender) dplyr::bind_rows(bundeslaender) else NULL

if (nrow(germany_points_for_maps) < 3) {
  empty_plot("Not enough Germany observations were available for the distribution maps.")
} else if (germany_distribution_map_type == "contours_points") {
  plot_points_contours_map(
    points_data = germany_points_for_maps,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: points plus density contours",
    subtitle = "Contours show broader hotspots, while semi-transparent points keep the underlying records visible.",
    caption = paste("Displayed points:", format_number(nrow(germany_points_for_maps)), "of", format_number(nrow(germany)), "Germany observations."),
    xlim = germany_xlim,
    ylim = germany_ylim,
    point_alpha = 0.10,
    point_size = 0.22,
    bins = 10
  )
} else if (germany_distribution_map_type == "smooth_fill") {
  plot_filled_density_map(
    points_data = germany_points_for_maps,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: smooth filled density",
    subtitle = "A cleaner hotspot view that is easy to read in a report.",
    caption = "This shows recording density in the dataset, not true tree abundance.",
    xlim = germany_xlim,
    ylim = germany_ylim,
    bins = 10,
    alpha = 0.90
  )
} else if (germany_distribution_map_type == "grid") {
  plot_grid_density_map(
    points_data = germany,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: gridded heatmap",
    subtitle = "Each square counts records; square-root scaling keeps smaller hotspots visible.",
    caption = "This map is useful when you want a less smoothed, more count-based view.",
    xlim = germany_xlim,
    ylim = germany_ylim,
    bins = 65
  )
} else {
  plot_points_only_map(
    points_data = germany_points_for_maps,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: sampled points",
    subtitle = "A simple raw-point reference map.",
    caption = paste("Displayed points:", format_number(nrow(germany_points_for_maps)), "of", format_number(nrow(germany)), "Germany observations."),
    xlim = germany_xlim,
    ylim = germany_ylim,
    point_alpha = 0.10,
    point_size = 0.22
  )
}
```

**Interpretation.** Darker/higher-density areas show places with more recorded observations. This should be interpreted as observation intensity, because a place can have many records due to active platforms, active observers, or easier access.

### Germany density map comparison

These extra maps are kept on purpose so you can compare styles in the rendered HTML.

```{r germany-distribution-contours, fig.width=8.5, fig.height=7}
if (nrow(germany_points_for_maps) < 3) {
  empty_plot("Not enough Germany observations were available for the comparison maps.")
} else {
  plot_points_contours_map(
    points_data = germany_points_for_maps,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: points plus density contours",
    subtitle = "Best balanced option: visible points plus coloured hotspot contours.",
    caption = "Good compromise between raw records and broad hotspot structure.",
    xlim = germany_xlim,
    ylim = germany_ylim,
    point_alpha = 0.10,
    point_size = 0.22,
    bins = 10
  )
}
```

```{r germany-distribution-filled, fig.width=8.5, fig.height=7}
if (nrow(germany_points_for_maps) < 3) {
  empty_plot("Not enough Germany observations were available for the comparison maps.")
} else {
  plot_filled_density_map(
    points_data = germany_points_for_maps,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: smooth filled density",
    subtitle = "Clean presentation option, but smoothing can make sparse areas look more continuous.",
    caption = "Best for a clean visual summary of hotspots.",
    xlim = germany_xlim,
    ylim = germany_ylim,
    bins = 10,
    alpha = 0.90
  )
}
```

```{r germany-distribution-grid, fig.width=8.5, fig.height=7}
if (nrow(germany) < 1) {
  empty_plot("No Germany observations were available for the grid map.")
} else {
  plot_grid_density_map(
    points_data = germany,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: gridded heatmap",
    subtitle = "Each square counts observations directly, so this view is less smoothed than the density maps.",
    caption = "Useful when you want visible count cells instead of a smoothed surface.",
    xlim = germany_xlim,
    ylim = germany_ylim,
    bins = 65
  )
}
```

```{r germany-distribution-points, fig.width=8.5, fig.height=7}
if (nrow(germany_points_for_maps) < 1) {
  empty_plot("No Germany observations were available for the point map.")
} else {
  plot_points_only_map(
    points_data = germany_points_for_maps,
    boundary_data = germany_boundary_for_maps,
    title = "Germany observations: sampled points",
    subtitle = "Raw-point reference map with no smoothing.",
    caption = paste("Displayed points:", format_number(nrow(germany_points_for_maps)), "of", format_number(nrow(germany)), "Germany observations."),
    xlim = germany_xlim,
    ylim = germany_ylim,
    point_alpha = 0.10,
    point_size = 0.22
  )
}
```

**Quick recommendation.** If you want one final Germany map, I would keep **points + contours** as the main option and **smooth filled density** as the backup presentation option.

## Observations by year

**Why this visual is needed:** the year chart checks whether the dataset is balanced through time or dominated by a few recent years.
This matters because spatial maps can be influenced by when observations were collected.

```{r observations-by-year, fig.width=8, fig.height=4.8}
year_counts <- germany |>
  filter(!is.na(year), year >= 1900, year <= as.integer(format(Sys.Date(), "%Y"))) |>
  count(year, name = "observations") |>
  arrange(year)

if (nrow(year_counts) == 0) {
  empty_plot("No usable year values were found in the Germany rows.")
} else {
  ggplot(year_counts, aes(x = year, y = observations)) +
    geom_col(width = 0.75) +
    geom_line(linewidth = 0.6, color = "gray25") +
    geom_point(size = 1.8, color = "gray25") +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    scale_y_continuous(labels = scales::comma_format()) +
    labs(
      title = "Germany observations by year",
      subtitle = "Annual counts show when the Germany subset was most strongly represented.",
      x = "Year",
      y = "Observations"
    ) +
    theme_gtree()
}
```

**Interpretation.** Peaks in this chart should be read as peaks in recording activity. They do not automatically mean that more trees existed in those years.

## Observations by source on the country map

**Why this visual is needed:** different data sources can have different spatial footprints.
The map helps check whether one source dominates certain areas of Germany.

```{r source-groups}
source_counts_raw <- germany |>
  count(source, name = "observations", sort = TRUE) |>
  mutate(
    share = observations / sum(observations),
    source_short = make_source_label(source)
  )

top_sources <- source_counts_raw |>
  slice_head(n = 5) |>
  pull(source)

germany_sources <- germany |>
  mutate(
    source_group = if_else(source %in% top_sources, make_source_label(source), "Other sources"),
    source_group = stringr::str_wrap(source_group, width = 20)
  )

source_counts <- germany_sources |>
  count(source_group, name = "observations", sort = TRUE) |>
  mutate(
    share = observations / sum(observations),
    label = paste0(format_number(observations), " (", format_percent(share), ")"),
    source_group = factor(source_group, levels = rev(source_group))
  )

source_map_data <- germany_sources |>
  group_by(source_group) |>
  group_modify(~ sample_up_to(.x, max_points_per_source_group)) |>
  ungroup() |>
  mutate(source_group = factor(source_group, levels = levels(source_counts$source_group)))
```

```{r source-map, fig.width=10, fig.height=7}
p <- ggplot()

if (has_bundeslaender) {
  p <- p +
    geom_sf(data = bundeslaender, fill = "gray97", color = "gray70", linewidth = 0.2)
}

p +
  geom_point(
    data = source_map_data,
    aes(x = longitude, y = latitude),
    alpha = 0.18,
    size = 0.45
  ) +
  facet_wrap(~ source_group, ncol = 3) +
  coord_sf(xlim = germany_xlim, ylim = germany_ylim, expand = FALSE) +
  labs(
    title = "Spatial footprint of major observation sources in Germany",
    subtitle = "Each panel shows sampled observation points from one source group.",
    caption = paste("Maximum displayed points per source group:", format_number(max_points_per_source_group))
  ) +
  theme_map_gtree()
```

**Interpretation.** This map is useful for spotting source bias. If one source is dense in one region and another source is dense elsewhere, regional patterns may partly reflect data collection rather than biology.

## Observations by source as a bar chart

**Why this visual is needed:** the map shows geography, but the bar chart makes it easier to compare source sizes.
Both are needed because they answer different questions.

```{r source-bar-chart, fig.width=8.5, fig.height=5}
ggplot(source_counts, aes(x = source_group, y = observations)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.20))) +
  labs(
    title = "Germany observations by source group",
    subtitle = "The same grouping is used in the country map above.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree()
```

**Interpretation.** The largest bar shows which source contributes the most records. This is important because a dominant source can strongly shape the spatial pattern seen on the maps.

## Short note on the two Pl@ntNet source names

`Pl@ntNet observations` and `Pl@ntNet auto ID` are both Pl@ntNet-derived, but they do not mean exactly the same thing. In this report, I keep them separate because they use different validation routes: one is the regular shared-observation dataset, and the other is the automatically identified occurrences dataset.

## Observations mapped to Bundesländer

**Why this step is needed:** the original data has coordinates, not a ready-made Bundesland field.
To compare federal states, the points must first be spatially joined to the Bundesländer polygons.

```{r bundesland-spatial-join}
if (has_bundeslaender) {
  germany_bundeslaender <- sf::st_join(
    germany_points,
    bundeslaender,
    left = FALSE,
    join = sf::st_within
  ) |>
    sf::st_drop_geometry() |>
    tibble::as_tibble()

  bundesland_counts <- germany_bundeslaender |>
    group_by(bundesland) |>
    summarise(
      observations = n(),
      species = dplyr::n_distinct(level3_species),
      sources = dplyr::n_distinct(source),
      .groups = "drop"
    ) |>
    arrange(desc(observations))

  bundesland_map <- bundeslaender |>
    left_join(bundesland_counts, by = "bundesland") |>
    mutate(
      observations = dplyr::coalesce(observations, 0L),
      species = dplyr::coalesce(species, 0L),
      sources = dplyr::coalesce(sources, 0L)
    )
} else {
  germany_bundeslaender <- tibble::tibble()
  bundesland_counts <- tibble::tibble()
  bundesland_map <- NULL
}

bundesland_join_check <- tibble::tibble(
  metric = c(
    "Germany observations before Bundesland join",
    "Germany observations matched to a Bundesland",
    "Bundesländer with at least one observation"
  ),
  value = c(
    format_number(nrow(germany)),
    format_number(nrow(germany_bundeslaender)),
    format_number(nrow(bundesland_counts))
  )
)

nice_table(bundesland_join_check, "Bundesland spatial join check")
```

```{r bundesland-table}
if (nrow(bundesland_counts) == 0) {
  nice_table(tibble::tibble(note = "Bundesländer counts are unavailable because the boundary file was not loaded."))
} else {
  bundesland_counts |>
    mutate(
      observations = format_number(observations),
      species = format_number(species),
      sources = format_number(sources)
    ) |>
    nice_table("Observations mapped to Bundesländer")
}
```

**Interpretation.** This table is the cleanest check that the spatial join worked. If many Germany points were not matched, the boundary data or coordinate quality should be checked before trusting the maps.

## Bundesländer bar chart

**Why this visual is needed:** the bar chart is the best way to rank federal states because each state is compared on the same axis.

```{r bundesland-bar-chart, fig.width=8.5, fig.height=6}
if (nrow(bundesland_counts) == 0) {
  empty_plot("Bundesländer counts are unavailable because the boundary file was not loaded.")
} else {
  ggplot(bundesland_counts, aes(x = reorder(bundesland, observations), y = observations)) +
    geom_col() +
    geom_text(aes(label = format_number(observations)), hjust = -0.05, size = 3) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
    labs(
      title = "Germany observations by Bundesland",
      subtitle = "Observation counts are based on spatially joining coordinates to GISCO NUTS1 boundaries.",
      x = NULL,
      y = "Observations"
    ) +
    theme_gtree()
}
```

**Interpretation.** This chart shows which Bundesländer dominate the German subset. It should not be read as a direct comparison of tree abundance because the number of observations also depends on recording effort.

## Bundesländer map

**Why this visual is needed:** the map complements the bar chart by showing where high-record states are located geographically.

```{r bundesland-map, fig.width=8, fig.height=7}
if (is.null(bundesland_map)) {
  empty_plot("Bundesländer map is unavailable because the boundary file was not loaded.")
} else {
  ggplot(bundesland_map) +
    geom_sf(aes(fill = observations), color = "white", linewidth = 0.35) +
    scale_fill_viridis_c(option = "C", trans = "sqrt", labels = scales::comma_format()) +
    labs(
      title = "Observation frequency by Bundesland",
      subtitle = "Darker Bundesländer contain more Germany observations in the dataset.",
      fill = "Observations",
      caption = "Boundaries: GISCO NUTS1 regions. Counts: GlobalGeoTree Germany subset."
    ) +
    theme_map_gtree()
}
```

**Interpretation.** This map is useful for spatial context, but it is less precise for ranking than the bar chart. Larger states can also look visually more important because they occupy more map area.

# Berlin

## Load Berlin district boundaries

Berlin is handled separately because it needs a finer boundary layer than Bundesländer.
The public fallback file contains many small Berlin local-area polygons, so I describe the results as Berlin district/local-area results.

```{r load-berlin-boundaries}
berlin_result <- read_geojson_local_or_url(
  berlin_local_path,
  berlin_district_url,
  "Berlin district/local-area"
)

berlin_districts <- berlin_result$data

if (!is.null(berlin_districts)) {
  name_col <- intersect(c("name", "bezirk", "district", "gemeinde_name", "gen", "ortsteil"), names(berlin_districts))[1]

  if (is.na(name_col)) {
    berlin_districts$district <- paste("Area", seq_len(nrow(berlin_districts)))
  } else {
    berlin_districts$district <- as.character(berlin_districts[[name_col]])
  }

  berlin_districts <- berlin_districts |>
    select(district, geometry) |>
    filter(!is.na(district), district != "")
}

has_berlin_districts <- !is.null(berlin_districts) && nrow(berlin_districts) > 0

berlin_boundary_status <- tibble::tibble(
  item = c("Berlin district/local-area boundary status", "Number of Berlin polygons"),
  value = c(
    berlin_result$note,
    ifelse(has_berlin_districts, format_number(nrow(berlin_districts)), "0")
  )
)

nice_table(berlin_boundary_status, "Berlin boundary check")
```

## Berlin observations table

**Why this table is needed:** it shows how Berlin observations were selected and gives the main record counts before the Berlin maps.

```{r berlin-filter-and-summary}
if (has_berlin_districts) {
  district_bbox <- sf::st_bbox(berlin_districts)

  germany_points_near_berlin <- germany_points |>
    filter(
      longitude >= district_bbox[["xmin"]], longitude <= district_bbox[["xmax"]],
      latitude >= district_bbox[["ymin"]], latitude <= district_bbox[["ymax"]]
    )

  berlin_joined <- sf::st_join(
    germany_points_near_berlin,
    berlin_districts,
    left = FALSE,
    join = sf::st_within
  )

  berlin <- berlin_joined |>
    sf::st_drop_geometry() |>
    tibble::as_tibble()

  berlin_filter_method <- "Spatial join to Berlin district/local-area polygons"
} else {
  berlin <- germany |>
    filter(
      longitude >= berlin_bbox$lon_min,
      longitude <= berlin_bbox$lon_max,
      latitude >= berlin_bbox$lat_min,
      latitude <= berlin_bbox$lat_max
    ) |>
    mutate(district = NA_character_)

  berlin_filter_method <- "Approximate Berlin bounding box fallback"
}

berlin_year_text <- if (all(is.na(berlin$year))) {
  "No usable year values"
} else {
  paste(range(berlin$year, na.rm = TRUE), collapse = "-")
}

berlin_summary <- tibble::tibble(
  metric = c(
    "Berlin filtering method",
    "Berlin observations",
    "Berlin recorded species",
    "Available source values",
    "Available years"
  ),
  value = c(
    berlin_filter_method,
    format_number(nrow(berlin)),
    format_number(dplyr::n_distinct(berlin$level3_species)),
    format_number(dplyr::n_distinct(berlin$source)),
    berlin_year_text
  )
)

nice_table(berlin_summary, "Berlin observation summary")
```

**Interpretation.** I removed the old observation/species bar chart because it put two very different measures on one axis. This table is clearer.

## Berlin observation sources

**Why this visual is needed:** Berlin observations are not equally distributed across sources, so source composition is important context for every Berlin map.

```{r berlin-source-summary}
berlin_source_counts <- berlin |>
  mutate(source_short = make_source_label(source)) |>
  count(source_short, name = "observations", sort = TRUE) |>
  mutate(
    share = observations / sum(observations),
    label = paste0(format_number(observations), " (", format_percent(share), ")"),
    source_short = factor(source_short, levels = rev(source_short))
  )

berlin_source_counts |>
  mutate(
    observations = format_number(observations),
    share = format_percent(share)
  ) |>
  nice_table("Berlin observations by source")
```

```{r berlin-source-bar, fig.width=8.5, fig.height=5}
ggplot(berlin_source_counts, aes(x = source_short, y = observations)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Berlin observations by source",
    subtitle = "Short labels are used because the original source names are very long.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree()
```

**Interpretation.** This chart tells us which platforms shape the Berlin subset most. If one source dominates, the Berlin hotspots may partly show that source's recording behaviour.

## Spatial distribution of Berlin observations

**Why this visual is needed:** this is the clearest Berlin plot because it shows where observations cluster inside the city.
The density layer avoids overplotting, and the local-area borders give spatial context.

```{r berlin-spatial-distribution, fig.width=8, fig.height=7}
berlin_points_for_map <- sample_up_to(berlin, max_berlin_points)

if (nrow(berlin_points_for_map) < 3) {
  empty_plot("Not enough Berlin observations were available for a density map.")
} else if (has_berlin_districts) {
  ggplot() +
    stat_density_2d_filled(
      data = berlin_points_for_map,
      aes(x = longitude, y = latitude, fill = after_stat(level)),
      contour_var = "ndensity",
      bins = 8,
      alpha = 0.78
    ) +
    geom_sf(data = berlin_districts, fill = NA, color = "gray25", linewidth = 0.35) +
    geom_point(data = berlin_points_for_map, aes(x = longitude, y = latitude), alpha = 0.14, size = 0.35) +
    coord_sf() +
    scale_fill_viridis_d(option = "C", direction = -1, name = "Density") +
    labs(
      title = "Tree observation density in Berlin",
      subtitle = "Filled contours show observation density; local-area borders provide spatial context.",
      x = NULL,
      y = NULL
    ) +
    theme_map_gtree()
} else {
  ggplot(berlin_points_for_map, aes(x = longitude, y = latitude)) +
    stat_density_2d_filled(contour_var = "ndensity", bins = 8, alpha = 0.8) +
    geom_point(alpha = 0.12, size = 0.35) +
    coord_fixed() +
    scale_fill_viridis_d(option = "C", direction = -1, name = "Density") +
    labs(
      title = "Tree observation density in Berlin",
      subtitle = "District borders were unavailable, so this map uses the Berlin bounding-box fallback.",
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_gtree()
}
```

**Interpretation.** The density map highlights observation hotspots inside Berlin. Like the Germany maps, it is a map of recorded observations, not a direct map of all trees in the city.

## Berlin local areas: observations and species

**Why this section is needed:** the density map shows hotspots, but district/local-area summaries make the pattern easier to compare and check.

```{r berlin-district-summary}
if ("district" %in% names(berlin) && any(!is.na(berlin$district))) {
  berlin_district_counts <- berlin |>
    filter(!is.na(district), district != "") |>
    group_by(district) |>
    summarise(
      observations = n(),
      species = dplyr::n_distinct(level3_species),
      sources = dplyr::n_distinct(source),
      .groups = "drop"
    ) |>
    arrange(desc(observations))
} else {
  berlin_district_counts <- tibble::tibble()
}

if (nrow(berlin_district_counts) > 0) {
  berlin_district_counts |>
    slice_head(n = 20) |>
    mutate(
      observations = format_number(observations),
      species = format_number(species),
      sources = format_number(sources)
    ) |>
    nice_table("Top Berlin local areas by observation count")
} else {
  nice_table(tibble::tibble(note = "Berlin local-area counts are unavailable because polygons were not loaded."))
}
```

```{r berlin-district-bar, fig.width=8.5, fig.height=6}
if (nrow(berlin_district_counts) == 0) {
  empty_plot("Berlin local-area counts are unavailable because polygons were not loaded.")
} else {
  berlin_district_counts |>
    slice_head(n = 20) |>
    ggplot(aes(x = reorder(district, observations), y = observations)) +
    geom_col() +
    geom_text(aes(label = format_number(observations)), hjust = -0.05, size = 3) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
    labs(
      title = "Berlin observations by local area",
      subtitle = "Only the top 20 areas are shown to keep the chart readable.",
      x = NULL,
      y = "Observations"
    ) +
    theme_gtree()
}
```

```{r berlin-species-bar, fig.width=8.5, fig.height=6}
if (nrow(berlin_district_counts) == 0) {
  empty_plot("Berlin local-area species counts are unavailable because polygons were not loaded.")
} else {
  berlin_district_counts |>
    arrange(desc(species)) |>
    slice_head(n = 20) |>
    ggplot(aes(x = reorder(district, species), y = species)) +
    geom_col() +
    geom_text(aes(label = format_number(species)), hjust = -0.05, size = 3) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
    labs(
      title = "Recorded species by Berlin local area",
      subtitle = "This counts distinct scientific species names recorded in each area.",
      x = NULL,
      y = "Recorded species"
    ) +
    theme_gtree()
}
```

```{r berlin-district-map, fig.width=8, fig.height=7}
if (has_berlin_districts && nrow(berlin_district_counts) > 0) {
  berlin_district_map <- berlin_districts |>
    left_join(berlin_district_counts, by = "district") |>
    mutate(observations = dplyr::coalesce(observations, 0L))

  ggplot(berlin_district_map) +
    geom_sf(aes(fill = observations), color = "white", linewidth = 0.25) +
    scale_fill_viridis_c(option = "C", trans = "sqrt", labels = scales::comma_format()) +
    labs(
      title = "Observation frequency by Berlin local area",
      subtitle = "Square-root scaling keeps both high and medium-count areas visible.",
      fill = "Observations"
    ) +
    theme_map_gtree()
} else {
  empty_plot("Berlin local-area map is unavailable because polygons were not loaded.")
}
```

**Interpretation.** The bar charts are better for comparing exact ranks, while the map is better for seeing where the high-record areas are located.

## Top recorded species in Berlin

**Why this visual is needed:** this gives a simple biological summary after the spatial summaries. It should still be read as recorded observations, not a census of tree abundance.

```{r berlin-top-species}
berlin_top_species <- berlin |>
  count(level3_species, sort = TRUE, name = "observations") |>
  mutate(share = observations / sum(observations))

berlin_top_species |>
  slice_head(n = 15) |>
  mutate(
    observations = format_number(observations),
    share = format_percent(share)
  ) |>
  nice_table("Top recorded species in Berlin")
```

```{r berlin-top-species-bar, fig.width=8.5, fig.height=6}
berlin_top_species |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(level3_species, observations), y = observations)) +
  geom_col() +
  geom_text(aes(label = format_number(observations)), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Most recorded tree species in Berlin",
    subtitle = "Counts show observations in GlobalGeoTree, not true abundance in the city.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree()
```

# Nordrhein-Westfalen (NRW)

## Load NRW boundaries

**Why this step is needed:** NRW needs one boundary for filtering observations and NUTS3 boundaries for district-level comparisons.
NUTS3 regions correspond approximately to Kreise and kreisfreie Städte.

```{r load-nrw-boundaries}
nuts3_result <- read_geojson_local_or_url(
  nuts3_local_path,
  gisco_nuts3_url,
  "NUTS3"
)

nuts3 <- nuts3_result$data

nrw_boundary <- NULL
nrw_districts <- NULL

if (has_bundeslaender) {
  nrw_boundary <- bundeslaender |>
    filter(stringr::str_detect(stringr::str_to_lower(bundesland), "nordrhein|north rhine|westfalen|westphalia"))
}

if (!is.null(nuts3)) {
  if (!("nuts_id" %in% names(nuts3)) && "id" %in% names(nuts3)) {
    nuts3$nuts_id <- as.character(nuts3$id)
  }

  if (!("nuts_id" %in% names(nuts3))) {
    nuts3$nuts_id <- NA_character_
  }

  name_col <- intersect(c("name_latn", "nuts_name", "name", "nuts_id"), names(nuts3))[1]

  if (is.na(name_col)) {
    nuts3$district_name <- paste("NRW district", seq_len(nrow(nuts3)))
  } else {
    nuts3$district_name <- as.character(nuts3[[name_col]])
  }

  nrw_districts <- nuts3 |>
    filter(stringr::str_starts(as.character(nuts_id), "DEA"), as.character(nuts_id) != "DEA") |>
    mutate(district_id = as.character(nuts_id)) |>
    select(district_id, district_name, geometry)
}

has_nrw_boundary <- !is.null(nrw_boundary) && nrow(nrw_boundary) > 0
has_nrw_districts <- !is.null(nrw_districts) && nrow(nrw_districts) > 0

nrw_boundary_status <- tibble::tibble(
  item = c("NRW boundary available", "NUTS3 boundary status", "NRW district polygons"),
  value = c(
    ifelse(has_nrw_boundary, "Yes", "No"),
    nuts3_result$note,
    ifelse(has_nrw_districts, format_number(nrow(nrw_districts)), "0")
  )
)

nice_table(nrw_boundary_status, "NRW boundary check")
```

## Filter observations to NRW

**Why this table is needed:** before mapping NRW, we need to know how many observations and species were found and whether filtering used real boundaries or a fallback box.

```{r filter-nrw}
if (has_nrw_boundary) {
  nrw_joined <- sf::st_join(
    germany_points,
    nrw_boundary |> select(geometry),
    join = sf::st_within,
    left = FALSE
  )

  nrw_filter_note <- "Spatial join to official NUTS1 NRW boundary"
} else {
  nrw_joined <- germany_points |>
    filter(
      dplyr::between(longitude, nrw_bbox$lon_min, nrw_bbox$lon_max),
      dplyr::between(latitude, nrw_bbox$lat_min, nrw_bbox$lat_max)
    )

  nrw_filter_note <- "Approximate NRW bounding box fallback"
}

nrw <- nrw_joined |>
  sf::st_drop_geometry() |>
  tibble::as_tibble()

nrw_year_text <- if (all(is.na(nrw$year))) {
  "No usable year values"
} else {
  paste(range(nrw$year, na.rm = TRUE), collapse = "-")
}

nrw_summary <- tibble::tibble(
  metric = c(
    "NRW filtering method",
    "NRW observations",
    "NRW recorded species",
    "NRW recorded genera",
    "NRW recorded families",
    "Available source values",
    "Available years"
  ),
  value = c(
    nrw_filter_note,
    format_number(nrow(nrw)),
    format_number(dplyr::n_distinct(nrw$level3_species)),
    format_number(dplyr::n_distinct(nrw$level2_genus)),
    format_number(dplyr::n_distinct(nrw$level1_family)),
    format_number(dplyr::n_distinct(nrw$source)),
    nrw_year_text
  )
)

nice_table(nrw_summary, "NRW observation summary")
```

**Interpretation.** This table is a necessary quality check. If the boundary join is used, the NRW subset is more reliable than a simple bounding-box filter.

## NRW observation sources

**Why this visual is needed:** the NRW PDF showed that a few data sources contribute most records, so source composition should stay in the unified report.

```{r nrw-source-summary}
nrw_source_counts <- nrw |>
  mutate(source_short = make_source_label(source)) |>
  count(source_short, name = "observations", sort = TRUE) |>
  mutate(
    share = observations / sum(observations),
    label = paste0(format_number(observations), " (", format_percent(share), ")"),
    source_short = factor(source_short, levels = rev(source_short))
  )

nrw_source_counts |>
  mutate(
    observations = format_number(observations),
    share = format_percent(share)
  ) |>
  nice_table("NRW observations by source")
```

```{r nrw-source-bar, fig.width=8.5, fig.height=5}
ggplot(nrw_source_counts, aes(x = source_short, y = observations)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "NRW observations by source",
    subtitle = "Short labels are used because the original source names are very long.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree()
```

**Interpretation.** This chart explains source bias before the maps. A dominant source can make one region look more important simply because that platform has more records there.

## NRW observation density maps

**Why this visual is needed:** a map is needed to see the spatial distribution inside NRW, but a raw point map can be too crowded.
Because the best style is not obvious, this report keeps several NRW density visualisations for direct comparison.

**My recommendation:** the **points + contours** map is the best balanced option. It keeps the underlying point pattern visible, but it also shows smoother hotspot zones. The **smooth filled density** map is also good for presentation.

```{r nrw-density-main, fig.width=8, fig.height=7}
nrw_points_for_map <- sample_up_to(nrw, max_nrw_points)
nrw_boundary_for_map <- if (has_nrw_boundary) nrw_boundary else NULL

if (nrow(nrw_points_for_map) < 3) {
  empty_plot("Not enough NRW observations were available for the density maps.")
} else {
  plot_points_contours_map(
    points_data = nrw_points_for_map,
    boundary_data = nrw_boundary_for_map,
    title = "NRW observations: density contours plus points",
    subtitle = "Contours show broad concentration zones while sampled points remain visible.",
    caption = paste("Displayed points:", format_number(nrow(nrw_points_for_map)), "of", format_number(nrow(nrw)), "NRW observations."),
    point_alpha = 0.16,
    point_size = 0.55,
    bins = 9
  )
}
```

**Interpretation.** This map is good for seeing NRW hotspots, but it should not be interpreted as true tree abundance. It is a map of where records are concentrated.

### NRW density map comparison

```{r nrw-density-contours, fig.width=8, fig.height=7}
if (nrow(nrw_points_for_map) < 3) {
  empty_plot("Not enough NRW observations were available for the comparison maps.")
} else {
  plot_points_contours_map(
    points_data = nrw_points_for_map,
    boundary_data = nrw_boundary_for_map,
    title = "NRW observations: density contours plus points",
    subtitle = "Best balanced option. It shows both the hotspot shape and the underlying point cloud.",
    caption = "Good when you want more structure than a point map, but more honesty than a fully smoothed surface.",
    point_alpha = 0.16,
    point_size = 0.55,
    bins = 9
  )
}
```

```{r nrw-density-filled, fig.width=8, fig.height=7}
if (nrow(nrw_points_for_map) < 3) {
  empty_plot("Not enough NRW observations were available for the comparison maps.")
} else {
  plot_filled_density_map(
    points_data = nrw_points_for_map,
    boundary_data = nrw_boundary_for_map,
    title = "NRW observations: smooth filled density",
    subtitle = "A clear hotspot view, but smoothing can make sparse areas look more continuous than they are.",
    caption = "Best for a clean presentation-style hotspot map.",
    bins = 9,
    alpha = 0.90
  )
}
```

```{r nrw-density-grid, fig.width=8, fig.height=7}
if (nrow(nrw) < 1) {
  empty_plot("No NRW observations were available for the grid map.")
} else {
  plot_grid_density_map(
    points_data = nrw,
    boundary_data = nrw_boundary_for_map,
    title = "NRW observations: gridded heatmap",
    subtitle = "Each square counts records directly, so the result is less smoothed and easier to compare cell by cell.",
    caption = "Useful for a more count-based spatial summary.",
    bins = 45
  )
}
```

```{r nrw-density-points, fig.width=8, fig.height=7}
if (nrow(nrw_points_for_map) < 1) {
  empty_plot("No NRW observations were available for the point map.")
} else {
  plot_points_only_map(
    points_data = nrw_points_for_map,
    boundary_data = nrw_boundary_for_map,
    title = "NRW observations: sampled points",
    subtitle = "Raw-point reference map with no smoothing.",
    caption = paste("Displayed points:", format_number(nrow(nrw_points_for_map)), "of", format_number(nrow(nrw)), "NRW observations."),
    point_alpha = 0.16,
    point_size = 0.50
  )
}
```

**Quick recommendation.** If you want to keep two NRW maps, I would keep **density contours plus points** and **smooth filled density**. The contour version is more informative, while the filled version is cleaner.

## NRW districts: observations and species

**Why this step is needed:** NUTS3 districts provide a useful middle scale between individual points and the whole NRW region.

```{r nrw-district-join}
if (has_nrw_districts) {
  nrw_obs_sf <- sf::st_as_sf(
    nrw,
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )

  nrw_district_joined <- sf::st_join(
    nrw_obs_sf,
    nrw_districts |> select(district_id, district_name, geometry),
    join = sf::st_within,
    left = FALSE
  ) |>
    sf::st_drop_geometry() |>
    tibble::as_tibble()

  nrw_district_counts <- nrw_district_joined |>
    filter(!is.na(district_name), district_name != "") |>
    group_by(district_id, district_name) |>
    summarise(
      observations = n(),
      species = dplyr::n_distinct(level3_species),
      sources = dplyr::n_distinct(source),
      .groups = "drop"
    ) |>
    arrange(desc(observations))

  nrw_district_map <- nrw_districts |>
    left_join(nrw_district_counts, by = c("district_id", "district_name")) |>
    mutate(
      observations = dplyr::coalesce(observations, 0L),
      species = dplyr::coalesce(species, 0L),
      sources = dplyr::coalesce(sources, 0L)
    )
} else {
  nrw_district_counts <- tibble::tibble()
  nrw_district_map <- NULL
}

if (nrow(nrw_district_counts) > 0) {
  nrw_district_counts |>
    slice_head(n = 20) |>
    mutate(
      observations = format_number(observations),
      species = format_number(species),
      sources = format_number(sources)
    ) |>
    nice_table("Top NRW NUTS3 districts by observation count")
} else {
  nice_table(tibble::tibble(note = "NRW district counts are unavailable because NUTS3 polygons were not loaded."))
}
```

```{r nrw-district-bar, fig.width=8.5, fig.height=6}
if (nrow(nrw_district_counts) == 0) {
  empty_plot("NRW district counts are unavailable because NUTS3 polygons were not loaded.")
} else {
  nrw_district_counts |>
    slice_head(n = 20) |>
    ggplot(aes(x = reorder(district_name, observations), y = observations)) +
    geom_col() +
    geom_text(aes(label = format_number(observations)), hjust = -0.05, size = 3) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
    labs(
      title = "NRW observations by NUTS3 district",
      subtitle = "Only the top 20 districts are shown to keep the chart readable.",
      x = NULL,
      y = "Observations"
    ) +
    theme_gtree()
}
```

```{r nrw-district-species-bar, fig.width=8.5, fig.height=6}
if (nrow(nrw_district_counts) == 0) {
  empty_plot("NRW district species counts are unavailable because NUTS3 polygons were not loaded.")
} else {
  nrw_district_counts |>
    arrange(desc(species)) |>
    slice_head(n = 20) |>
    ggplot(aes(x = reorder(district_name, species), y = species)) +
    geom_col() +
    geom_text(aes(label = format_number(species)), hjust = -0.05, size = 3) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
    labs(
      title = "Recorded species by NRW NUTS3 district",
      subtitle = "This counts distinct scientific species names recorded in each district.",
      x = NULL,
      y = "Recorded species"
    ) +
    theme_gtree()
}
```

```{r nrw-district-map, fig.width=8, fig.height=7}
if (is.null(nrw_district_map)) {
  empty_plot("NRW district map is unavailable because NUTS3 polygons were not loaded.")
} else {
  ggplot(nrw_district_map) +
    geom_sf(aes(fill = observations), color = "white", linewidth = 0.15) +
    scale_fill_viridis_c(option = "C", trans = "sqrt", labels = scales::comma_format()) +
    labs(
      title = "Observation frequency across NRW NUTS3 districts",
      subtitle = "Square-root scaling improves readability when a few districts have very high counts.",
      fill = "Observations"
    ) +
    theme_map_gtree()
}
```

**Interpretation.** The district bar charts are best for ranking. The district map is best for seeing whether high-record districts are clustered geographically.

## Sources on the NRW map

**Why this visual is needed:** the source bar chart shows size, but this faceted map checks whether the largest sources cover the same parts of NRW.

```{r nrw-source-map, fig.width=9, fig.height=7}
nrw_top_source_names <- nrw |>
  count(source, sort = TRUE) |>
  slice_head(n = 4) |>
  pull(source)

nrw_source_points <- nrw |>
  filter(source %in% nrw_top_source_names) |>
  mutate(source_group = make_source_label(source)) |>
  group_by(source_group) |>
  group_modify(~ sample_up_to(.x, as.integer(max_nrw_points / 4))) |>
  ungroup()

if (nrow(nrw_source_points) == 0) {
  empty_plot("No source-specific NRW points were available for mapping.")
} else {
  p <- ggplot()

  if (has_nrw_boundary) {
    p <- p + geom_sf(data = nrw_boundary, fill = "gray97", color = "gray60", linewidth = 0.35)
  }

  p +
    geom_point(data = nrw_source_points, aes(x = longitude, y = latitude), alpha = 0.22, size = 0.45) +
    facet_wrap(~ source_group, ncol = 2) +
    coord_sf() +
    labs(
      title = "Spatial patterns of the largest NRW observation sources",
      subtitle = "Facets show the four largest source groups. Points are sampled for readability.",
      x = NULL,
      y = NULL
    ) +
    theme_map_gtree(base_size = 11)
}
```

**Interpretation.** If source panels look different, NRW spatial patterns may partly reflect source coverage. This makes the source map useful next to the overall density map.

## Top recorded species in NRW

**Why this visual is needed:** this gives a simple taxonomic summary for NRW without using optional GBIF common-name lookups.
Scientific names are kept because they are more reproducible.

```{r nrw-top-species}
nrw_top_species <- nrw |>
  count(level3_species, sort = TRUE, name = "observations") |>
  mutate(share = observations / sum(observations))

nrw_top_species |>
  slice_head(n = 20) |>
  mutate(
    observations = format_number(observations),
    share = format_percent(share)
  ) |>
  nice_table("Top recorded species in NRW")
```

```{r nrw-top-species-bar, fig.width=8.5, fig.height=6}
nrw_top_species |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(level3_species, observations), y = observations)) +
  geom_col() +
  geom_text(aes(label = format_number(observations)), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma_format(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Most recorded tree species in NRW",
    subtitle = "Counts show observations in GlobalGeoTree, not true abundance in NRW.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree()
```

## Final interpretation notes

The most useful visualisations are the ones that compare the same records in different ways: Germany maps for national spatial patterns, year counts for time, source charts for data bias, Bundesländer and NRW district charts for regional comparison, and Berlin density/local-area plots for city-level clustering.

The map and bar-chart pairs are intentionally kept together. Maps are good for spatial patterns, but bar charts are better for exact comparisons.

All maps should be interpreted as maps of **recording intensity** in `GlobalGeoTree.csv`. High values can reflect true tree presence, but they can also reflect where observers and data platforms are more active.
