---
title: "GlobalGeoTree Bayern: Observations, Sources, Districts, and Common Names"
author: "GlobalGeoTree team"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

## Purpose

This report focuses on **Bayern (Bavaria)**. It answers four practical inspection questions:

1. How many GlobalGeoTree observations are located in Bayern?
2. How many recorded species are located in Bayern?
3. Which data sources contribute Bayern observations?
4. How are tree observations distributed across Bayern districts?

The report also includes an **optional common-name enrichment step**. Scientific names remain the primary identifier, but the script can try to add English and German common names using GBIF through the optional `rgbif` package. Results are cached in a local CSV file so the report does not need to query GBIF every time.

> Important note: common names are not globally standardized. The report treats German and English names as *GBIF vernacular names where available*, not as guaranteed official names.

## Setup

```{r setup}
# Required packages only
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(sf)

set.seed(42)

# -------------------------------------------------------------------------
# User configuration
# -------------------------------------------------------------------------

# Main observations file. Put the full 6M file in the same folder if possible.
if (file.exists("GlobalGeoTree.csv")) {
  observations_path <- "GlobalGeoTree.csv"
} else if (file.exists("GlobalGeoTree-10kEval-900.csv")) {
  observations_path <- "GlobalGeoTree-10kEval-900.csv"
} else if (file.exists("GlobalGeoTree-10kEval-90.csv")) {
  observations_path <- "GlobalGeoTree-10kEval-90.csv"
} else {
  stop("No observations file found. Expected GlobalGeoTree_full.csv, GlobalGeoTree-10kEval-900.csv, or GlobalGeoTree-10kEval-90.csv in the same folder as this QMD.")
}

# Optional category lookup. This is not required for the Bayern inspection,
# but it is useful if you also want Rare/Common/Frequent labels where available.
category_lookup_path <- if (file.exists("GlobalGeoTree-10kEval-900.csv")) {
  "GlobalGeoTree-10kEval-900.csv"
} else if (file.exists("GlobalGeoTree-10kEval-90.csv")) {
  "GlobalGeoTree-10kEval-90.csv"
} else {
  NA_character_
}

# Optional GBIF common-name lookup.
# Set to FALSE if you want fully offline rendering.
enable_gbif_lookup <- TRUE
common_names_cache_path <- "species_common_names_gbif.csv"

# Number of points to plot on maps. Keeps HTML rendering readable.
max_points_on_map <- 8000

# If internet is unavailable, place local GISCO boundary files in either the root
# folder or data/external/ with these names.
local_nuts1_paths <- c(
  "germany_nuts1_2021.geojson",
  "data/external/germany_nuts1_2021.geojson"
)

local_nuts3_paths <- c(
  "germany_nuts3_2021.geojson",
  "data/external/germany_nuts3_2021.geojson"
)

# GISCO NUTS boundary files. These are used only if local files are not found.
gisco_nuts1_url <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_4326_LEVL_1.geojson"
gisco_nuts3_url <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_4326_LEVL_3.geojson"

# Rough fallback bounding box for Bayern if boundary files cannot be loaded.
state_bbox <- list(
  lon_min = 8.80,
  lon_max = 13.90,
  lat_min = 47.20,
  lat_max = 50.60
)
```

```{r helper-functions}
theme_gtree <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(size = base_size, color = "gray35"),
      plot.caption = element_text(size = base_size - 2, color = "gray45"),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")
    )
}

nice_table <- function(data, caption = NULL) {
  knitr::kable(data, caption = caption, format.args = list(big.mark = ","))
}

sample_points_for_plot <- function(data, max_n = 8000) {
  if (nrow(data) <= max_n) {
    data
  } else {
    dplyr::slice_sample(data, n = max_n)
  }
}

standardize_names <- function(data) {
  names(data) <- names(data) |>
    stringr::str_trim() |>
    stringr::str_replace_all("[^A-Za-z0-9]+", "_") |>
    stringr::str_replace_all("_+", "_") |>
    stringr::str_replace_all("_$", "") |>
    stringr::str_to_lower()
  data
}

read_geojson_from_local_or_url <- function(local_paths, url, label) {
  existing_path <- local_paths[file.exists(local_paths)][1]

  if (!is.na(existing_path)) {
    boundary <- sf::st_read(existing_path, quiet = TRUE)
    source_note <- paste("Loaded", label, "from", existing_path)
    return(list(data = sf::st_transform(sf::st_make_valid(boundary), 4326), note = source_note))
  }

  tmp_geojson <- tempfile(fileext = ".geojson")
  try_download <- try(
    utils::download.file(url, tmp_geojson, quiet = TRUE, mode = "wb"),
    silent = TRUE
  )

  if (!inherits(try_download, "try-error") && file.exists(tmp_geojson)) {
    boundary <- try(sf::st_read(tmp_geojson, quiet = TRUE), silent = TRUE)
    if (!inherits(boundary, "try-error")) {
      source_note <- paste("Downloaded", label, "from GISCO.")
      return(list(data = sf::st_transform(sf::st_make_valid(boundary), 4326), note = source_note))
    }
  }

  list(data = NULL, note = paste("Could not load", label, "boundaries. Falling back where possible."))
}

first_non_missing <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) NA_character_ else x[[1]]
}
```

## Load observations

```{r load-observations}
obs_raw <- data.table::fread(observations_path, showProgress = FALSE) |>
  standardize_names()

required_cols <- c("country_code", "level3_species", "longitude", "latitude")
missing_cols <- setdiff(required_cols, names(obs_raw))
if (length(missing_cols) > 0) {
  stop("The observations file is missing required columns: ", paste(missing_cols, collapse = ", "))
}

obs <- obs_raw |>
  mutate(
    across(where(is.character), str_squish),
    country_code = str_to_upper(country_code),
    level3_species = str_squish(level3_species),
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude),
    year = if ("year" %in% names(obs_raw)) as.integer(year) else NA_integer_,
    source = if ("source" %in% names(obs_raw)) str_squish(as.character(source)) else "Unknown",
    species_key_lookup = if ("species_key" %in% names(obs_raw)) as.character(species_key) else NA_character_
  ) |>
  filter(
    !is.na(longitude), !is.na(latitude),
    between(longitude, -180, 180),
    between(latitude, -90, 90),
    !is.na(level3_species), level3_species != "",
    !is.na(country_code), country_code != ""
  )

file_summary <- tibble::tibble(
  item = c("Input file", "Rows in input file", "Rows after coordinate/species cleaning", "Countries after cleaning"),
  value = c(
    observations_path,
    format(nrow(obs_raw), big.mark = ","),
    format(nrow(obs), big.mark = ","),
    format(n_distinct(obs$country_code), big.mark = ",")
  )
)

nice_table(file_summary, "Input data summary")
```

## Preview of the dataset

```{r head-table}
obs |>
  select(any_of(c(
    "sample_id", "country_code", "level0", "level1_family", "level2_genus",
    "level3_species", "location", "source", "species_key", "year", "longitude", "latitude", "category"
  ))) |>
  head(10) |>
  nice_table("First 10 cleaned rows")
```

## Load Bayern boundaries

```{r load-boundaries}
nuts1_result <- read_geojson_from_local_or_url(local_nuts1_paths, gisco_nuts1_url, "NUTS1")
nuts3_result <- read_geojson_from_local_or_url(local_nuts3_paths, gisco_nuts3_url, "NUTS3")

nuts1 <- nuts1_result$data
nuts3 <- nuts3_result$data

state_boundary <- NULL
state_districts <- NULL

if (!is.null(nuts1)) {
  nuts1 <- nuts1 |>
    mutate(
      nuts_id = if ("NUTS_ID" %in% names(nuts1)) as.character(NUTS_ID) else if ("nuts_id" %in% names(nuts1)) as.character(nuts_id) else NA_character_,
      name_latn = if ("NAME_LATN" %in% names(nuts1)) as.character(NAME_LATN) else if ("name_latn" %in% names(nuts1)) as.character(name_latn) else NA_character_,
      cntr_code = if ("CNTR_CODE" %in% names(nuts1)) as.character(CNTR_CODE) else if ("cntr_code" %in% names(nuts1)) as.character(cntr_code) else NA_character_
    )

  state_boundary <- nuts1 |>
    filter(
      nuts_id == "DE2" |
        str_detect(str_to_lower(name_latn), "bayern|bavaria")
    ) |>
    select(nuts_id, name_latn, geometry)
}

if (!is.null(nuts3)) {
  nuts3 <- nuts3 |>
    mutate(
      nuts_id = if ("NUTS_ID" %in% names(nuts3)) as.character(NUTS_ID) else if ("nuts_id" %in% names(nuts3)) as.character(nuts_id) else NA_character_,
      district_name = if ("NAME_LATN" %in% names(nuts3)) as.character(NAME_LATN) else if ("name_latn" %in% names(nuts3)) as.character(name_latn) else NA_character_,
      cntr_code = if ("CNTR_CODE" %in% names(nuts3)) as.character(CNTR_CODE) else if ("cntr_code" %in% names(nuts3)) as.character(cntr_code) else NA_character_
    )

  state_districts <- nuts3 |>
    filter(str_starts(nuts_id, "DE2"), nuts_id != "DE2") |>
    select(district_id = nuts_id, district_name, geometry)
}

boundary_summary <- tibble::tibble(
  item = c("NUTS1 boundary status", "NUTS3 boundary status", "Bayern boundary available", "Bayern district polygons"),
  value = c(
    nuts1_result$note,
    nuts3_result$note,
    ifelse(!is.null(state_boundary) && nrow(state_boundary) > 0, "Yes", "No"),
    ifelse(!is.null(state_districts), format(nrow(state_districts), big.mark = ","), "0")
  )
)

nice_table(boundary_summary, "Boundary data status")
```

## Filter observations to Bayern

If NUTS boundaries are available, the report identifies Bayern observations by spatially joining German observations to the Bayern polygon. If boundaries are not available, it falls back to an approximate Bayern bounding box.

```{r filter-nrw}
germany_obs <- obs |>
  filter(country_code %in% c("DE", "DEU", "GER", "GERMANY"))

if (!is.null(state_boundary) && nrow(state_boundary) > 0) {
  germany_sf <- sf::st_as_sf(germany_obs, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
  state_obs_sf <- sf::st_join(
    germany_sf,
    state_boundary |> select(state_name = name_latn, geometry),
    join = sf::st_within,
    left = FALSE
  )
  state_filter_note <- "Bayern observations were identified using the official NUTS1 Bayern boundary."
} else {
  state_obs <- germany_obs |>
    filter(
      between(longitude, state_bbox$lon_min, state_bbox$lon_max),
      between(latitude, state_bbox$lat_min, state_bbox$lat_max)
    )
  state_obs_sf <- sf::st_as_sf(state_obs, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
  state_filter_note <- "Bayern observations were approximated using a bounding box because no Bayern boundary was available."
}

state_obs <- sf::st_drop_geometry(state_obs_sf)

state_summary <- tibble::tibble(
  item = c(
    "Filter method",
    "Bayern observations",
    "Bayern recorded species",
    "Bayern recorded genera",
    "Bayern recorded families",
    "Years represented"
  ),
  value = c(
    state_filter_note,
    format(nrow(state_obs), big.mark = ","),
    format(n_distinct(state_obs$level3_species), big.mark = ","),
    if ("level2_genus" %in% names(state_obs)) format(n_distinct(state_obs$level2_genus), big.mark = ",") else "Not available",
    if ("level1_family" %in% names(state_obs)) format(n_distinct(state_obs$level1_family), big.mark = ",") else "Not available",
    if (all(is.na(state_obs$year))) "Not available" else paste(range(state_obs$year, na.rm = TRUE), collapse = "–")
  )
)

nice_table(state_summary, "Bayern summary")
```

## Optional: add Rare/Common/Frequent categories

The full GlobalGeoTree file may not contain `category`. If not, this report tries to add category labels from the larger evaluation file. This only labels species present in the evaluation lookup, so category coverage may be incomplete.

```{r optional-category-lookup}
category_lookup <- NULL
category_note <- "No category lookup was applied."

if ("category" %in% names(state_obs)) {
  state_obs_with_category <- state_obs |>
    mutate(category = str_squish(as.character(category)))
  category_note <- "Category was already present in the main observations file."
} else if (!is.na(category_lookup_path) && file.exists(category_lookup_path)) {
  eval_raw <- data.table::fread(category_lookup_path, showProgress = FALSE) |>
    standardize_names()

  if (all(c("level3_species", "category") %in% names(eval_raw))) {
    eval_clean <- eval_raw |>
      mutate(
        level3_species = str_squish(as.character(level3_species)),
        species_key_lookup = if ("species_key" %in% names(eval_raw)) as.character(species_key) else NA_character_,
        category = str_squish(as.character(category))
      ) |>
      filter(!is.na(level3_species), level3_species != "", !is.na(category), category != "")

    category_lookup <- eval_clean |>
      count(species_key_lookup, level3_species, category, name = "lookup_records") |>
      group_by(species_key_lookup, level3_species) |>
      slice_max(lookup_records, n = 1, with_ties = FALSE) |>
      ungroup() |>
      select(species_key_lookup, level3_species, category)

    # First join by species_key where available.
    state_obs_with_category <- state_obs |>
      left_join(
        category_lookup |>
          filter(!is.na(species_key_lookup), species_key_lookup != "") |>
          select(species_key_lookup, category_by_key = category),
        by = "species_key_lookup"
      ) |>
      left_join(
        category_lookup |>
          group_by(level3_species) |>
          summarise(category_by_name = first_non_missing(category), .groups = "drop"),
        by = "level3_species"
      ) |>
      mutate(category = coalesce(category_by_key, category_by_name)) |>
      select(-category_by_key, -category_by_name)

    category_note <- paste("Category was joined from", category_lookup_path, "by species_key where possible, then by scientific species name.")
  } else {
    state_obs_with_category <- state_obs
    category_note <- paste("Category lookup file was found, but it did not contain both level3_species and category:", category_lookup_path)
  }
} else {
  state_obs_with_category <- state_obs
}

category_coverage <- tibble::tibble(
  item = c(
    "Category method",
    "Bayern observations with known category",
    "Bayern species with known category"
  ),
  value = c(
    category_note,
    if ("category" %in% names(state_obs_with_category)) format(sum(!is.na(state_obs_with_category$category)), big.mark = ",") else "0",
    if ("category" %in% names(state_obs_with_category)) format(n_distinct(state_obs_with_category$level3_species[!is.na(state_obs_with_category$category)]), big.mark = ",") else "0"
  )
)

nice_table(category_coverage, "Category coverage in Bayern")
```

## Optional: add English and German common names

This section first checks for a local cache file `species_common_names_gbif.csv`. If common names for some Bayern species are missing and `rgbif` is installed, the report tries to retrieve English and German vernacular names from GBIF and updates the cache.

If this section is too slow, set `enable_gbif_lookup <- FALSE` in the setup chunk. The report will still render using scientific names.

```{r common-name-lookup}
state_species <- state_obs |>
  distinct(species_key_lookup, level3_species) |>
  mutate(
    species_key_lookup = as.character(species_key_lookup),
    scientific_name = level3_species
  ) |>
  select(species_key_lookup, scientific_name)

empty_common_names <- tibble::tibble(
  species_key_lookup = character(),
  scientific_name = character(),
  common_name_en = character(),
  common_name_de = character()
)

cached_common_names <- if (file.exists(common_names_cache_path)) {
  data.table::fread(common_names_cache_path, showProgress = FALSE) |>
    as_tibble() |>
    mutate(
      species_key_lookup = as.character(species_key_lookup),
      scientific_name = as.character(scientific_name),
      common_name_en = as.character(common_name_en),
      common_name_de = as.character(common_name_de)
    )
} else {
  empty_common_names
}

missing_for_lookup <- state_species |>
  anti_join(cached_common_names, by = c("species_key_lookup", "scientific_name"))

get_gbif_common_names_one <- function(species_key_lookup, scientific_name) {
  if (!requireNamespace("rgbif", quietly = TRUE)) {
    return(tibble::tibble(
      species_key_lookup = species_key_lookup,
      scientific_name = scientific_name,
      common_name_en = NA_character_,
      common_name_de = NA_character_
    ))
  }

  result <- tryCatch({
    gbif_key <- suppressWarnings(as.integer(species_key_lookup))

    if (is.na(gbif_key)) {
      backbone <- try(rgbif::name_backbone(name = scientific_name), silent = TRUE)
      if (!inherits(backbone, "try-error") && !is.null(backbone$usageKey)) {
        gbif_key <- backbone$usageKey
      }
    }

    if (is.na(gbif_key)) {
      return(tibble::tibble(
        species_key_lookup = species_key_lookup,
        scientific_name = scientific_name,
        common_name_en = NA_character_,
        common_name_de = NA_character_
      ))
    }

    vern <- rgbif::name_usage(key = gbif_key, data = "vernacularNames", limit = 200)

    if (is.list(vern) && "data" %in% names(vern)) {
      vern <- vern$data
    }

    if (is.null(vern) || nrow(vern) == 0) {
      return(tibble::tibble(
        species_key_lookup = species_key_lookup,
        scientific_name = scientific_name,
        common_name_en = NA_character_,
        common_name_de = NA_character_
      ))
    }

    vern <- as_tibble(vern)
    names(vern) <- names(vern) |>
      stringr::str_replace_all("([a-z0-9])([A-Z])", "\\1_\\2") |>
      stringr::str_to_lower()

    if (!all(c("vernacular_name", "language") %in% names(vern))) {
      return(tibble::tibble(
        species_key_lookup = species_key_lookup,
        scientific_name = scientific_name,
        common_name_en = NA_character_,
        common_name_de = NA_character_
      ))
    }

    en_name <- vern |>
      filter(language %in% c("eng", "en")) |>
      pull(vernacular_name) |>
      unique() |>
      first_non_missing()

    de_name <- vern |>
      filter(language %in% c("deu", "de", "ger")) |>
      pull(vernacular_name) |>
      unique() |>
      first_non_missing()

    tibble::tibble(
      species_key_lookup = species_key_lookup,
      scientific_name = scientific_name,
      common_name_en = en_name,
      common_name_de = de_name
    )
  }, error = function(e) {
    tibble::tibble(
      species_key_lookup = species_key_lookup,
      scientific_name = scientific_name,
      common_name_en = NA_character_,
      common_name_de = NA_character_
    )
  })

  result
}

if (enable_gbif_lookup && nrow(missing_for_lookup) > 0 && requireNamespace("rgbif", quietly = TRUE)) {
  newly_found_common_names <- bind_rows(Map(
    get_gbif_common_names_one,
    missing_for_lookup$species_key_lookup,
    missing_for_lookup$scientific_name
  ))

  common_names <- bind_rows(cached_common_names, newly_found_common_names) |>
    distinct(species_key_lookup, scientific_name, .keep_all = TRUE)

  data.table::fwrite(common_names, common_names_cache_path)
  common_name_note <- paste("Common names were updated using GBIF and cached in", common_names_cache_path)
} else {
  common_names <- cached_common_names
  common_name_note <- if (!enable_gbif_lookup) {
    "GBIF lookup disabled. Scientific names are used unless a local cache already exists."
  } else if (!requireNamespace("rgbif", quietly = TRUE)) {
    "Package 'rgbif' is not installed. Scientific names are used unless a local cache already exists. Install rgbif to enable GBIF lookup."
  } else {
    "All Bayern species were already covered by the local common-name cache."
  }
}

state_obs_named <- state_obs_with_category |>
  left_join(common_names, by = c("species_key_lookup", "level3_species" = "scientific_name")) |>
  mutate(
    display_name = case_when(
      !is.na(common_name_de) & common_name_de != "" ~ paste0(common_name_de, " (", level3_species, ")"),
      !is.na(common_name_en) & common_name_en != "" ~ paste0(common_name_en, " (", level3_species, ")"),
      TRUE ~ level3_species
    )
  )

common_name_summary <- tibble::tibble(
  item = c(
    "Common-name method",
    "Bayern species total",
    "Bayern species with English common name",
    "Bayern species with German common name"
  ),
  value = c(
    common_name_note,
    format(n_distinct(state_obs_named$level3_species), big.mark = ","),
    format(n_distinct(state_obs_named$level3_species[!is.na(state_obs_named$common_name_en) & state_obs_named$common_name_en != ""]), big.mark = ","),
    format(n_distinct(state_obs_named$level3_species[!is.na(state_obs_named$common_name_de) & state_obs_named$common_name_de != ""]), big.mark = ",")
  )
)

nice_table(common_name_summary, "Common-name coverage")
```

## Bayern observations and species

```{r observations-species-summary}
summary_metrics <- tibble::tibble(
  metric = c("Observations", "Recorded species", "Recorded genera", "Recorded families", "Data sources"),
  value = c(
    nrow(state_obs_named),
    n_distinct(state_obs_named$level3_species),
    if ("level2_genus" %in% names(state_obs_named)) n_distinct(state_obs_named$level2_genus) else NA_integer_,
    if ("level1_family" %in% names(state_obs_named)) n_distinct(state_obs_named$level1_family) else NA_integer_,
    n_distinct(state_obs_named$source)
  )
)

nice_table(summary_metrics, "Bayern observations and taxonomic coverage")
```

## Sources of Bayern observations

```{r source-summary}
source_summary <- state_obs_named |>
  count(source, sort = TRUE, name = "observations") |>
  mutate(share = observations / sum(observations))

source_summary |>
  head(15) |>
  mutate(
    observations = format(observations, big.mark = ","),
    share = percent(share, accuracy = 0.1)
  ) |>
  nice_table("Top 15 data sources in Bayern")
```

```{r source-plot, fig.width=8, fig.height=5}
source_summary |>
  slice_head(n = 10) |>
  mutate(source = reorder(source, observations)) |>
  ggplot(aes(x = observations, y = source)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Main sources of tree observations in Bayern",
    subtitle = "The plot shows the ten largest sources by observation count.",
    x = "Number of observations",
    y = NULL
  ) +
  theme_gtree()
```

## Bayern observation map

```{r nrw-map, fig.width=8, fig.height=7}
plot_points <- state_obs_sf |>
  sample_points_for_plot(max_points_on_map)

if (!is.null(state_boundary) && nrow(state_boundary) > 0) {
  ggplot() +
    geom_sf(data = state_boundary, fill = "gray96", color = "gray45", linewidth = 0.4) +
    geom_point(
      data = plot_points,
      aes(x = longitude, y = latitude),
      alpha = 0.20,
      size = 0.45
    ) +
    coord_sf() +
    labs(
      title = "Spatial distribution of GlobalGeoTree observations in Bayern",
      subtitle = "Each point is one observation. Points are sampled for readability if the dataset is large.",
      x = NULL,
      y = NULL,
      caption = paste("Displayed points:", format(nrow(plot_points), big.mark = ","), "of", format(nrow(state_obs_sf), big.mark = ","), "Bayern observations")
    ) +
    theme_gtree()
} else {
  ggplot(plot_points, aes(x = longitude, y = latitude)) +
    geom_point(alpha = 0.20, size = 0.45) +
    coord_fixed(
      xlim = c(state_bbox$lon_min, state_bbox$lon_max),
      ylim = c(state_bbox$lat_min, state_bbox$lat_max)
    ) +
    labs(
      title = "Approximate spatial distribution of observations in Bayern",
      subtitle = "Boundary file unavailable; points are shown within an approximate Bayern bounding box.",
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_gtree()
}
```

## Bayern districts: observations and species

The district analysis uses **NUTS3 regions** in Bayern. In Germany these correspond approximately to Kreise and kreisfreie Städte.

```{r district-join}
if (!is.null(state_districts) && nrow(state_districts) > 0) {
  state_obs_district_sf <- sf::st_join(
    state_obs_sf,
    state_districts |> select(district_id, district_name, geometry),
    join = sf::st_within,
    left = TRUE
  )

  state_obs_district <- sf::st_drop_geometry(state_obs_district_sf) |>
    left_join(
      common_names,
      by = c("species_key_lookup", "level3_species" = "scientific_name")
    ) |>
    mutate(
      display_name = case_when(
        !is.na(common_name_de) & common_name_de != "" ~ paste0(common_name_de, " (", level3_species, ")"),
        !is.na(common_name_en) & common_name_en != "" ~ paste0(common_name_en, " (", level3_species, ")"),
        TRUE ~ level3_species
      )
    )

  district_summary <- state_obs_district |>
    filter(!is.na(district_name), district_name != "") |>
    group_by(district_id, district_name) |>
    summarise(
      observations = n(),
      species = n_distinct(level3_species),
      sources = n_distinct(source),
      .groups = "drop"
    ) |>
    arrange(desc(observations))
} else {
  state_obs_district_sf <- NULL
  state_obs_district <- NULL
  district_summary <- tibble::tibble()
}

if (nrow(district_summary) == 0) {
  nice_table(tibble::tibble(message = "No district-level boundary data available. District plots are skipped."), "District status")
} else {
  district_summary |>
    mutate(observations = format(observations, big.mark = ",")) |>
    head(15) |>
    nice_table("Top 15 Bayern districts by observation count")
}
```

```{r district-barplot, fig.width=8, fig.height=7}
if (nrow(district_summary) > 0) {
  district_summary |>
    slice_head(n = 20) |>
    mutate(district_name = reorder(district_name, observations)) |>
    ggplot(aes(x = observations, y = district_name)) +
    geom_col() +
    scale_x_continuous(labels = comma) +
    labs(
      title = "Bayern observations by district",
      subtitle = "NUTS3 districts with the highest number of observations.",
      x = "Number of observations",
      y = NULL
    ) +
    theme_gtree()
}
```

```{r district-species-barplot, fig.width=8, fig.height=7}
if (nrow(district_summary) > 0) {
  district_summary |>
    arrange(desc(species)) |>
    slice_head(n = 20) |>
    mutate(district_name = reorder(district_name, species)) |>
    ggplot(aes(x = species, y = district_name)) +
    geom_col() +
    scale_x_continuous(labels = comma) +
    labs(
      title = "Recorded species by Bayern district",
      subtitle = "Districts with the highest number of distinct recorded species.",
      x = "Number of recorded species",
      y = NULL
    ) +
    theme_gtree()
}
```

```{r district-map, fig.width=8, fig.height=7}
if (!is.null(state_districts) && nrow(district_summary) > 0) {
  district_map_data <- state_districts |>
    left_join(district_summary, by = c("district_id", "district_name")) |>
    mutate(observations = coalesce(observations, 0L))

  ggplot(district_map_data) +
    geom_sf(aes(fill = observations), color = "white", linewidth = 0.15) +
    scale_fill_viridis_c(labels = comma, option = "C", trans = "sqrt") +
    labs(
      title = "Observation frequency across Bayern districts",
      subtitle = "Darker districts contain more GlobalGeoTree observations. Square-root scaling improves readability.",
      fill = "Observations",
      x = NULL,
      y = NULL
    ) +
    theme_gtree()
}
```

## Sources on the Bayern map

```{r source-map, fig.width=9, fig.height=7}
top_sources <- state_obs_named |>
  count(source, sort = TRUE) |>
  slice_head(n = 4) |>
  pull(source)

source_points <- state_obs_sf |>
  filter(source %in% top_sources) |>
  sample_points_for_plot(max_points_on_map)

if (!is.null(state_boundary) && nrow(state_boundary) > 0 && length(top_sources) > 0) {
  ggplot() +
    geom_sf(data = state_boundary, fill = "gray96", color = "gray55", linewidth = 0.3) +
    geom_point(
      data = source_points,
      aes(x = longitude, y = latitude),
      alpha = 0.25,
      size = 0.45
    ) +
    facet_wrap(~ source) +
    coord_sf() +
    labs(
      title = "Spatial patterns of the largest observation sources in Bayern",
      subtitle = "Facets show the four largest source groups. Points are sampled for readability.",
      x = NULL,
      y = NULL
    ) +
    theme_gtree(base_size = 11)
}
```

## Top recorded species with common names

```{r top-species-table}
top_species <- state_obs_named |>
  count(level3_species, common_name_de, common_name_en, display_name, sort = TRUE, name = "observations") |>
  mutate(share = observations / sum(observations)) |>
  select(
    display_name,
    scientific_name = level3_species,
    german_common_name = common_name_de,
    english_common_name = common_name_en,
    observations,
    share
  )

top_species |>
  head(20) |>
  mutate(
    observations = format(observations, big.mark = ","),
    share = percent(share, accuracy = 0.1)
  ) |>
  nice_table("Top 20 recorded species in Bayern")
```

```{r top-species-plot, fig.width=8, fig.height=7}
top_species |>
  slice_head(n = 15) |>
  mutate(display_name = reorder(display_name, observations)) |>
  ggplot(aes(x = observations, y = display_name)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Most frequently recorded tree species in Bayern",
    subtitle = "Common names are shown where available; scientific names remain the primary identifier.",
    x = "Number of observations",
    y = NULL
  ) +
  theme_gtree()
```

## Category composition, if available

```{r category-summary}
if ("category" %in% names(state_obs_named) && sum(!is.na(state_obs_named$category)) > 0) {
  category_summary <- state_obs_named |>
    filter(!is.na(category), category != "") |>
    group_by(category) |>
    summarise(
      observations = n(),
      species = n_distinct(level3_species),
      .groups = "drop"
    ) |>
    mutate(observation_share = observations / sum(observations)) |>
    arrange(desc(observations))

  category_summary |>
    mutate(
      observations = format(observations, big.mark = ","),
      observation_share = percent(observation_share, accuracy = 0.1)
    ) |>
    nice_table("Rare/Common/Frequent coverage in Bayern, where known")
} else {
  nice_table(tibble::tibble(message = "No category labels are available for Bayern observations in the current files."), "Category status")
}
```

```{r category-plot, fig.width=7, fig.height=4.5}
if ("category" %in% names(state_obs_named) && sum(!is.na(state_obs_named$category)) > 0) {
  category_summary |>
    ggplot(aes(x = category, y = observations)) +
    geom_col() +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Known category composition of Bayern observations",
      subtitle = "This only covers species that could be matched to a category lookup.",
      x = "Dataset-defined category",
      y = "Number of observations"
    ) +
    theme_gtree()
}
```

## Interpretation notes

The Bayern subset gives a regional view of GlobalGeoTree observations. The observation map and district-level charts describe where data are concentrated, but they should not be interpreted directly as true tree abundance. High counts may reflect a combination of tree presence, citizen-science activity, institutional data availability, and source-specific sampling behavior.

The common-name enrichment is useful for making the report more readable. However, common names are incomplete and not standardized across all languages or regions. For reproducibility, all analysis should continue to use scientific names or species keys as the stable identifiers.
