---
title: "Germany Check: Full Dataset, Species Counts, and Category Lookup"
author: "GlobalGeoTree team"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
---

# Purpose

This short diagnostic report checks Germany in the GlobalGeoTree observations file.

It answers three questions:

1. How many observations are available for Germany?
2. How many recorded species are available for Germany?
3. How many Germany observations and species can be assigned to the dataset-defined categories `Rare`, `Common`, and `Frequent`?

The category check is especially important because the full observations dataset may not contain a `Category` column. In that case, this report uses the larger evaluation dataset `GlobalGeoTree-10kEval-900.csv` as a category lookup table.

# Setup

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)

options(scipen = 999)

focus_country_code <- "DE"
focus_country_name <- "Germany"

# Main observations file. Prefer the full dataset if available.
observation_candidates <- c(
  "GlobalGeoTree_full.csv",
  "GlobalGeoTree.csv",
  "GlobalGeoTree-6m.csv",
  "GlobalGeoTree-10kEval-900.csv",
  "GlobalGeoTree-10kEval-90.csv"
)

observations_path <- observation_candidates[file.exists(observation_candidates)][1]

if (is.na(observations_path)) {
  stop("No observations file found. Put GlobalGeoTree_full.csv or an eval CSV in the same folder as this QMD.")
}

# Category lookup. Prefer the larger 900-species eval dataset.
eval_candidates <- c(
  "GlobalGeoTree-10kEval-900.csv",
  "GlobalGeoTree-10kEval-90.csv"
)

eval_lookup_path <- eval_candidates[file.exists(eval_candidates)][1]

# Optional taxonomy catalog. It is not used for category, but can help verify taxonomy.
catalog_candidates <- c(
  "Tree_species_catalog.csv",
  "Tree_species_catalog(1).csv"
)

catalog_path <- catalog_candidates[file.exists(catalog_candidates)][1]

theme_gtree <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(color = "gray35"),
      plot.caption = element_text(color = "gray45", size = base_size - 2),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )
}

fmt_int <- function(x) format(x, big.mark = ",", scientific = FALSE)
```

# Input files

```{r input-files}
input_files <- tibble::tibble(
  file_role = c("Main observations file", "Category lookup file", "Optional species catalog"),
  file_path = c(
    observations_path,
    ifelse(is.na(eval_lookup_path), "Not found", eval_lookup_path),
    ifelse(is.na(catalog_path), "Not found", catalog_path)
  )
)

knitr::kable(input_files, caption = "Files used in this diagnostic report")
```

# Load data and inspect the head

```{r load-data}
obs_raw <- fread(observations_path, showProgress = FALSE)

# Standardize column names only where needed.
old_names <- names(obs_raw)
names(obs_raw) <- str_to_lower(names(obs_raw))

# The full dataset and eval files may use Category or category.
if ("category" %in% names(obs_raw)) {
  obs_raw <- obs_raw |> mutate(category = str_squish(as.character(category)))
}

required_observation_cols <- c("country_code", "level3_species", "species_key")
missing_required <- setdiff(required_observation_cols, names(obs_raw))

if (length(missing_required) > 0) {
  stop("The observations file is missing required columns: ", paste(missing_required, collapse = ", "))
}

obs_clean <- obs_raw |>
  mutate(
    country_code = str_to_upper(str_squish(as.character(country_code))),
    level3_species = str_squish(as.character(level3_species)),
    species_key = as.character(species_key)
  )

if ("latitude" %in% names(obs_clean)) {
  obs_clean <- obs_clean |> mutate(latitude = as.numeric(latitude))
}

if ("longitude" %in% names(obs_clean)) {
  obs_clean <- obs_clean |> mutate(longitude = as.numeric(longitude))
}

obs_de <- obs_clean |>
  filter(country_code == focus_country_code)
```

The table below shows the first rows of the main observations dataset. This is useful for checking which variables are available before any filtering or joins.

```{r head-full}
obs_clean |>
  head(10) |>
  knitr::kable(caption = "First 10 rows of the main observations dataset")
```

The table below shows the first rows after filtering to Germany.

```{r head-germany}
obs_de |>
  head(10) |>
  knitr::kable(caption = "First 10 Germany rows after filtering country_code == 'DE'")
```

# Germany observation and species counts

```{r germany-basic-counts}
germany_observations <- nrow(obs_de)
germany_species <- n_distinct(obs_de$level3_species)
germany_species_keys <- n_distinct(obs_de$species_key)

basic_counts <- tibble::tibble(
  metric = c(
    "Germany observations",
    "Germany recorded species names",
    "Germany recorded species keys"
  ),
  value = c(
    germany_observations,
    germany_species,
    germany_species_keys
  )
)

knitr::kable(
  basic_counts |> mutate(value = fmt_int(value)),
  caption = "Germany subset size before category lookup"
)
```

The Germany subset contains **`r fmt_int(germany_observations)` observations** and **`r fmt_int(germany_species)` recorded species names**.

# Category source and lookup logic

The category variable is handled in this order:

1. If the main observations file already contains `category`, use it directly.
2. If the main observations file does not contain `category`, use the evaluation dataset as a lookup table.
3. Prefer joining by `species_key` because it is more stable than species names.
4. If `species_key` is not available, fall back to joining by `level3_species`.

The species catalog is **not** used to assign `Rare`, `Common`, or `Frequent`, because the catalog does not normally contain category labels. It is a taxonomy reference, not a category lookup.

```{r category-lookup}
main_has_category <- "category" %in% names(obs_clean)

if (main_has_category) {
  obs_de_cat <- obs_de |>
    mutate(category_source = "Main observations file")
  
  category_source_note <- "The main observations file already contains a category column. No eval lookup was required."
  lookup_summary <- tibble::tibble(
    metric = c("Lookup file used", "Lookup species with category"),
    value = c("Not needed", "Not needed")
  )
  
} else {
  if (is.na(eval_lookup_path)) {
    stop("The main observations file has no category column and no eval lookup file was found.")
  }
  
  eval_raw <- fread(eval_lookup_path, showProgress = FALSE)
  names(eval_raw) <- str_to_lower(names(eval_raw))
  
  required_eval_cols <- c("level3_species", "species_key", "category")
  missing_eval <- setdiff(required_eval_cols, names(eval_raw))
  
  if (length(missing_eval) > 0) {
    stop("The eval lookup file is missing required columns: ", paste(missing_eval, collapse = ", "))
  }
  
  eval_lookup <- eval_raw |>
    mutate(
      level3_species = str_squish(as.character(level3_species)),
      species_key = as.character(species_key),
      category = str_squish(as.character(category))
    ) |>
    filter(!is.na(category), category != "") |>
    distinct(species_key, level3_species, category)
  
  # Check whether any species_key has conflicting categories.
  conflicting_keys <- eval_lookup |>
    count(species_key, name = "n_rows") |>
    filter(n_rows > 1)
  
  # Resolve category by species_key first.
  lookup_by_key <- eval_lookup |>
    group_by(species_key) |>
    summarise(
      category = dplyr::first(category),
      category_species_name = dplyr::first(level3_species),
      .groups = "drop"
    )
  
  # Fallback lookup by species name.
  lookup_by_species <- eval_lookup |>
    group_by(level3_species) |>
    summarise(
      category_by_species_name = dplyr::first(category),
      .groups = "drop"
    )
  
  obs_de_cat <- obs_de |>
    left_join(lookup_by_key, by = "species_key") |>
    left_join(lookup_by_species, by = "level3_species") |>
    mutate(
      category = coalesce(category, category_by_species_name),
      category_join_method = case_when(
        !is.na(category) & !is.na(category_species_name) ~ "Matched by species_key",
        !is.na(category) & is.na(category_species_name) ~ "Matched by species name",
        TRUE ~ "No category match"
      ),
      category_source = paste("Eval lookup:", eval_lookup_path)
    ) |>
    select(-category_by_species_name)
  
  category_source_note <- paste(
    "The main observations file does not contain category. Categories were joined from",
    eval_lookup_path,
    "using species_key first and species name as fallback."
  )
  
  lookup_summary <- tibble::tibble(
    metric = c(
      "Lookup file used",
      "Lookup rows",
      "Lookup species keys with category",
      "Lookup species names with category",
      "Species keys with conflicting category rows"
    ),
    value = c(
      eval_lookup_path,
      fmt_int(nrow(eval_lookup)),
      fmt_int(n_distinct(eval_lookup$species_key)),
      fmt_int(n_distinct(eval_lookup$level3_species)),
      fmt_int(nrow(conflicting_keys))
    )
  )
}

knitr::kable(lookup_summary, caption = "Category lookup source and coverage")
```

`r category_source_note`

# Category coverage in Germany

```{r category-coverage}
known_category_obs <- obs_de_cat |>
  filter(!is.na(category), category != "")

unknown_category_obs <- obs_de_cat |>
  filter(is.na(category) | category == "")

known_category_species <- known_category_obs |>
  distinct(level3_species) |>
  nrow()

unknown_category_species <- obs_de_cat |>
  filter(is.na(category) | category == "") |>
  distinct(level3_species) |>
  nrow()

coverage_summary <- tibble::tibble(
  metric = c(
    "Germany observations",
    "Germany recorded species",
    "Germany observations with known category",
    "Germany recorded species with known category",
    "Germany observations without known category",
    "Germany recorded species without known category"
  ),
  value = c(
    nrow(obs_de_cat),
    n_distinct(obs_de_cat$level3_species),
    nrow(known_category_obs),
    known_category_species,
    nrow(unknown_category_obs),
    unknown_category_species
  ),
  share = c(
    1,
    1,
    nrow(known_category_obs) / nrow(obs_de_cat),
    known_category_species / n_distinct(obs_de_cat$level3_species),
    nrow(unknown_category_obs) / nrow(obs_de_cat),
    unknown_category_species / n_distinct(obs_de_cat$level3_species)
  )
)

coverage_summary_display <- coverage_summary |>
  mutate(
    value = fmt_int(value),
    share = scales::percent(share, accuracy = 0.1)
  )

knitr::kable(coverage_summary_display, caption = "Known and unknown category coverage for Germany")
```

Interpretation:

- `Germany observations with known category` counts observation rows whose species could be matched to `Rare`, `Common`, or `Frequent`.
- `Germany recorded species with known category` counts unique species names in Germany whose category is known.
- If the full dataset does not contain `category`, these values depend on how much overlap exists between Germany species and the eval lookup file.

# Rare, Common, and Frequent relation

```{r category-observation-summary}
category_obs_summary <- known_category_obs |>
  count(category, name = "observations") |>
  mutate(share_observations = observations / sum(observations)) |>
  arrange(desc(observations))

knitr::kable(
  category_obs_summary |>
    mutate(
      observations = fmt_int(observations),
      share_observations = scales::percent(share_observations, accuracy = 0.1)
    ),
  caption = "Germany observations by known category"
)
```

```{r category-species-summary}
category_species_summary <- known_category_obs |>
  distinct(category, level3_species) |>
  count(category, name = "species") |>
  mutate(share_species = species / sum(species)) |>
  arrange(desc(species))

knitr::kable(
  category_species_summary |>
    mutate(
      species = fmt_int(species),
      share_species = scales::percent(share_species, accuracy = 0.1)
    ),
  caption = "Germany recorded species by known category"
)
```

```{r category-observation-chart, fig.width=7, fig.height=4.5}
ggplot(category_obs_summary, aes(x = reorder(category, observations), y = observations)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Germany observations with known category",
    subtitle = "Only species matched to the category lookup are included.",
    x = NULL,
    y = "Observation count",
    caption = "Category labels come from the main file if available; otherwise from the evaluation lookup."
  ) +
  theme_gtree()
```

```{r category-species-chart, fig.width=7, fig.height=4.5}
ggplot(category_species_summary, aes(x = reorder(category, species), y = species)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Germany species with known category",
    subtitle = "This chart counts species, not observation rows.",
    x = NULL,
    y = "Number of species",
    caption = "If many Germany species are unmatched, this chart should be interpreted as partial coverage only."
  ) +
  theme_gtree()
```

# Which Germany species received a category?

```{r matched-species-table}
matched_species <- known_category_obs |>
  count(level3_species, species_key, category, name = "observations") |>
  arrange(category, desc(observations))

knitr::kable(
  matched_species |>
    mutate(observations = fmt_int(observations)),
  caption = "Germany species with known category"
)
```

# Which Germany species are still unmatched?

```{r unmatched-species-table}
unmatched_species <- unknown_category_obs |>
  count(level3_species, species_key, name = "observations") |>
  arrange(desc(observations))

knitr::kable(
  unmatched_species |>
    head(30) |>
    mutate(observations = fmt_int(observations)),
  caption = "Top 30 Germany species without a known category"
)
```

# Conclusion

This diagnostic report separates two questions that are easy to confuse:

1. **How many observations and species exist for Germany in the main dataset?**  
   This is calculated directly from the full observations file after filtering `country_code == 'DE'`.

2. **How many Germany observations and species have a known `Rare`, `Common`, or `Frequent` category?**  
   This depends on whether the main file contains `category`. If it does not, category can only be assigned to species that overlap with the eval lookup dataset.

Therefore, if the category comes from `GlobalGeoTree-10kEval-900.csv`, the Rare/Common/Frequent analysis is a **matched-subset analysis**, not a complete classification of all Germany species.
