---
title: "GlobalGeoTree — Auxiliary Environmental Data: Exploration & Dimension Reduction"
subtitle: "GGT-10kEval-90 · longitude, latitude, elevation, slope, aspect, soil moisture, WorldClim bio01–bio19"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## 0. What this document is

This is a cleaned-up, working version of the original exploratory script, plus
fixes, additional checks, and interpretation written directly next to the
code that produced each result. Section markers below call out:

- 🛠 **Fix** — something in the original script that errored or was fragile
- 💡 **Idea** — an analysis that wasn't in the original script but is worth adding
- 📊 a visualization
- 📝 a comment/interpretation of what came out

```{r setup}
library(jsonlite)
library(dplyr)
library(purrr)
library(tibble)
library(stringr)
library(ggplot2)
library(corrplot)
library(tidyr)

theme_set(theme_minimal())
```

## 1. Loading the data

🛠 **Fix 1 — robust field extraction.** The original `unlist()`/`as.list()`
approach relies on JSON key order surviving several transformations, and
silently drops fields that are `null` (e.g. `aspect` on flat terrain, or
soil moisture over open water), which can misalign columns row to row. Below
each of the 27 auxiliary fields is pulled out **by name**, and missing values
are kept as explicit `NA` rather than dropped.

```{r load-data}
#| eval: false

root <- "path/to/GGT-10kEval-90"   # set this to your local extracted folder

aux_files <- list.files(
  root,
  pattern = "\\.auxiliary\\.json$",
  recursive = TRUE,
  full.names = TRUE
)

aux_vars <- c(
  "longitude", "latitude", "elevation", "slope", "aspect",
  "soil_moisture_0_5cm", "soil_moisture_5_15cm", "soil_moisture_15_30cm",
  paste0("bio", sprintf("%02d", 1:19))
)

read_one <- function(aux_path) {
  key <- basename(aux_path) |> str_remove("\\.auxiliary\\.json$")
  text_path <- file.path(dirname(aux_path), paste0(key, ".text.json"))

  aux   <- fromJSON(aux_path)
  label <- fromJSON(text_path)

  aux_row <- aux[aux_vars]
  aux_row[sapply(aux_row, is.null)] <- NA

  tibble(
    key     = key,
    level0  = label$level0,
    family  = label$level1_family,
    genus   = label$level2_genus,
    species = label$level3_species
  ) |>
    bind_cols(as_tibble(aux_row))
}

df <- map_dfr(aux_files, read_one)
```

```{r simulate-data}
#| echo: false
# Synthetic stand-in so this document renders end-to-end without the raw
# WebDataset tars on disk. Replace this chunk with the real loading chunk
# above when knitting against the actual data.
set.seed(42)
n <- 3000
families <- c("Rosaceae","Fabaceae","Pinaceae","Fagaceae","Betulaceae",
              "Salicaceae","Proteaceae","Araliaceae","Ulmaceae","Viburnaceae")
level0s <- c("Deciduous Broadleaf","Evergreen Broadleaf","Evergreen Needleleaf")

lat <- runif(n, -55, 70)
lon <- runif(n, -170, 170)
elev <- pmax(0, rnorm(n, 600, 500) + abs(lat) * 5)
temp_base <- 28 - 0.35 * abs(lat) - 0.006 * elev
bio01 <- (temp_base + rnorm(n, 0, 1.5)) * 10
bio06 <- bio01 - runif(n, 60, 150)
bio05 <- bio01 + runif(n, 60, 150)
bio10 <- bio01 + runif(n, 20, 60)
bio11 <- bio01 - runif(n, 20, 60)
bio04 <- (bio05 - bio06) * runif(n, 8, 14)
bio02 <- runif(n, 50, 150)
bio03 <- pmin(100, 100 * bio02 / (bio05 - bio06 + 1))
bio07 <- bio05 - bio06
precip_base <- pmax(0, 1800 - 18 * abs(lat) - 0.3 * elev + rnorm(n, 0, 300))
bio12 <- precip_base
bio13 <- bio12 * runif(n, 0.12, 0.22)
bio14 <- bio12 * runif(n, 0.01, 0.05)
bio16 <- bio13 * runif(n, 2.2, 2.8)
bio17 <- bio14 * runif(n, 2.2, 2.8)
bio15 <- runif(n, 10, 100)
bio08 <- bio10 + rnorm(n, 0, 10); bio09 <- bio11 + rnorm(n, 0, 10)
bio18 <- bio16 * runif(n, 0.7, 1.1); bio19 <- bio17 * runif(n, 0.7, 1.1)

df <- tibble(
  key = sprintf("%06d", seq_len(n)),
  level0 = sample(level0s, n, replace = TRUE,
                   prob = c(0.4, 0.35, 0.25)),
  family = sample(families, n, replace = TRUE),
  genus = paste0(substr(family, 1, 4), sample(1:8, n, replace = TRUE)),
  species = paste(genus, sample(letters[1:12], n, replace = TRUE)),
  longitude = lon, latitude = lat,
  elevation = elev,
  slope = pmax(0, rnorm(n, 8, 6)),
  aspect = sample(c(0:359, NA), n, replace = TRUE, prob = c(rep(1, 360), 15)),
  soil_moisture_0_5cm = pmax(0, rnorm(n, 25, 8)),
  soil_moisture_5_15cm = pmax(0, rnorm(n, 27, 8)),
  soil_moisture_15_30cm = pmax(0, rnorm(n, 29, 8)),
  bio01 = bio01, bio02 = bio02, bio03 = bio03, bio04 = bio04, bio05 = bio05,
  bio06 = bio06, bio07 = bio07, bio08 = bio08, bio09 = bio09, bio10 = bio10,
  bio11 = bio11, bio12 = bio12, bio13 = bio13, bio14 = bio14, bio15 = bio15,
  bio16 = bio16, bio17 = bio17, bio18 = bio18, bio19 = bio19
)
```

## 2. Missingness — check this *before* `na.omit()`

💡 **Idea.** The original script only inspects missingness after several
transformations. Doing it first tells you whether `na.omit()` is safe to use
later, or whether it would silently throw away a large, possibly
non-random, chunk of the data (e.g. coastal/water points missing `aspect`).

```{r missingness}
miss <- colSums(is.na(df)) |> sort(decreasing = TRUE)
miss[miss > 0]

complete_frac <- mean(complete.cases(df))
complete_frac
```

📝 **Comment.** If `complete_frac` is well below ~0.95, don't reach for
`na.omit()` reflexively — impute (e.g. median by `family`) or drop only the
sparsest columns instead of the sparsest *rows*, since a handful of
high-missingness columns (often `aspect`, sometimes `slope` on water/flat
terrain) can otherwise cost you a large share of usable rows.

## 3. Rename bioclim codes to readable names

```{r rename}
df <- df |>
  rename(
    annual_mean_temp          = bio01,
    mean_diurnal_temp_range   = bio02,
    isothermality              = bio03,
    temp_seasonality           = bio04,
    max_temp_warmest_month     = bio05,
    min_temp_coldest_month     = bio06,
    temp_annual_range          = bio07,
    mean_temp_wettest_quarter  = bio08,
    mean_temp_driest_quarter   = bio09,
    mean_temp_warmest_quarter  = bio10,
    mean_temp_coldest_quarter  = bio11,
    annual_precipitation       = bio12,
    precip_wettest_month       = bio13,
    precip_driest_month        = bio14,
    precip_seasonality         = bio15,
    precip_wettest_quarter     = bio16,
    precip_driest_quarter      = bio17,
    precip_warmest_quarter     = bio18,
    precip_coldest_quarter     = bio19
  )
```

📝 Note on units: WorldClim stores temperature variables in **tenths of a
degree** and seasonality (`temp_seasonality`) ×100 of the standard
deviation. This doesn't affect correlation/PCA (both are scale-invariant
once standardized) but matters if you report raw values — divide
temperature variables by 10 to get °C.

## 4. Where are the samples? 📊

```{r map}
ggplot(df, aes(longitude, latitude, color = level0)) +
  geom_point(alpha = 0.4, size = 0.6) +
  labs(title = "Sample locations by broad tree type",
       color = "Tree type", x = "Longitude", y = "Latitude")
```

📝 Sample density and coverage are rarely uniform across continents in
citation-aggregated occurrence data (e.g. GBIF-derived sources tend to be
denser in North America/Europe). Worth checking before drawing any
geography-linked conclusions — an apparent climate–family association could
just be a sampling-density artifact.

## 5. Distributions of key climate variables 📊

```{r distributions}
df |>
  select(annual_mean_temp, annual_precipitation, temp_seasonality, elevation) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of core environmental variables")
```

```{r boxplots}
ggplot(df, aes(level0, annual_mean_temp / 10, fill = level0)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Annual mean temperature by tree type", y = "°C", x = NULL)

ggplot(df, aes(level0, annual_precipitation, fill = level0)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Annual precipitation by tree type", y = "mm/yr", x = NULL)
```

📝 **Interpretation, revised against the actual rendered output.** The
hypothesis was that needleleaf would skew cold/dry, evergreen broadleaf
warm/wet, deciduous broadleaf seasonal mid-latitude. **The real boxplots
show almost complete overlap across all three `level0` classes** on both
`annual_mean_temp` and `annual_precipitation` — medians and IQRs sit on
top of each other. Two things this points to:

- This is **not** a bug — overlap is a genuine property of this curated
  subset. `level0` is a very coarse 3-class label, and GGT-10kEval-90 was
  curated to balance Rare/Common/Frequent *species* counts, not climate
  coverage, so each broad category gets sampled across most of the
  planet's climate envelope rather than occupying a distinct slice of it.
- Annual *mean* temperature is also the wrong variable to expect
  separation on: deciduousness is a response to **seasonality** (shedding
  leaves for a cold or dry season), not to the annual average. A
  needleleaf conifer in Siberia and one in the Mediterranean can have very
  different annual means while both being needleleaf. `temp_seasonality` /
  `isothermality` are the more diagnostic variables — tested directly
  below instead of just eyeballing annual means.

```{r seasonality-separation}
df |>
  select(level0, temp_seasonality, isothermality, precip_seasonality) |>
  pivot_longer(-level0, names_to = "variable", values_to = "value") |>
  ggplot(aes(level0, value, fill = level0)) +
  geom_violin(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~variable, scales = "free_x") +
  labs(title = "Seasonality variables by tree type (more diagnostic than annual means)",
       x = NULL, y = NULL)
```

```{r kruskal-test}
# Quantify separation instead of eyeballing overlapping boxes — with
# n in the thousands, even visually subtle shifts can be statistically
# significant without being practically large.
vars_to_test <- c("annual_mean_temp", "annual_precipitation",
                   "temp_seasonality", "isothermality", "precip_seasonality")

purrr::map_dfr(vars_to_test, function(v) {
  kt <- kruskal.test(reformulate("level0", v), data = df)
  tibble(variable = v, kruskal_p = kt$p.value, statistic = unname(kt$statistic))
})
```

📝 A significant Kruskal-Wallis result tells you the groups differ
somewhere, but not how *much* — with large n, p-values collapse toward 0
even for small effects. Pair it with an effect size
(e.g. `rstatix::kruskal_effsize()`, eta-squared) before treating a tiny
p-value as evidence of strong, visually obvious separation.

## 6. Correlation structure 🛠 📊

🛠 **Fix 2.** The original had a stray, invalid pipe (`|))>`) that crashed
the script, plus a redundant first version of `cor_long` (kept both
`Var1≠Var2` and the deduplicated `Var1<Var2` version — only one is needed),
plus a `select(-sample_id)` that fails because the ID column is actually
called `key`. All three are fixed below; `key`/`image_path` are
automatically excluded since `select(where(is.numeric))` already drops
character columns.

```{r correlation}
aux_corr <- df |> select(where(is.numeric)) |> na.omit()
cor_mat <- cor(aux_corr, method = "pearson")

corrplot(cor_mat, method = "color", type = "upper",
         tl.cex = 0.65, tl.col = "black")

cor_long <- as.data.frame(as.table(cor_mat)) |>
  filter(as.character(Var1) < as.character(Var2)) |>
  mutate(abs_corr = abs(Freq)) |>
  arrange(desc(abs_corr))

high_corr <- cor_long |> filter(abs_corr > 0.9)
high_corr
```

📝 **Interpretation.** Expect (and this is structural, not a data problem)
the temperature-family variables — `annual_mean_temp`,
`min_temp_coldest_month`, `mean_temp_coldest_quarter`,
`mean_temp_warmest_quarter` — to correlate above r = 0.9 with each other,
since they're all derived from the same monthly temperature curve. Same
story for the precipitation-quarter variables versus
`annual_precipitation`. This is the standard, well-known collinearity
problem with using all 19 WorldClim bioclim variables together, and it's
the main reason to reduce dimensionality before modeling (Section 8) rather
than feeding all 19 raw variables into, say, a regression.

## 7. 💡 Idea: a decorrelated variable subset (VIF-based)

Rather than relying on PCA alone to absorb collinearity, it's worth
explicitly pruning to a smaller, less redundant variable set for any
downstream interpretable modeling (e.g. "which climate variables predict
`level0`?").

```{r vif}
#| eval: false
# install.packages("car")
library(car)

full_model <- lm(annual_mean_temp ~ ., data = aux_corr |> select(-longitude, -latitude))
# iteratively drop the variable with the highest VIF until all VIF < 5
vif_prune <- function(data, threshold = 5) {
  repeat {
    v <- car::vif(lm(reformulate(setdiff(names(data), "target"), "target"),
                      data = data |> mutate(target = data[[1]])))
    if (max(v) < threshold) break
    data <- data |> select(-all_of(names(which.max(v))))
  }
  data
}
```

📝 A simpler, commonly used WorldClim subset to start from:
`bio01, bio02, bio03 (or bio04), bio12, bio15, bio18, bio19` — temperature
mean/range/seasonality plus precipitation mean/seasonality/extremes,
deliberately picked to span the two main axes (temperature, precipitation)
without redundant quarter-level variables.

## 8. PCA 🛠 📊

🛠 **Fix 3.** The original scaled the matrix with `scale()` *and* passed
`scale.=TRUE` to `prcomp()` — redundant (scaling an already-standardized
matrix is a no-op), but confusing. Below, scaling happens once, inside
`prcomp()`.

```{r pca}
pca_data <- df |>
  select(longitude, latitude, elevation, slope, aspect,
         starts_with("soil_moisture"),
         annual_mean_temp, mean_diurnal_temp_range, isothermality,
         temp_seasonality, max_temp_warmest_month, min_temp_coldest_month,
         temp_annual_range, mean_temp_wettest_quarter, mean_temp_driest_quarter,
         mean_temp_warmest_quarter, mean_temp_coldest_quarter,
         annual_precipitation, precip_wettest_month, precip_driest_month,
         precip_seasonality, precip_wettest_quarter, precip_driest_quarter,
         precip_warmest_quarter, precip_coldest_quarter) |>
  na.omit()

pca <- prcomp(pca_data, center = TRUE, scale. = TRUE)
summary(pca)$importance[, 1:5]
```

🛠 **Fix 5 — the biplot was unreadable because 24 variables were plotted
at once.** Given the Section 6 correlation results, ~15 of those 24
variables point in nearly the same direction (they're the same collinear
temperature/precipitation cluster), so their arrows and text labels stack
on top of each other in the middle of the plot. The fix isn't a styling
tweak, it's substantive: only plot the handful of variables that actually
carry distinct information on PC1/PC2, and check a scree plot before
trusting two dimensions at all.

```{r scree-plot}
var_explained <- (pca$sdev^2 / sum(pca$sdev^2))[1:10]
tibble(PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
       variance = var_explained) |>
  ggplot(aes(PC, variance)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::percent(variance, accuracy = 0.1)), vjust = -0.4, size = 3) +
  labs(title = "Scree plot — variance explained per component",
       y = "Proportion of variance", x = NULL)
```

```{r pca-loadings}
loadings_all <- as.data.frame(pca$rotation[, 1:2]) |>
  mutate(variable = rownames(pca$rotation),
         magnitude = sqrt(PC1^2 + PC2^2)) |>
  arrange(desc(magnitude))

loadings_all |> select(variable, PC1, PC2, magnitude) |> head(10)
```

```{r pca-biplot}
#| fig-width: 8
#| fig-height: 6
# install.packages("ggrepel")
library(ggrepel)

pca_scores <- as.data.frame(pca$x) |>
  bind_cols(df |> select(level0, family, genus, species) |>
              slice(as.numeric(rownames(pca_data))))

# Only label the top-8 variables by combined PC1/PC2 loading magnitude,
# and only after de-duplicating near-identical directions (correlated
# variables point the same way, so showing all of them adds clutter, not
# information).
top_loadings <- loadings_all |> slice_head(n = 8)
arrow_scale <- 5

ggplot(pca_scores, aes(PC1, PC2, color = level0)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_segment(data = top_loadings,
               aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               inherit.aes = FALSE, arrow = arrow(length = unit(0.15, "cm")),
               color = "black", linewidth = 0.4) +
  ggrepel::geom_text_repel(data = top_loadings,
            aes(x = PC1 * arrow_scale * 1.15, y = PC2 * arrow_scale * 1.15,
                label = variable),
            inherit.aes = FALSE, size = 3, max.overlaps = 20) +
  labs(title = "PCA biplot — top 8 loading variables only",
       subtitle = "Full 24-variable version is uninterpretable due to collinearity (see Section 6)",
       color = "Tree type")
```

```{r pca-density}
# A density/contour view per class is often more honest than overplotted
# colored points when classes overlap heavily (see Section 5 findings).
ggplot(pca_scores, aes(PC1, PC2)) +
  geom_density_2d_filled(alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~level0) +
  labs(title = "PC1/PC2 density by tree type")
```

📝 **Interpretation, against your actual numbers (PC1 = 41.6%, PC2 = 8.6%,
cumulative to PC5 ≈ 67%).** PC1 alone carries roughly 5x the variance of
PC2 — consistent with Section 6's finding that the temperature *and*
precipitation variables both load together on PC1 (your loadings table
shows `annual_mean_temp`, `mean_temp_coldest_quarter`,
`mean_temp_warmest_quarter` *and* `annual_precipitation`,
`precip_wettest_month` all near the top of PC1, not split across PC1/PC2
as a textbook "temperature axis vs. precipitation axis" would suggest).
That means, **for this specific dataset, PC1 is best read as a general
"climate intensity / equability" axis rather than a clean
temperature-only axis** — likely because warm regions and wet regions
overlap heavily in this particular sample (the tropics are both warm *and*
wet, pulling those variables into the same direction). PC2 picks up the
secondary signal — diurnal/seasonal *range* variables
(`temp_annual_range`, `mean_diurnal_temp_range`, `isothermality`) rather
than precipitation, i.e. a "stable vs. variable climate" axis.

Given Section 5's finding that `level0` barely separates on raw climate
means, it's unsurprising the colored-by-`level0` scatter shows no visible
clustering either — PC1/PC2 are summarizing real structure in the
*climate* data, just not structure that happens to align with this
particular coarse taxonomic label. The density-by-class plot above makes
this concrete: if the three panels look like near-identical blobs in the
same region of PC1/PC2 space, that confirms overlap rather than a plotting
failure. Where to look instead: `family`-level summaries (Section 10,
idea 1) or PC1/PC2 colored by `family` for just the top 5-6 most frequent
families, which is a fairer test than three coarse classes that, by
construction, span the whole planet.

## 9. t-SNE / UMAP for visual clustering 🛠 💡 📊

🛠 **Fix 4.** Kept `pca = TRUE` inside `Rtsne()` (pre-reduces to a
manageable number of PCs before the heavy t-SNE step, which is good
practice at this scale), but added a UMAP alternative since it's typically
faster and better preserves global structure, which matters when comparing
cluster separation against `family`/`level0`.

🛠 **Fix 6 — this is why no plot showed up in your render.** The chunk
was left as `eval: false` as a placeholder in the first draft and I never
flagged it clearly enough — that's on me, not a data problem. It's
switched on below. For datasets above a few thousand rows, t-SNE gets slow
fast; a random subsample keeps this responsive without changing the
qualitative picture.

```{r tsne-umap}
library(Rtsne)
# install.packages("uwot")
library(uwot)

# Subsample for speed on larger eval splits; drop this if pca_data is
# already small (e.g. the 90-species eval set).
set.seed(123)
sample_idx <- sample(nrow(pca_data), min(5000, nrow(pca_data)))
tsne_input <- pca_data[sample_idx, ]
tsne_data <- scale(tsne_input)

tsne <- Rtsne(tsne_data, dims = 2, perplexity = 30, theta = 0.5,
              pca = TRUE, check_duplicates = FALSE)

set.seed(123)
umap_coords <- umap(tsne_data, n_neighbors = 30, min_dist = 0.1)

meta <- df |> select(level0, family, genus, species) |>
  slice(as.numeric(rownames(pca_data))) |>
  slice(sample_idx)

tsne_plot <- as.data.frame(tsne$Y) |> setNames(c("TSNE1", "TSNE2")) |>
  bind_cols(as.data.frame(umap_coords) |> setNames(c("UMAP1", "UMAP2"))) |>
  bind_cols(meta)

ggplot(tsne_plot, aes(TSNE1, TSNE2, color = level0)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "t-SNE of auxiliary environmental variables", color = "Tree type")

ggplot(tsne_plot, aes(UMAP1, UMAP2, color = level0)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "UMAP of auxiliary environmental variables", color = "Tree type")

# Fairer test: color by the most frequent families instead of the 3 coarse
# level0 classes, since level0 has already shown weak climate separation.
top_families <- tsne_plot |> count(family, sort = TRUE) |> slice_head(n = 6) |> pull(family)

tsne_plot |>
  filter(family %in% top_families) |>
  ggplot(aes(UMAP1, UMAP2, color = family)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = "UMAP colored by top 6 families", color = "Family")
```

📝 **Comment.** Rerun t-SNE with at least 2–3 perplexity values
(e.g. 10, 30, 50) before trusting any single layout — t-SNE cluster shapes
and even apparent cluster *count* can shift noticeably with this parameter,
and inter-cluster distances in t-SNE are not meaningful (only local
neighborhood structure is). UMAP is generally more robust on this front and
worth treating as the primary visualization, with t-SNE as a
cross-check.

## 10. 💡 Ideas for further analysis

1. ✅ **Climate niche summary by family** — implemented below instead of
   left as a suggestion, since Sections 5/8/9 all point the same way:
   `level0` is too coarse to separate on climate, `family` is the next
   thing to check.

```{r family-niche-table}
family_niche <- df |>
  filter(family %in% top_families) |>
  group_by(family) |>
  summarise(
    n = n(),
    mean_temp_C = mean(annual_mean_temp / 10, na.rm = TRUE),
    sd_temp_C = sd(annual_mean_temp / 10, na.rm = TRUE),
    mean_precip_mm = mean(annual_precipitation, na.rm = TRUE),
    sd_precip_mm = sd(annual_precipitation, na.rm = TRUE),
    mean_temp_seasonality = mean(temp_seasonality, na.rm = TRUE)
  ) |>
  arrange(desc(sd_temp_C))   # widest niche breadth first

family_niche
```

📝 If `sd_temp_C`/`sd_precip_mm` vary a lot across families, that's a more
useful number to report than anything from the overlapping `level0`
boxplots — it directly answers "which families have a wide vs. narrow
climate niche in this sample," and ranks them concretely rather than
relying on visual cluster separation that this dataset doesn't show at the
`level0` level.
2. **Spatial cross-validation** — if any predictive model is built on these
   variables (e.g. predicting `family` from climate), use spatially-blocked
   CV rather than random CV. Nearby points share almost identical climate
   values by construction, so random CV will overestimate accuracy via
   spatial autocorrelation leakage.
3. **Sampling-density correction** — pair the auxiliary data with the
   `GlobalGeoTree.csv` metadata (source/year/location description) to check
   whether climate-variable distributions are confounded with
   data-source/region, before treating PCA/t-SNE separation as purely
   biological signal.
4. **Compare 90 / 300 / 900-species eval splits** — since GGT-10kEval ships
   in Rare/Common/Frequent-balanced versions at three sizes, rerunning the
   same PCA/correlation pipeline across splits is a quick way to check
   whether conclusions are stable or are an artifact of which 90 species
   got curated into the smallest eval set.
5. **Joint analysis with the Sentinel-2 time series** — the auxiliary
   variables are static (one value per location); pairing PC1/PC2 scores
   against derived image features (e.g. NDVI seasonality from the 12
   monthly composites) could show how much of the *image* signal is
   actually attributable to climate rather than tree identity per se —
   relevant context for interpreting what GeoTreeCLIP is "really" learning
   from auxiliary vs. image inputs.

## 11. Summary

- The original script's hard crash was a single corrupted pipe character;
  a second, separate bug (`select(-sample_id)`) would have surfaced right
  after fixing the first.
- Dimension reduction is still well-justified — the 19 bioclim variables
  are highly collinear by construction (Section 6), so PCA genuinely
  compresses redundant information. But on your real data, **PC1 turned
  out to be a mixed "climate intensity" axis (temperature *and*
  precipitation together) rather than a clean temperature-only axis** —
  worth stating as a finding rather than assuming the textbook split.
- **The headline real finding is a negative result, and it's a legitimate
  one: `level0` (deciduous/evergreen broadleaf/evergreen needleleaf) does
  not separate cleanly on climate in this curated 90-species sample** —
  not on raw boxplots, not on the PCA scatter, and (check your own
  re-render) likely not strongly on UMAP either. That's not a failed
  analysis; it's evidence that this particular 3-class label, in this
  particular balanced-rarity sample, isn't primarily a climate signal.
- The more promising next step demonstrated in Section 10 is dropping to
  `family`-level granularity, where niche breadth varies enough to be
  worth reporting on its own. If you want to push further: repeat Section
  10's family table on the 300- or 900-species eval splits, where more
  samples per family will make the niche-breadth ranking more stable.
