---
title: "GlobalGeoTree — Auxiliary Environmental Data: Exploration & Dimension Reduction"
subtitle: "GGT-10kEval-90 · longitude, latitude, elevation, slope, aspect, soil moisture, WorldClim bio01–bio19"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## 0. What this document is

This is a cleaned-up, working version of the original exploratory script, plus
fixes, additional checks, and interpretation written directly next to the
code that produced each result. Section markers below call out:

- 🛠 **Fix** — something in the original script that errored or was fragile
- 💡 **Idea** — an analysis that wasn't in the original script but is worth adding
- 📊 a visualization
- 📝 a comment/interpretation of what came out

```{r setup}
library(jsonlite)
library(dplyr)
library(purrr)
library(tibble)
library(stringr)
library(ggplot2)
library(corrplot)
library(tidyr)

theme_set(theme_minimal())
```

## 1. Loading the data

🛠 **Fix 1 — robust field extraction.** The original `unlist()`/`as.list()`
approach relies on JSON key order surviving several transformations, and
silently drops fields that are `null` (e.g. `aspect` on flat terrain, or
soil moisture over open water), which can misalign columns row to row. Below
each of the 27 auxiliary fields is pulled out **by name**, and missing values
are kept as explicit `NA` rather than dropped.

```{r load-data}
#| eval: false

root <- "/home/nremezova/Dokumente/data_visualization/data-viz-final-proj/GGT-10kEval-90"   # set this to your local extracted folder

aux_files <- list.files(
  root,
  pattern = "\\.auxiliary\\.json$",
  recursive = TRUE,
  full.names = TRUE
)

aux_vars <- c(
  "longitude", "latitude", "elevation", "slope", "aspect",
  "soil_moisture_0_5cm", "soil_moisture_5_15cm", "soil_moisture_15_30cm",
  paste0("bio", sprintf("%02d", 1:19))
)

read_one <- function(aux_path) {
  key <- basename(aux_path) |> str_remove("\\.auxiliary\\.json$")
  text_path <- file.path(dirname(aux_path), paste0(key, ".text.json"))

  aux   <- fromJSON(aux_path)
  label <- fromJSON(text_path)

  aux_row <- aux[aux_vars]
  aux_row[sapply(aux_row, is.null)] <- NA

  tibble(
    key     = key,
    level0  = label$level0,
    family  = label$level1_family,
    genus   = label$level2_genus,
    species = label$level3_species
  ) |>
    bind_cols(as_tibble(aux_row))
}

df <- map_dfr(aux_files, read_one)
```

```{r simulate-data}
#| echo: false
# Synthetic stand-in so this document renders end-to-end without the raw
# WebDataset tars on disk. Replace this chunk with the real loading chunk
# above when knitting against the actual data.
set.seed(42)
n <- 3000
families <- c("Rosaceae","Fabaceae","Pinaceae","Fagaceae","Betulaceae",
              "Salicaceae","Proteaceae","Araliaceae","Ulmaceae","Viburnaceae")
level0s <- c("Deciduous Broadleaf","Evergreen Broadleaf","Evergreen Needleleaf")

lat <- runif(n, -55, 70)
lon <- runif(n, -170, 170)
elev <- pmax(0, rnorm(n, 600, 500) + abs(lat) * 5)
temp_base <- 28 - 0.35 * abs(lat) - 0.006 * elev
bio01 <- (temp_base + rnorm(n, 0, 1.5)) * 10
bio06 <- bio01 - runif(n, 60, 150)
bio05 <- bio01 + runif(n, 60, 150)
bio10 <- bio01 + runif(n, 20, 60)
bio11 <- bio01 - runif(n, 20, 60)
bio04 <- (bio05 - bio06) * runif(n, 8, 14)
bio02 <- runif(n, 50, 150)
bio03 <- pmin(100, 100 * bio02 / (bio05 - bio06 + 1))
bio07 <- bio05 - bio06
precip_base <- pmax(0, 1800 - 18 * abs(lat) - 0.3 * elev + rnorm(n, 0, 300))
bio12 <- precip_base
bio13 <- bio12 * runif(n, 0.12, 0.22)
bio14 <- bio12 * runif(n, 0.01, 0.05)
bio16 <- bio13 * runif(n, 2.2, 2.8)
bio17 <- bio14 * runif(n, 2.2, 2.8)
bio15 <- runif(n, 10, 100)
bio08 <- bio10 + rnorm(n, 0, 10); bio09 <- bio11 + rnorm(n, 0, 10)
bio18 <- bio16 * runif(n, 0.7, 1.1); bio19 <- bio17 * runif(n, 0.7, 1.1)

df <- tibble(
  key = sprintf("%06d", seq_len(n)),
  level0 = sample(level0s, n, replace = TRUE,
                   prob = c(0.4, 0.35, 0.25)),
  family = sample(families, n, replace = TRUE),
  genus = paste0(substr(family, 1, 4), sample(1:8, n, replace = TRUE)),
  species = paste(genus, sample(letters[1:12], n, replace = TRUE)),
  longitude = lon, latitude = lat,
  elevation = elev,
  slope = pmax(0, rnorm(n, 8, 6)),
  aspect = sample(c(0:359, NA), n, replace = TRUE, prob = c(rep(1, 360), 15)),
  soil_moisture_0_5cm = pmax(0, rnorm(n, 25, 8)),
  soil_moisture_5_15cm = pmax(0, rnorm(n, 27, 8)),
  soil_moisture_15_30cm = pmax(0, rnorm(n, 29, 8)),
  bio01 = bio01, bio02 = bio02, bio03 = bio03, bio04 = bio04, bio05 = bio05,
  bio06 = bio06, bio07 = bio07, bio08 = bio08, bio09 = bio09, bio10 = bio10,
  bio11 = bio11, bio12 = bio12, bio13 = bio13, bio14 = bio14, bio15 = bio15,
  bio16 = bio16, bio17 = bio17, bio18 = bio18, bio19 = bio19
)
```

## 2. Missingness — check this *before* `na.omit()`

💡 **Idea.** The original script only inspects missingness after several
transformations. Doing it first tells you whether `na.omit()` is safe to use
later, or whether it would silently throw away a large, possibly
non-random, chunk of the data (e.g. coastal/water points missing `aspect`).

```{r missingness}
miss <- colSums(is.na(df)) |> sort(decreasing = TRUE)
miss[miss > 0]

complete_frac <- mean(complete.cases(df))
complete_frac
```

📝 **Comment.** If `complete_frac` is well below ~0.95, don't reach for
`na.omit()` reflexively — impute (e.g. median by `family`) or drop only the
sparsest columns instead of the sparsest *rows*, since a handful of
high-missingness columns (often `aspect`, sometimes `slope` on water/flat
terrain) can otherwise cost you a large share of usable rows.

## 3. Rename bioclim codes to readable names

```{r rename}
df <- df |>
  rename(
    annual_mean_temp          = bio01,
    mean_diurnal_temp_range   = bio02,
    isothermality              = bio03,
    temp_seasonality           = bio04,
    max_temp_warmest_month     = bio05,
    min_temp_coldest_month     = bio06,
    temp_annual_range          = bio07,
    mean_temp_wettest_quarter  = bio08,
    mean_temp_driest_quarter   = bio09,
    mean_temp_warmest_quarter  = bio10,
    mean_temp_coldest_quarter  = bio11,
    annual_precipitation       = bio12,
    precip_wettest_month       = bio13,
    precip_driest_month        = bio14,
    precip_seasonality         = bio15,
    precip_wettest_quarter     = bio16,
    precip_driest_quarter      = bio17,
    precip_warmest_quarter     = bio18,
    precip_coldest_quarter     = bio19
  )
```

📝 Note on units: WorldClim stores temperature variables in **tenths of a
degree** and seasonality (`temp_seasonality`) ×100 of the standard
deviation. This doesn't affect correlation/PCA (both are scale-invariant
once standardized) but matters if you report raw values — divide
temperature variables by 10 to get °C.

## 4. Where are the samples? 📊

```{r map}
ggplot(df, aes(longitude, latitude, color = level0)) +
  geom_point(alpha = 0.4, size = 0.6) +
  labs(title = "Sample locations by broad tree type",
       color = "Tree type", x = "Longitude", y = "Latitude")
```

📝 Sample density and coverage are rarely uniform across continents in
citation-aggregated occurrence data (e.g. GBIF-derived sources tend to be
denser in North America/Europe). Worth checking before drawing any
geography-linked conclusions — an apparent climate–family association could
just be a sampling-density artifact.

## 5. Distributions of key climate variables 📊

```{r distributions}
df |>
  select(annual_mean_temp, annual_precipitation, temp_seasonality, elevation) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of core environmental variables")
```

```{r boxplots}
ggplot(df, aes(level0, annual_mean_temp / 10, fill = level0)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Annual mean temperature by tree type", y = "°C", x = NULL)

ggplot(df, aes(level0, annual_precipitation, fill = level0)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Annual precipitation by tree type", y = "mm/yr", x = NULL)
```

📝 **Interpretation.** Evergreen needleleaf species should skew toward
colder/drier climates, evergreen broadleaf toward warmer/wetter, and
deciduous broadleaf toward seasonal mid-latitude climates with a strong
`temp_seasonality` signal. If your real data doesn't show this separation
at all, that's a useful diagnostic — it suggests either heavy label noise
or that `level0` isn't capturing climate-relevant functional variation as
cleanly as expected.

## 6. Correlation structure 🛠 📊

🛠 **Fix 2.** The original had a stray, invalid pipe (`|))>`) that crashed
the script, plus a redundant first version of `cor_long` (kept both
`Var1≠Var2` and the deduplicated `Var1<Var2` version — only one is needed),
plus a `select(-sample_id)` that fails because the ID column is actually
called `key`. All three are fixed below; `key`/`image_path` are
automatically excluded since `select(where(is.numeric))` already drops
character columns.

```{r correlation}
aux_corr <- df |> select(where(is.numeric)) |> na.omit()
cor_mat <- cor(aux_corr, method = "pearson")

corrplot(cor_mat, method = "color", type = "upper",
         tl.cex = 0.65, tl.col = "black")

cor_long <- as.data.frame(as.table(cor_mat)) |>
  filter(as.character(Var1) < as.character(Var2)) |>
  mutate(abs_corr = abs(Freq)) |>
  arrange(desc(abs_corr))

high_corr <- cor_long |> filter(abs_corr > 0.9)
high_corr
```

📝 **Interpretation.** Expect (and this is structural, not a data problem)
the temperature-family variables — `annual_mean_temp`,
`min_temp_coldest_month`, `mean_temp_coldest_quarter`,
`mean_temp_warmest_quarter` — to correlate above r = 0.9 with each other,
since they're all derived from the same monthly temperature curve. Same
story for the precipitation-quarter variables versus
`annual_precipitation`. This is the standard, well-known collinearity
problem with using all 19 WorldClim bioclim variables together, and it's
the main reason to reduce dimensionality before modeling (Section 8) rather
than feeding all 19 raw variables into, say, a regression.

## 7. 💡 Idea: a decorrelated variable subset (VIF-based)

Rather than relying on PCA alone to absorb collinearity, it's worth
explicitly pruning to a smaller, less redundant variable set for any
downstream interpretable modeling (e.g. "which climate variables predict
`level0`?").

```{r vif}
#| eval: false
# install.packages("car")
library(car)

full_model <- lm(annual_mean_temp ~ ., data = aux_corr |> select(-longitude, -latitude))
# iteratively drop the variable with the highest VIF until all VIF < 5
vif_prune <- function(data, threshold = 5) {
  repeat {
    v <- car::vif(lm(reformulate(setdiff(names(data), "target"), "target"),
                      data = data |> mutate(target = data[[1]])))
    if (max(v) < threshold) break
    data <- data |> select(-all_of(names(which.max(v))))
  }
  data
}
```

📝 A simpler, commonly used WorldClim subset to start from:
`bio01, bio02, bio03 (or bio04), bio12, bio15, bio18, bio19` — temperature
mean/range/seasonality plus precipitation mean/seasonality/extremes,
deliberately picked to span the two main axes (temperature, precipitation)
without redundant quarter-level variables.

## 8. PCA 🛠 📊

🛠 **Fix 3.** The original scaled the matrix with `scale()` *and* passed
`scale.=TRUE` to `prcomp()` — redundant (scaling an already-standardized
matrix is a no-op), but confusing. Below, scaling happens once, inside
`prcomp()`.

```{r pca}
pca_data <- df |>
  select(longitude, latitude, elevation, slope, aspect,
         starts_with("soil_moisture"),
         annual_mean_temp, mean_diurnal_temp_range, isothermality,
         temp_seasonality, max_temp_warmest_month, min_temp_coldest_month,
         temp_annual_range, mean_temp_wettest_quarter, mean_temp_driest_quarter,
         mean_temp_warmest_quarter, mean_temp_coldest_quarter,
         annual_precipitation, precip_wettest_month, precip_driest_month,
         precip_seasonality, precip_wettest_quarter, precip_driest_quarter,
         precip_warmest_quarter, precip_coldest_quarter) |>
  na.omit()

pca <- prcomp(pca_data, center = TRUE, scale. = TRUE)
summary(pca)$importance[, 1:5]
```

```{r pca-loadings}
as.data.frame(pca$rotation[, 1:2]) |>
  mutate(variable = rownames(pca$rotation)) |>
  arrange(desc(abs(PC1))) |>
  head(10)
```

```{r pca-biplot}
pca_scores <- as.data.frame(pca$x) |>
  bind_cols(df |> select(level0, family, genus, species) |>
              slice(as.numeric(rownames(pca_data))))

loadings <- as.data.frame(pca$rotation[, 1:2])
loadings$variable <- rownames(loadings)
arrow_scale <- 4

ggplot(pca_scores, aes(PC1, PC2, color = level0)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               inherit.aes = FALSE, arrow = arrow(length = unit(0.15, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1 * arrow_scale * 1.15, y = PC2 * arrow_scale * 1.15,
                label = variable),
            inherit.aes = FALSE, size = 2.8) +
  labs(title = "PCA biplot of auxiliary environmental variables",
       color = "Tree type")
```

📝 **Interpretation.** Given the correlation structure from Section 6,
expect:

- **PC1** to load heavily on the temperature cluster (often with elevation
  loading oppositely, via the lapse rate) — a general
  "warm/low-elevation ↔ cold/high-elevation" axis.
- **PC2** to load on the precipitation cluster — a "wet ↔ dry" axis,
  largely orthogonal to PC1.
- Together these two axes typically explain a large share of total
  variance for bioclim data (often 50–70%+) *because* of the redundancy
  identified above — PCA is doing real work here, not just noise reduction.
- `level0` (deciduous/evergreen/needleleaf) is the label most likely to
  separate visibly along PC1, since it's strongly linked to seasonality and
  cold tolerance. `family`/`genus`/`species` will separate far less
  cleanly — environment doesn't fully determine taxonomy (see Section 10).

## 9. t-SNE / UMAP for visual clustering 🛠 💡 📊

🛠 **Fix 4.** Kept `pca = TRUE` inside `Rtsne()` (pre-reduces to a
manageable number of PCs before the heavy t-SNE step, which is good
practice at this scale), but added a UMAP alternative since it's typically
faster and better preserves global structure, which matters when comparing
cluster separation against `family`/`level0`.

```{r tsne-umap}
#| eval: false
library(Rtsne)
# install.packages("uwot")
library(uwot)

tsne_data <- pca_data |> scale()

set.seed(123)
tsne <- Rtsne(tsne_data, dims = 2, perplexity = 30, theta = 0.5,
              pca = TRUE, check_duplicates = FALSE)

set.seed(123)
umap_coords <- umap(tsne_data, n_neighbors = 30, min_dist = 0.1)

tsne_plot <- as.data.frame(tsne$Y) |> setNames(c("TSNE1", "TSNE2")) |>
  bind_cols(as.data.frame(umap_coords) |> setNames(c("UMAP1", "UMAP2"))) |>
  bind_cols(df |> select(level0, family, genus, species) |>
              slice(as.numeric(rownames(pca_data))))

ggplot(tsne_plot, aes(TSNE1, TSNE2, color = level0)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "t-SNE of auxiliary environmental variables", color = "Tree type")

ggplot(tsne_plot, aes(UMAP1, UMAP2, color = level0)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "UMAP of auxiliary environmental variables", color = "Tree type")
```

📝 **Comment.** Rerun t-SNE with at least 2–3 perplexity values
(e.g. 10, 30, 50) before trusting any single layout — t-SNE cluster shapes
and even apparent cluster *count* can shift noticeably with this parameter,
and inter-cluster distances in t-SNE are not meaningful (only local
neighborhood structure is). UMAP is generally more robust on this front and
worth treating as the primary visualization, with t-SNE as a
cross-check.

## 10. 💡 Ideas for further analysis

1. **Climate niche summary by family/species** — instead of (or alongside)
   unsupervised clustering, compute per-family or per-species means/SDs of
   the core climate variables (`annual_mean_temp`, `annual_precipitation`,
   `temp_seasonality`) and rank families by niche breadth. This turns the
   PCA/t-SNE intuition into a concrete, citable table.
2. **Spatial cross-validation** — if any predictive model is built on these
   variables (e.g. predicting `family` from climate), use spatially-blocked
   CV rather than random CV. Nearby points share almost identical climate
   values by construction, so random CV will overestimate accuracy via
   spatial autocorrelation leakage.
3. **Sampling-density correction** — pair the auxiliary data with the
   `GlobalGeoTree.csv` metadata (source/year/location description) to check
   whether climate-variable distributions are confounded with
   data-source/region, before treating PCA/t-SNE separation as purely
   biological signal.
4. **Compare 90 / 300 / 900-species eval splits** — since GGT-10kEval ships
   in Rare/Common/Frequent-balanced versions at three sizes, rerunning the
   same PCA/correlation pipeline across splits is a quick way to check
   whether conclusions are stable or are an artifact of which 90 species
   got curated into the smallest eval set.
5. **Joint analysis with the Sentinel-2 time series** — the auxiliary
   variables are static (one value per location); pairing PC1/PC2 scores
   against derived image features (e.g. NDVI seasonality from the 12
   monthly composites) could show how much of the *image* signal is
   actually attributable to climate rather than tree identity per se —
   relevant context for interpreting what GeoTreeCLIP is "really" learning
   from auxiliary vs. image inputs.

## 11. Summary

- The original script's hard crash was a single corrupted pipe character;
  a second, separate bug (`select(-sample_id)`) would have surfaced right
  after fixing the first.
- Dimension reduction is well-justified here, not just permissible — the
  19 bioclim variables are highly collinear by construction, so PCA/UMAP
  genuinely compress redundant information rather than discarding signal.
- The clearest expected signal is climate ↔ `level0` (broad functional
  type), via temperature seasonality and elevation. Climate ↔ fine-grained
  taxonomy (`genus`/`species`) is expected to be much weaker, since
  environment constrains but doesn't determine species identity.
