---
title: "GlobalGeoTree: Source Bias and Temporal Development"
subtitle: "RQ2 - Source and Platform Bias, and RQ3 - Temporal Development, combined into one report"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(data.table)   # fast reading of the large CSV
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "gray35"),
      plot.caption = element_text(color = "gray45", size = 9),
      legend.position = "bottom"
    )
)

# one color per source group, reused in every chart in this report so the
# same source always has the same color everywhere
source_colors <- c(
  "iNaturalist"                = "#8ECAE6",
  "Pl@ntNet"                   = "#FF9DA7",
  "GBIF"                       = "#264653",
  "Wikidata/Wikipedia"         = "#B07AA1",
  "Institutional collections"  = "#9C6644",
  "Inventory/survey"           = "#B2DF8A",
  "Other source"               = "#2CA02C",
  "Unknown source"             = "#E41A1C"
)
source_group_levels <- names(source_colors)
```

# Research Questions

This report combines two related research questions, because they share the same data and the same key idea: **GlobalGeoTree is not a neutral, evenly-collected dataset** - it is built from thousands of different sources (mostly citizen-science apps), and that has consequences for how the geographic and temporal patterns in it should be read.

**RQ2 - Source and Platform Bias.** To what extent do data sources shape the observed geographic and taxonomic patterns in GlobalGeoTree?

1. Which sources contribute the largest share of observations?
2. Is the dataset dominated by citizen-science platforms such as iNaturalist?
3. Do different sources cover different countries or regions?
4. Are some sources associated with specific years, countries, or species groups?

**RQ3 - Temporal Development.** How has the number and distribution of tree observations changed from 2015 to 2024?

1. Has the number of observations increased over time?
2. Which years show the strongest growth or decline?
3. Did geographic coverage expand over time?
4. Are temporal patterns mainly driven by specific sources?

Both analyses are descriptive. Strong source-specific patterns should be read as evidence of recording and platform bias, not only as biological patterns in tree distributions.

# Data Loading

```{r config}
start_year <- 2015
end_year <- 2024

top_n_sources    <- 10   # how many individual sources to show by name
top_n_countries  <- 15
top_n_taxa       <- 12
top_n_source_maps <- 4   # how many source groups get their own map panel

snapshot_years <- c(2015, 2018, 2021, 2024)   # years shown in the RQ3 map grid
max_map_points <- 200000                       # cap on points per map, for readability
```

```{r column-mapping}
# the real column names in GlobalGeoTree.csv, checked once with names(raw).
# this is the only place column names should need to be edited.
col_source      <- "source"
col_country     <- "country_code"
col_species     <- "level3_species"
col_family      <- "level1_family"
col_genus       <- "level2_genus"
col_year        <- "year"
col_lat         <- "latitude"
col_lon         <- "longitude"
```

```{r load-and-clean}
raw <- fread("GlobalGeoTree.csv", showProgress = FALSE)

# -----------------------------------------------------------------------
# how source names are grouped into broad categories:
#   - iNaturalist, Pl@ntNet, GBIF, Wikidata/Wikipedia, Institutional
#     collections, and Inventory/survey are matched by keywords in the
#     source name (e.g. "herbarium" -> Institutional collections)
#   - "Unknown source" = the source field was empty or explicitly said
#     unknown/missing/n.a.
#   - "Other source" = everything left over: thousands of individually
#     named sources (national biodiversity portals, government forestry
#     databases, regional citizen-science projects, museum exports, etc.)
#     that don't match any of the patterns above. It is a catch-all
#     bucket, not a single source - see the "What is in Other source?"
#     section below for what's actually inside it.
# -----------------------------------------------------------------------
classify_source <- function(x) {
  x_lower <- str_to_lower(str_trim(x))
  case_when(
    str_detect(x_lower, "inaturalist|inat")                     ~ "iNaturalist",
    str_detect(x_lower, "plantnet|pl@ntnet")                    ~ "Pl@ntNet",
    str_detect(x_lower, "gbif")                                  ~ "GBIF",
    str_detect(x_lower, "wikidata|wikipedia")                    ~ "Wikidata/Wikipedia",
    str_detect(x_lower, "herbar|museum|botanic|arboretum")       ~ "Institutional collections",
    str_detect(x_lower, "inventory|census|survey|monitoring")    ~ "Inventory/survey",
    str_detect(x_lower, "^(unknown( source)?|missing|n/?a|na|none|null)$") ~ "Unknown source",
    TRUE                                                         ~ "Other source"
  )
}

df <- raw %>%
  mutate(
    source_raw   = ifelse(is.na(.data[[col_source]]) | str_trim(.data[[col_source]]) == "",
                           "Unknown source", str_trim(.data[[col_source]])),
    source_group = factor(classify_source(source_raw), levels = source_group_levels),

    country_value     = str_to_upper(str_trim(.data[[col_country]])),
    species_value     = str_trim(.data[[col_species]]),
    family_value      = str_trim(.data[[col_family]]),
    genus_value       = str_trim(.data[[col_genus]]),
    year_value        = as.integer(.data[[col_year]]),
    latitude_value    = suppressWarnings(as.numeric(.data[[col_lat]])),
    longitude_value   = suppressWarnings(as.numeric(.data[[col_lon]]))
  ) %>%
  filter(
    !is.na(country_value), country_value != "",
    !is.na(species_value), species_value != ""
  )

has_coordinates <- TRUE  # latitude/longitude are required columns in this file
```

```{r data-summary}
data_summary <- tibble(
  Measure = c(
    "Observation file", "Period analysed", "Rows analysed",
    "Distinct sources", "Distinct source groups", "Distinct countries",
    "Distinct species"
  ),
  Value = c(
    "GlobalGeoTree.csv",
    paste(start_year, end_year, sep = "-"),
    comma(nrow(df)),
    comma(n_distinct(df$source_raw)),
    comma(n_distinct(df$source_group)),
    comma(n_distinct(df$country_value)),
    comma(n_distinct(df$species_value))
  )
)
knitr::kable(data_summary, align = "ll")
```

This report uses `r comma(nrow(df))` observations.

# Source Concentration

**Why this visualisation is needed.** Before comparing countries, years, or species, we first need to know whether the dataset is balanced across sources. If one source is extremely large, it can shape the whole dataset.

```{r source-share-bar, fig.width=9.5, fig.height=5.8}
# Horizontal bars are used here because source names are long.
# The percentage labels show each source's share of all observations.
top_sources <- df %>%
  count(source_raw, sort = TRUE, name = "observations") %>%
  mutate(share = observations / sum(observations)) %>%
  slice_head(n = top_n_sources) %>%
  mutate(source_short = reorder(str_trunc(source_raw, 45), observations))

ggplot(top_sources, aes(x = source_short, y = observations)) +
  geom_col(fill = "#2A7F62", width = 0.72) +
  geom_text(
    aes(label = percent(share, accuracy = 0.1)),
    hjust = -0.08,
    size = 3.2
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Largest observation sources in GlobalGeoTree",
    subtitle = "The top source is large enough to shape the overall dataset.",
    x = NULL,
    y = "Observations",
    caption = "Long source names are shortened for readability."
  )
```

# What Is in "Other Source"?

*Why we look at this:* "Other source" is the second-largest group in this dataset, but unlike iNaturalist or Pl@ntNet it is not one platform - it is everything that didn't match a recognised keyword pattern. Before using it in any chart, it's worth showing what's actually inside it.

```{r other-source-data}
other_source_top <- df %>%
  filter(source_group == "Other source") %>%
  count(source_raw, sort = TRUE, name = "observations") %>%
  mutate(share_of_other = observations / sum(observations)) %>%
  slice_head(n = 15)

n_other_sources <- df %>% filter(source_group == "Other source") %>% summarise(n = n_distinct(source_raw)) %>% pull(n)
```

```{r other-source-plot, fig.width=9, fig.height=5.5}
ggplot(other_source_top, aes(x = reorder(str_trunc(source_raw, 45), observations), y = observations)) +
  geom_col(fill = "#2A7F62", width = 0.72) +
  geom_text(aes(label = percent(share_of_other, accuracy = 0.1)), hjust = -0.08, size = 3.1) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.16))) +
  labs(
    title = "The 15 largest individual sources inside \"Other source\"",
    subtitle = paste0("\"Other source\" is made up of ", comma(n_other_sources), " differently-named sources in total."),
    x = NULL, y = "Observations",
    caption = "Percentages are share of the Other source group, not of the whole dataset."
  )
```

**Interpretation.** "Other source" is not a single platform - it is a long tail of `r comma(n_other_sources)` individually named sources (national biodiversity portals, regional/government forestry or environmental-impact datasets, local citizen-science platforms such as Artportalen or Observation.org, and similar projects). Treat any chart that shows "Other source" as a group as a *mixture*, not as one consistent collection method - the countries and years where "Other source" is large are typically being driven by a handful of large national datasets rather than a single shared bias.

```{r other-source-name-audit}
other_source_name_audit <- df %>%
  filter(source_group == "Other source") %>%
  count(source_raw, sort = TRUE, name = "observations") %>%
  mutate(share_of_other = observations / sum(observations)) %>%
  slice_head(n = 50)

knitr::kable(other_source_name_audit, align = "lrr")
```

# Source Concentration

*Why we look at this:* before comparing countries, years, or species, we first need to know whether the dataset is balanced across sources. If one source is extremely large, it shapes the whole dataset (RQ2 subquestions 1-2).

```{r source-concentration-data}
source_counts <- df %>%
  count(source_raw, sort = TRUE, name = "observations") %>%
  mutate(
    share = observations / sum(observations),
    cumulative_share = cumsum(share),
    source_rank = row_number()
  )

source_group_counts <- df %>%
  count(source_group, name = "observations") %>%
  mutate(share = observations / sum(observations))

inat_share    <- source_group_counts$share[source_group_counts$source_group == "iNaturalist"]
citizen_share <- sum(source_group_counts$share[source_group_counts$source_group %in% c("iNaturalist", "Pl@ntNet")])
```

The source structure is highly concentrated: **iNaturalist alone contributes `r percent(inat_share, accuracy = 0.1)`** of observations, and **iNaturalist plus Pl@ntNet together contribute `r percent(citizen_share, accuracy = 0.1)`**.

```{r source-pareto, fig.width=9, fig.height=5}
# Pareto chart: bars are each source's individual share, the line is the
# running total. A line that reaches close to 100% after only a few bars
# means a small number of sources explain almost everything.
ggplot(source_counts %>% slice_head(n = 50), aes(x = source_rank)) +
  geom_col(aes(y = share), fill = "#2A7F62", alpha = 0.75, width = 0.85) +
  geom_line(aes(y = cumulative_share), color = "#C23B22", linewidth = 1) +
  scale_x_continuous(breaks = c(1, 5, 10, 25, 50)) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(
    title = "A small number of sources explain most observations",
    subtitle = "Bars show each source's individual share; the red line shows cumulative share.",
    x = "Source rank", y = "Share of all observations"
  )
```

```{r source-group-bar, fig.width=8.5, fig.height=4.5}
ggplot(source_group_counts, aes(x = reorder(source_group, observations), y = observations, fill = source_group)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_text(aes(label = percent(share, accuracy = 0.1)), hjust = -0.08, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = source_colors) +
  labs(
    title = "Observation share by broad source group",
    subtitle = "Citizen-science platforms dominate the grouped source structure.",
    x = NULL, y = "Observations"
  )
```

```{r citizen-science-bar, fig.width=7, fig.height=4}
citizen_summary <- df %>%
  mutate(is_citizen = source_group %in% c("iNaturalist", "Pl@ntNet")) %>%
  count(is_citizen, name = "observations") %>%
  mutate(
    share = observations / sum(observations),
    label = ifelse(is_citizen, "Citizen-science source", "Other or unclear source")
  )

ggplot(citizen_summary, aes(x = label, y = observations, fill = is_citizen)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = percent(share, accuracy = 0.1)), vjust = -0.4, size = 4) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.14))) +
  scale_fill_manual(values = c("TRUE" = "#8ECAE6", "FALSE" = "#9C6644")) +
  labs(
    title = "Citizen-science platforms dominate the observations",
    subtitle = "Citizen-science is defined here as iNaturalist plus Pl@ntNet.",
    x = NULL, y = "Observations"
  )
```

**Interpretation.** A few large platforms dominate the evidence base, so global summaries are strongly affected by the behaviour, geography, and taxonomy covered by those platforms specifically.

# Temporal Development

*Why we look at this:* before anything more detailed, we need the basic shape of the data over time - is it going up, down, or flat (RQ3 subquestions 1-2) - and whether that shape is really about trees, or about one platform growing (RQ2 subquestion 4 / RQ3 subquestion 4).

```{r yearly-counts}
yearly_counts <- df %>%
  filter(year_value >= start_year, year_value <= end_year) %>%
  count(year_value, name = "observations") %>%
  right_join(tibble(year_value = start_year:end_year), by = "year_value") %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations)) %>%
  arrange(year_value) %>%
  mutate(absolute_change = observations - lag(observations))

growth_summary <- yearly_counts %>% filter(!is.na(absolute_change))
strongest_growth  <- growth_summary %>% slice_max(absolute_change, n = 1, with_ties = FALSE)
strongest_decline <- growth_summary %>% slice_min(absolute_change, n = 1, with_ties = FALSE)

first_year_count <- yearly_counts$observations[yearly_counts$year_value == start_year]
last_year_count  <- yearly_counts$observations[yearly_counts$year_value == end_year]
overall_change   <- (last_year_count - first_year_count) / first_year_count
```

Between `r start_year` and `r end_year`, observations changed from **`r comma(first_year_count)`** to **`r comma(last_year_count)`**, an overall change of **`r percent(overall_change, accuracy = 0.1)`**.

```{r line-observations-by-year, fig.width=9, fig.height=4.6}
ggplot(yearly_counts, aes(x = year_value, y = observations)) +
  geom_line(color = "#2A7F62", linewidth = 1.1) +
  geom_point(color = "#2A7F62", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(title = "Number of tree observations by year", x = "Year", y = "Observations")
```

```{r yearly-growth-plot, fig.width=9, fig.height=4.6}
ggplot(growth_summary, aes(x = year_value, y = absolute_change, fill = absolute_change >= 0)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "gray35", linewidth = 0.4) +
  geom_text(aes(label = comma(absolute_change)),
            vjust = ifelse(growth_summary$absolute_change >= 0, -0.35, 1.15), size = 3.1) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.12, 0.16))) +
  scale_fill_manual(values = c("TRUE" = "#2A7F62", "FALSE" = "#C23B22")) +
  labs(
    title = "Year-to-year change in observations", x = "Year", y = "Change in observations",
    caption = paste0("Note: ", end_year, " is the last year in the export and may not be complete yet, so part of\n",
                      "the apparent decline could be a data-cutoff effect rather than a real drop in recording.")
  )
```

**Interpretation.** The strongest growth occurs in **`r strongest_growth$year_value`** (+`r comma(strongest_growth$absolute_change)`), and the strongest decline in **`r strongest_decline$year_value`** (`r comma(strongest_decline$absolute_change)`). The `r end_year` figure should be treated with caution: if that year's data was exported before the year finished, the apparent decline may shrink once a complete year is available.

```{r source-year-data}
top_source_groups <- source_group_counts %>% arrange(desc(observations)) %>%
  slice_head(n = top_n_sources) %>% pull(source_group) %>% as.character()

# shared by both research questions: this is "is the dataset growing
# because of trees, or because of one platform" (RQ3 q4) and "are sources
# tied to specific years" (RQ2 q4) at the same time - one table answers both
source_year <- df %>%
  filter(year_value >= start_year, year_value <= end_year) %>%
  mutate(source_group_plot = ifelse(as.character(source_group) %in% top_source_groups,
                                     as.character(source_group), "Other source")) %>%
  count(year_value, source_group_plot, name = "observations") %>%
  right_join(expand.grid(year_value = start_year:end_year, source_group_plot = source_group_levels),
             by = c("year_value", "source_group_plot")) %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations)) %>%
  group_by(year_value) %>%
  mutate(share = observations / sum(observations)) %>%
  ungroup() %>%
  mutate(source_group_plot = factor(source_group_plot, levels = source_group_levels))
```

```{r stacked-area-source-year, fig.width=9, fig.height=5}
ggplot(source_year, aes(x = year_value, y = observations, fill = source_group_plot)) +
  geom_area(alpha = 0.9, linewidth = 0.15, color = "white") +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(title = "Observation volume by source group and year",
       subtitle = "A rising source group can drive the apparent growth of the full dataset.",
       x = "Year", y = "Observations", fill = "Source group")
```

```{r source-share-year, fig.width=9, fig.height=5}
ggplot(source_year, aes(x = year_value, y = share, fill = source_group_plot)) +
  geom_col(width = 0.9) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(title = "Source composition by year", subtitle = "Bars sum to 100% within each year.",
       x = "Year", y = "Share of observations", fill = "Source group",
       caption = "The most recent year may be incomplete depending on the data export date.")
```

```{r source-growth-plot, fig.width=9, fig.height=4.6}
largest_source_growth <- source_year %>%
  arrange(source_group_plot, year_value) %>%
  group_by(source_group_plot) %>%
  mutate(source_change = observations - lag(observations)) %>%
  ungroup() %>%
  filter(!is.na(source_change)) %>%
  slice_max(source_change, n = 8, with_ties = FALSE) %>%
  mutate(label = paste0(year_value, ": ", source_group_plot))

ggplot(largest_source_growth, aes(x = reorder(label, source_change), y = source_change, fill = source_group_plot)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(title = "Largest source-specific yearly increases", x = NULL, y = "Increase from previous year")
```

**Interpretation.** If most growth is concentrated in one source group, the temporal trend is source-driven, not a neutral increase in global tree recording. This ranks *absolute* increases, so later years (when the whole dataset is bigger) have a built-in advantage - a percentage-growth version would be a useful follow-up.

# Geographic Coverage Over Time

*Why we look at this:* more observations is not the same as broader observations - growth could just mean recording the same countries more intensively (RQ3 subquestion 3).

```{r country-coverage-data}
country_year <- df %>%
  filter(year_value >= start_year, year_value <= end_year) %>%
  count(year_value, country_value, name = "observations")

countries_per_year <- country_year %>%
  group_by(year_value) %>% summarise(countries_observed = n_distinct(country_value), .groups = "drop")

country_first_year <- country_year %>%
  group_by(country_value) %>% summarise(first_observed_year = min(year_value), .groups = "drop")

cumulative_countries <- tibble(year_value = start_year:end_year) %>%
  rowwise() %>%
  mutate(cumulative_countries = sum(country_first_year$first_observed_year <= year_value)) %>%
  ungroup()
```

```{r countries-per-year, fig.width=9, fig.height=4.6}
# axis starts at 0 so normal year-to-year wiggle doesn't look exaggerated
ggplot(countries_per_year, aes(x = year_value, y = countries_observed)) +
  geom_line(color = "#4E79A7", linewidth = 1.1) +
  geom_point(color = "#4E79A7", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Countries represented each year", x = "Year", y = "Countries with observations")
```

```{r cumulative-country-coverage, fig.width=9, fig.height=4.6}
ggplot(cumulative_countries, aes(x = year_value, y = cumulative_countries)) +
  geom_area(fill = "#8ECAE6", alpha = 0.55) +
  geom_line(color = "#1F5B74", linewidth = 1.1) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(title = "Cumulative country coverage over time", x = "Year", y = "Cumulative countries")
```

```{r new-countries-plot, fig.width=9, fig.height=4.4}
# start_year is excluded: every country present in the very first year of
# the window counts as "new" purely because that's where the window starts,
# not because it's genuinely a new entry - so only 2016+ is shown here
new_countries <- country_first_year %>%
  filter(first_observed_year > start_year) %>%
  count(first_observed_year, name = "new_countries") %>%
  right_join(tibble(first_observed_year = (start_year + 1):end_year), by = "first_observed_year") %>%
  mutate(new_countries = ifelse(is.na(new_countries), 0L, new_countries))

countries_at_start <- sum(country_first_year$first_observed_year == start_year)

ggplot(new_countries, aes(x = first_observed_year, y = new_countries)) +
  geom_col(fill = "#4E79A7", width = 0.72) +
  scale_x_continuous(breaks = (start_year + 1):end_year) +
  scale_y_continuous(labels = comma) +
  labs(title = "New countries entering the dataset by year", x = "First observed year in period", y = "New countries",
       caption = paste0(start_year, " is excluded: ", countries_at_start,
                         " countries already had observations in the first year of the window."))
```

**Interpretation.** If country counts increase over time, GlobalGeoTree is expanding geographically; if observations grow without much country expansion, growth is concentrated in already well-covered places.

# Countries and Regional Coverage

*Why we look at this:* a global source pattern can hide strong differences between countries - is each country's evidence base built the same way (RQ2 subquestion 3)?

```{r country-source-data}
top_country_values <- df %>% count(country_value, sort = TRUE) %>% slice_head(n = top_n_countries) %>% pull(country_value)

country_source <- df %>%
  filter(country_value %in% top_country_values) %>%
  mutate(source_group_plot = ifelse(as.character(source_group) %in% top_source_groups,
                                     as.character(source_group), "Other source")) %>%
  count(country_value, source_group_plot, name = "observations") %>%
  group_by(country_value) %>%
  mutate(country_total = sum(observations), share = observations / country_total) %>%
  ungroup()

country_order <- country_source %>% distinct(country_value, country_total) %>% arrange(country_total) %>% pull(country_value)
country_source <- country_source %>%
  mutate(country_value = factor(country_value, levels = country_order),
         source_group_plot = factor(source_group_plot, levels = source_group_levels))

country_dominant <- df %>%
  count(country_value, source_group, name = "observations") %>%
  group_by(country_value) %>%
  mutate(country_total = sum(observations), share = observations / country_total) %>%
  slice_max(observations, n = 1, with_ties = FALSE) %>%
  ungroup()
```

```{r country-source-stacked-share, fig.width=9, fig.height=5.8}
ggplot(country_source, aes(x = country_value, y = share, fill = source_group_plot)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(title = "Source composition differs by country",
       subtitle = paste0("Top ", top_n_countries, " countries by observations; bars sum to 100% within each country."),
       x = NULL, y = "Share of country observations", fill = "Source group")
```

**Interpretation.** Countries differ both in volume and in source mix - a country dominated by iNaturalist is not directly comparable to one dominated by local inventories or unknown sources.

```{r country-diversity-data}
# Shannon diversity -> "effective number of source groups" per country.
# 1 means one source dominates; higher means a more mixed source structure.
shannon_diversity <- function(counts) {
  p <- counts / sum(counts)
  p <- p[p > 0]
  -sum(p * log(p))
}

country_diversity <- df %>%
  count(country_value, source_group, name = "observations") %>%
  group_by(country_value) %>%
  summarise(
    total_observations = sum(observations),
    effective_sources = exp(shannon_diversity(observations)),
    dominant_source = as.character(source_group[which.max(observations)]),
    .groups = "drop"
  ) %>%
  arrange(desc(total_observations)) %>%
  slice_head(n = top_n_countries) %>%
  mutate(country_value = factor(country_value, levels = rev(country_value)))
```

```{r source-diversity-country, fig.width=9, fig.height=5.4}
ggplot(country_diversity, aes(x = country_value, y = effective_sources, fill = dominant_source)) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(title = "How source-diverse are the largest country samples?",
       subtitle = "Effective number of source groups is low when one source dominates a country.",
       x = NULL, y = "Effective number of source groups", fill = "Dominant source")
```

**Interpretation.** Low diversity means a country's evidence base is essentially one platform's user base, so any pattern found there is at higher risk of being platform-specific rather than biological.

# Sources and Taxonomic Groups

*Why we look at this:* source bias can also be taxonomic - some platforms may record ornamental or easy-to-photograph families more often than others (RQ2 subquestion 4).

```{r taxonomic-source-data}
df <- df %>% mutate(family_clean = ifelse(family_value == "", "Unknown family", family_value))

top_families <- df %>% count(family_clean, sort = TRUE) %>% slice_head(n = top_n_taxa) %>% pull(family_clean)

family_source <- df %>%
  filter(family_clean %in% top_families) %>%
  mutate(source_group_plot = ifelse(as.character(source_group) %in% top_source_groups,
                                     as.character(source_group), "Other source")) %>%
  count(family_clean, source_group_plot, name = "observations") %>%
  group_by(family_clean) %>%
  mutate(share = observations / sum(observations)) %>%
  ungroup()

family_order <- family_source %>% group_by(family_clean) %>% summarise(total = sum(observations)) %>% arrange(total) %>% pull(family_clean)
family_source <- family_source %>%
  mutate(family_clean = factor(family_clean, levels = family_order),
         source_group_plot = factor(source_group_plot, levels = source_group_levels))
```

```{r taxon-source-share, fig.width=9, fig.height=5.6}
ggplot(family_source, aes(x = family_clean, y = share, fill = source_group_plot)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(title = "Source composition by family", subtitle = "Bars sum to 100% within each taxonomic group.",
       x = NULL, y = "Share of observations", fill = "Source group")
```

**Interpretation.** If a family is mostly recorded by one platform, its geographic pattern may partly reflect that platform's user base rather than the tree's real distribution.

# Geographic Footprints of Sources and Time

*Why we look at this:* spatial clusters in the raw points may reflect user activity and platform coverage, not only real tree distributions (RQ2 subquestion 3 and RQ3 subquestion 3, viewed spatially).

```{r source-map-data}
map_source_groups <- source_group_counts %>%
  filter(!source_group %in% c("Unknown source", "Other source")) %>%
  arrange(desc(observations)) %>% slice_head(n = top_n_source_maps) %>% pull(source_group) %>% as.character()

map_data_by_source <- df %>%
  filter(as.character(source_group) %in% map_source_groups,
         !is.na(latitude_value), !is.na(longitude_value),
         between(latitude_value, -90, 90), between(longitude_value, -180, 180)) %>%
  group_by(source_group) %>%
  filter(row_number() %in% sample(n(), min(n(), max_map_points))) %>%
  ungroup()

world_outline <- if (requireNamespace("maps", quietly = TRUE)) ggplot2::map_data("world") else NULL
```

```{r source-faceted-map, fig.width=9.5, fig.height=6.6}
p <- ggplot()
if (!is.null(world_outline)) {
  p <- p + geom_polygon(data = world_outline, aes(x = long, y = lat, group = group),
                         fill = "gray96", color = "gray82", linewidth = 0.12, inherit.aes = FALSE)
}
p + geom_point(data = map_data_by_source, aes(x = longitude_value, y = latitude_value),
               alpha = 0.12, size = 0.22, color = "#1F5B74") +
  facet_wrap(~ source_group, ncol = 2) +
  coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(title = "Geographic coverage of major source groups",
       subtitle = "Different platforms leave visibly different spatial footprints.",
       x = "Longitude", y = "Latitude",
       caption = "Points show recording locations, not true species ranges. Points are sampled for readability.")
```

```{r map-snapshots-data}
snapshot_years <- snapshot_years[snapshot_years >= start_year & snapshot_years <= end_year]

map_data_by_year <- df %>%
  filter(year_value %in% snapshot_years,
         !is.na(latitude_value), !is.na(longitude_value),
         between(latitude_value, -90, 90), between(longitude_value, -180, 180)) %>%
  group_by(year_value) %>%
  filter(row_number() %in% sample(n(), min(n(), max_map_points))) %>%
  ungroup()
```

```{r map-snapshots-plot, fig.width=9.5, fig.height=6.6}
p <- ggplot()
if (!is.null(world_outline)) {
  p <- p + geom_polygon(data = world_outline, aes(x = long, y = lat, group = group),
                         fill = "gray96", color = "gray82", linewidth = 0.12, inherit.aes = FALSE)
}
p + geom_point(data = map_data_by_year, aes(x = longitude_value, y = latitude_value),
               alpha = 0.15, size = 0.22, color = "#1F5B74") +
  facet_wrap(~ year_value, ncol = 2) +
  coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
  labs(title = "Map snapshots of observations over time",
       subtitle = "Selected years show whether records spread geographically or intensify in existing regions.",
       x = "Longitude", y = "Latitude",
       caption = "Points are sampled per year for readability.")
```

```{r grid-coverage-data}
# split the world into 5x5 degree boxes - a finer-grained measure of spatial
# spread than country borders, useful for catching growth inside large countries
grid_first_year <- df %>%
  filter(!is.na(latitude_value), !is.na(longitude_value),
         between(latitude_value, -90, 90), between(longitude_value, -180, 180),
         year_value >= start_year, year_value <= end_year) %>%
  mutate(grid_cell = paste(floor(latitude_value / 5) * 5, floor(longitude_value / 5) * 5, sep = "_")) %>%
  distinct(year_value, grid_cell) %>%
  group_by(grid_cell) %>%
  summarise(first_observed_year = min(year_value), .groups = "drop")

cumulative_grid_cells <- tibble(year_value = start_year:end_year) %>%
  rowwise() %>%
  mutate(cumulative_grid_cells = sum(grid_first_year$first_observed_year <= year_value)) %>%
  ungroup()
```

```{r cumulative-grid-coverage, fig.width=9, fig.height=4.6}
ggplot(cumulative_grid_cells, aes(x = year_value, y = cumulative_grid_cells)) +
  geom_area(fill = "#B2DF8A", alpha = 0.65) +
  geom_line(color = "#2A7F62", linewidth = 1.1) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(title = "Cumulative spatial grid coverage over time",
       subtitle = "A 5-degree latitude/longitude grid approximates whether spatial coverage expands.",
       x = "Year", y = "Cumulative occupied grid cells")
```

**Interpretation.** Source maps and map snapshots distinguish two different processes: where each platform records data, and whether records over time spread into new regions or just intensify in existing ones.

```{r unknown-source-diagnostics, results='asis'}
unknown_sources <- df %>%
  filter(source_group == "Unknown source") %>%
  count(source_raw, sort = TRUE, name = "observations") %>%
  mutate(share = observations / nrow(df))

n_unknown <- sum(unknown_sources$observations)
share_unknown <- n_unknown / nrow(df)

if (n_unknown > 0) {
  cat("# Unknown Source Diagnostics\n\n")
  cat(
    "Out of ", comma(nrow(df)), " observations, ",
    comma(n_unknown), " (", percent(share_unknown, accuracy = 0.01),
    ") come from an unknown or missing source.\n\n",
    sep = ""
  )
  print(knitr::kable(unknown_sources, align = "lrr"))
}
```

# Overall Interpretation

This analysis shows that GlobalGeoTree is **not source-neutral and not time-neutral**:

- A few platforms - mainly iNaturalist and Pl@ntNet - dominate the observation base.
- "Other source" looks big in some charts, but it is a long tail of many different national/regional datasets, not one consistent collection method (see the dedicated section above).
- Source composition varies by country, year, and taxonomic family, so geographic and taxonomic patterns are partly platform-coverage patterns.
- The overall rise in observations from 2015-2024 is driven substantially by iNaturalist's growth, and the apparent decline in `r end_year` may be a data-export artifact rather than a real drop.
- Geographic coverage genuinely expanded (more countries, more grid cells over time), but that expansion is also unevenly distributed across platforms.

**Main limitation.** These figures describe patterns in the available records. They do not prove that any source is biased in a negative way - they show where source coverage is uneven and where interpretation needs extra care. For any downstream analysis, source composition should be reported alongside geographic, temporal, or taxonomic results, and key findings should ideally be re-checked with and without the largest citizen-science source.
