---
title: "Quarto Demo"
author: "Ulrike Grömping"
date: "13 April 2026"
lang: en
format:
  html:
    toc: true
    toc-float: true
    code-fold: true
    embed-resources: true
    number-sections: true
  pdf:
    toc: false
    keep-tex: true
    number-sections: true
execute:
  echo: true
  eval: true
---

```{r}
#| label: setup
#| include: false
params <- list(html=knitr::is_html_output())
knitr::opts_chunk$set(echo = params$html)
```

# First chapter

Quarto (Allaire et al. 2026) is the latest tool for integrating analysis and reporting. It handles R, Python and Julia, and possibly further languages. In this brief demo, I demonstrate some basic aspects of its use. Using it with R requires **knitr** (see, e.g., https://yihui.org/knitr/), and the possibilities of that package can also be used. It is possible to incorporate not only R code (like in this demo), but also, e.g., Python or Julia code with the respective code chunks.

In Chapter&nbsp;@sec-meinreflabel2 you find a figure with a simple caption that was created by the `fig-cap` entry in the chunk information, and a table that was created with function `knitr::kable`. In Chapter&nbsp;@sec-meinreflabel3 you find a figure whose caption contains underscores and a formula; in Quarto, such captions can be written directly in the `fig-cap` chunk option without needing a separate reference definition (which would be needed with bookdown). A caption can contain _underscores_ and formulae directly.

You can have unnumbered chapters (while most chapters are numbered), see the appendix.

With Quarto, it is easy to reference back and forth between sections, and to figures or tables that were created from code chunks. References to figures start with `@fig-`, references to tables with `@tbl-`, and section references with `@sec-`. The chunk label itself must carry the corresponding prefix (different from bookdown, where the chunk label does not have the prefix, but the reference will).

# Second chapter {#sec-meinreflabel2}

@tbl-cor shows the correlations in the dataset `swiss`.

```{r}
#| label: tbl-cor
#| tbl-cap: "The correlation matrix of the `swiss` data"
knitr::kable(round(cor(swiss),4), booktabs=TRUE, linesep="")
```

@fig-kde shows a kernel density estimate of the proportion of males involved in agricultural occupations.

```{r}
#| label: fig-kde
#| fig-cap: "Proportion of males involved in agricultural occupations in 47 French speaking swiss provinces (1888). Kernel density estimate."
#| fig-height: 3
#| fig-width: 4
plot(density(swiss$Agriculture), xlab="Agriculture", las=1, mgp=c(2,0.5,0))
```

# Third chapter {#sec-meinreflabel3}

@fig-scat shows a scatter plot with a more demanding figure caption.

```{r}
#| label: fig-scat
#| fig-cap: "A caption can contain \\_underscores\\_ and $\\sqrt v$ formulae."
#| fig-height: 3
#| fig-width: 4
plot(Fertility~Agriculture, swiss, las=1, mgp=c(2,0.5,0))
```

# Final chapter

Your final chapter ---&nbsp;like your introductory chapter ---&nbsp;will be read by every reader. Both these chapters are therefore particularly important.

Please note that this brief text focused on how to implement a report with the help of R package **bookdown**. There was no content of specific relevance for data visualization.

# References {.unnumbered}

Allaire, J., Teague, C., Scheidegger, C., Xie, Y., Dervieux, C., & Woodhull, G. (2026). Quarto (Version 1.9) [Computer software]. https://doi.org/10.5281/zenodo.5960048

# Appendix A {.unnumbered}

Example for an appendix

The following listing shows the data (do not do that with a large data set!).

```{r}
#| label: printdat
print(swiss)
```
