---
title: "RQ2 - Source and Platform Bias in GlobalGeoTree"
subtitle: "How data sources shape geographic and taxonomic patterns"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 3
    number-sections: false
    documentclass: article
    pdf-engine: xelatex
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Required packages:
# install.packages(c("data.table", "dplyr", "stringr", "ggplot2", "scales"))
#
# Optional for the country map:
# install.packages("maps")

library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)

theme_gtree <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(color = "gray35"),
      plot.caption = element_text(color = "gray45", size = base_size - 2),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      strip.background = element_rect(fill = "gray94", color = NA),
      plot.margin = margin(8, 14, 8, 8)
    )
}

theme_set(theme_gtree())

source_palette <- c(
  "iNaturalist" = "#8ECAE6",
  "Pl@ntNet" = "#FF9DA7",
  "GBIF" = "#4E79A7",
  "Wikidata/Wikipedia" = "#B07AA1",
  "Institutional collections" = "#1F77B4",
  "Inventory/survey" = "#B2DF8A",
  "Other source" = "#2CA02C",
  "Unknown source" = "#E41A1C",
  "Other or unclear source" = "#5B8FA8",
  "Citizen-science source" = "#C97E4F"
)

source_group_levels <- c(
  "iNaturalist",
  "Pl@ntNet",
  "GBIF",
  "Wikidata/Wikipedia",
  "Institutional collections",
  "Inventory/survey",
  "Other source",
  "Unknown source"
)
```

# Research Question

**RQ2 - Source and Platform Bias**  
To what extent do data sources shape the observed geographic and taxonomic patterns in **GlobalGeoTree**?

This report asks:

1. Which sources contribute the largest share of observations?
2. Is the dataset dominated by citizen-science platforms such as iNaturalist?
3. Do different sources cover different countries or regions?
4. Are some sources associated with specific years, countries, or species groups?

The analysis is descriptive. Strong source-specific patterns should be interpreted as evidence of **recording and platform bias**, not only as biological patterns in tree distributions.

# Data Loading

```{r user-config}
# -------------------------------------------------------------------------
# User configuration
# -------------------------------------------------------------------------

# Put this .qmd in the same folder as your data file.
# The full dataset is called GlobalGeoTree, without "_full".
observations_path <- "GlobalGeoTree.csv"

# Optional category lookup. This is used only if GlobalGeoTree itself has no
# Rare/Common/Frequent column.
eval_category_path <- "GlobalGeoTree-10kEval-900.csv"

# If the dataset is very large, set use_subsample <- TRUE for faster drafts.
# Use the full data for final counts.
use_subsample <- FALSE
sample_n <- 500000
set.seed(2026)

top_n_sources <- 10
top_n_countries <- 15
top_n_taxa <- 12
top_n_sources_for_maps <- 4
map_point_sample_n <- 200000
```

```{r helper-functions}
normalize_names <- function(x) {
  x |>
    str_replace_all("([a-z0-9])([A-Z])", "\\1_\\2") |>
    str_replace_all("[^A-Za-z0-9]+", "_") |>
    str_replace_all("_+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()
}

first_existing <- function(data, candidates, label, required = TRUE) {
  found <- candidates[candidates %in% names(data)]

  if (length(found) > 0) {
    return(found[[1]])
  }

  if (required) {
    stop(
      paste0(
        "Could not find a column for ", label, ". Tried: ",
        paste(candidates, collapse = ", "),
        "\nAvailable columns are:\n",
        paste(names(data), collapse = ", ")
      ),
      call. = FALSE
    )
  }

  NA_character_
}

clean_source_label <- function(x) {
  x <- as.character(x)
  x <- ifelse(is.na(x) | str_trim(x) == "", "Unknown source", x)
  str_squish(x)
}

source_group_label <- function(x) {
  source_clean <- str_to_lower(clean_source_label(x))

  case_when(
    str_detect(source_clean, "inaturalist|i-naturalist|inat") ~ "iNaturalist",
    str_detect(source_clean, "plantnet|pl@ntnet") ~ "Pl@ntNet",
    str_detect(source_clean, "gbif") ~ "GBIF",
    str_detect(source_clean, "wikidata|wikipedia") ~ "Wikidata/Wikipedia",
    str_detect(source_clean, "herbar|museum|botanic|arboretum") ~ "Institutional collections",
    str_detect(source_clean, "inventory|census|survey|monitoring") ~ "Inventory/survey",
    str_detect(source_clean, "unknown|missing|na") ~ "Unknown source",
    TRUE ~ "Other source"
  )
}

shorten_label <- function(x, width = 34) {
  str_trunc(as.character(x), width = width)
}

source_fill_scale <- function(..., na.value = "gray75") {
  scale_fill_manual(values = source_palette, drop = FALSE, na.value = na.value, ...)
}

source_color_scale <- function(..., na.value = "gray75") {
  scale_color_manual(values = source_palette, drop = FALSE, na.value = na.value, ...)
}

shannon_diversity <- function(x) {
  p <- x / sum(x)
  p <- p[p > 0]
  -sum(p * log(p))
}
```

```{r load-data}
if (!file.exists(observations_path)) {
  stop(
    paste0(
      "Could not find ", observations_path, ". Put this QMD in the same folder ",
      "as GlobalGeoTree.csv, or edit observations_path in the configuration chunk."
    ),
    call. = FALSE
  )
}

GlobalGeoTree <- fread(observations_path, showProgress = FALSE)
names(GlobalGeoTree) <- normalize_names(names(GlobalGeoTree))

if (use_subsample && nrow(GlobalGeoTree) > sample_n) {
  GlobalGeoTree <- GlobalGeoTree |>
    slice_sample(n = sample_n)
}

source_col <- first_existing(
  GlobalGeoTree,
  c(
    "source", "sources", "data_source", "dataset_source", "platform",
    "publisher", "publisher_title", "institution_code", "dataset_name",
    "basis_of_record", "recorded_by"
  ),
  "observation source"
)

country_col <- first_existing(
  GlobalGeoTree,
  c("country_code", "countrycode", "iso_a2", "iso2", "country", "level0_country", "country_name"),
  "country"
)

species_col <- first_existing(
  GlobalGeoTree,
  c("level3_species", "species", "scientific_name", "scientificname", "accepted_scientific_name", "taxon_name"),
  "species"
)

species_key_col <- first_existing(
  GlobalGeoTree,
  c("species_key", "specieskey", "taxon_key", "taxonkey", "gbif_species_key", "gbif_taxon_key"),
  "species key",
  required = FALSE
)

genus_col <- first_existing(GlobalGeoTree, c("level2_genus", "genus"), "genus", required = FALSE)
family_col <- first_existing(GlobalGeoTree, c("level1_family", "family"), "family", required = FALSE)
year_col <- first_existing(GlobalGeoTree, c("year", "observation_year", "event_year", "recorded_year"), "year", required = FALSE)
date_col <- first_existing(GlobalGeoTree, c("event_date", "eventdate", "date", "observed_on", "observation_date", "recorded_date"), "date", required = FALSE)
lat_col <- first_existing(GlobalGeoTree, c("latitude", "decimal_latitude", "decimallatitude", "lat", "y"), "latitude", required = FALSE)
lon_col <- first_existing(GlobalGeoTree, c("longitude", "decimal_longitude", "decimallongitude", "lon", "lng", "x"), "longitude", required = FALSE)
category_col <- first_existing(GlobalGeoTree, c("category", "frequency_category", "species_category", "rarity_category"), "Rare/Common/Frequent category", required = FALSE)

if (is.na(year_col) && !is.na(date_col)) {
  GlobalGeoTree <- GlobalGeoTree |>
    mutate(year_from_date = suppressWarnings(as.integer(str_sub(as.character(.data[[date_col]]), 1, 4))))
  year_col <- "year_from_date"
}

GlobalGeoTree <- GlobalGeoTree |>
  mutate(
    source_raw = clean_source_label(.data[[source_col]]),
    source_group = factor(source_group_label(source_raw), levels = source_group_levels),
    country_value = str_to_upper(str_squish(as.character(.data[[country_col]]))),
    species_value = str_squish(as.character(.data[[species_col]])),
    species_key_value = if (!is.na(species_key_col)) str_squish(as.character(.data[[species_key_col]])) else NA_character_,
    genus_value = if (!is.na(genus_col)) str_squish(as.character(.data[[genus_col]])) else NA_character_,
    family_value = if (!is.na(family_col)) str_squish(as.character(.data[[family_col]])) else NA_character_,
    category_value = if (!is.na(category_col)) str_squish(as.character(.data[[category_col]])) else NA_character_,
    year_value = if (!is.na(year_col)) suppressWarnings(as.integer(.data[[year_col]])) else NA_integer_,
    latitude_value = if (!is.na(lat_col)) suppressWarnings(as.numeric(.data[[lat_col]])) else NA_real_,
    longitude_value = if (!is.na(lon_col)) suppressWarnings(as.numeric(.data[[lon_col]])) else NA_real_
  ) |>
  filter(!is.na(source_raw), !is.na(country_value), !is.na(species_value))

category_source_label <- if (!is.na(category_col)) {
  "GlobalGeoTree category column"
} else {
  "Not available"
}
```

```{r optional-category-lookup}
if (is.na(category_col) && file.exists(eval_category_path)) {
  eval_data <- fread(eval_category_path, showProgress = FALSE)
  names(eval_data) <- normalize_names(names(eval_data))

  eval_category_col <- first_existing(
    eval_data,
    c("category", "frequency_category", "species_category", "rarity_category"),
    "evaluation category",
    required = FALSE
  )

  eval_species_key_col <- first_existing(
    eval_data,
    c("species_key", "specieskey", "taxon_key", "taxonkey", "gbif_species_key", "gbif_taxon_key"),
    "evaluation species key",
    required = FALSE
  )

  eval_species_col <- first_existing(
    eval_data,
    c("level3_species", "species", "scientific_name", "scientificname", "accepted_scientific_name", "taxon_name"),
    "evaluation species",
    required = FALSE
  )

  if (!is.na(eval_category_col)) {
    if (!is.na(species_key_col) && !is.na(eval_species_key_col)) {
      category_lookup <- eval_data |>
        transmute(
          species_key_value = str_squish(as.character(.data[[eval_species_key_col]])),
          category_lookup = str_squish(as.character(.data[[eval_category_col]]))
        ) |>
        filter(!is.na(species_key_value), species_key_value != "", !is.na(category_lookup), category_lookup != "") |>
        group_by(species_key_value) |>
        summarise(category_lookup = first(category_lookup), .groups = "drop")

      GlobalGeoTree <- GlobalGeoTree |>
        left_join(category_lookup, by = "species_key_value") |>
        mutate(category_value = category_lookup) |>
        select(-category_lookup)

      category_source_label <- paste0(eval_category_path, " joined by species_key")
    } else if (!is.na(eval_species_col)) {
      category_lookup <- eval_data |>
        transmute(
          species_value = str_squish(as.character(.data[[eval_species_col]])),
          category_lookup = str_squish(as.character(.data[[eval_category_col]]))
        ) |>
        filter(!is.na(species_value), species_value != "", !is.na(category_lookup), category_lookup != "") |>
        group_by(species_value) |>
        summarise(category_lookup = first(category_lookup), .groups = "drop")

      GlobalGeoTree <- GlobalGeoTree |>
        left_join(category_lookup, by = "species_value") |>
        mutate(category_value = category_lookup) |>
        select(-category_lookup)

      category_source_label <- paste0(eval_category_path, " joined by species name")
    }
  }
}

source_note <- if (use_subsample) {
  paste0("This report uses a random subsample of ", comma(nrow(GlobalGeoTree)), " observations.")
} else {
  paste0("This report uses ", comma(nrow(GlobalGeoTree)), " observations.")
}
```

```{r data-summary}
category_known_observations <- sum(!is.na(GlobalGeoTree$category_value) & GlobalGeoTree$category_value != "")
category_known_species <- GlobalGeoTree |>
  filter(!is.na(category_value), category_value != "") |>
  summarise(n_species = n_distinct(species_value)) |>
  pull(n_species)

data_summary <- tibble(
  Measure = c(
    "Observation file",
    "Rows analysed",
    "Distinct sources",
    "Distinct source groups",
    "Distinct countries",
    "Distinct species",
    "Year column available",
    "Coordinate columns available",
    "Category source",
    "Observations with known category",
    "Species with known category"
  ),
  Value = c(
    observations_path,
    comma(nrow(GlobalGeoTree)),
    comma(n_distinct(GlobalGeoTree$source_raw)),
    comma(n_distinct(GlobalGeoTree$source_group)),
    comma(n_distinct(GlobalGeoTree$country_value)),
    comma(n_distinct(GlobalGeoTree$species_value)),
    ifelse(!is.na(year_col), "Yes", "No"),
    ifelse(!is.na(lat_col) && !is.na(lon_col), "Yes", "No"),
    category_source_label,
    comma(category_known_observations),
    comma(category_known_species)
  )
)

knitr::kable(data_summary, align = "ll")
```

`r source_note`

# Headline Result

```{r source-share-data}
source_counts <- GlobalGeoTree |>
  count(source_raw, sort = TRUE, name = "observations") |>
  mutate(
    share = observations / sum(observations),
    cumulative_share = cumsum(share),
    source_rank = row_number(),
    source_short = reorder(shorten_label(source_raw), observations)
  )

source_group_counts <- GlobalGeoTree |>
  count(source_group, sort = TRUE, name = "observations") |>
  mutate(
    share = observations / sum(observations),
    source_group = factor(source_group, levels = source_group_levels)
  )

top_sources <- source_counts |>
  slice_head(n = top_n_sources)

inat_share <- source_group_counts |>
  filter(source_group == "iNaturalist") |>
  pull(share)

if (length(inat_share) == 0) {
  inat_share <- 0
}

citizen_groups <- c("iNaturalist", "Pl@ntNet")
citizen_share <- source_group_counts |>
  filter(source_group %in% citizen_groups) |>
  summarise(share = sum(share)) |>
  pull(share)

if (length(citizen_share) == 0) {
  citizen_share <- 0
}
```

The source structure is highly concentrated: **iNaturalist alone contributes `r percent(inat_share, accuracy = 0.1)`** of observations, while **iNaturalist and Pl@ntNet together contribute `r percent(citizen_share, accuracy = 0.1)`**. Therefore, many observed geographic, temporal, and taxonomic patterns should be read partly as platform-coverage patterns.

# Source Concentration

```{r source-share-bar, fig.width=9.5, fig.height=5.8}
ggplot(top_sources, aes(x = source_short, y = observations)) +
  geom_col(fill = "#2A7F62", width = 0.72) +
  geom_text(
    aes(label = percent(share, accuracy = 0.1)),
    hjust = -0.08,
    size = 3.2
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Largest observation sources in GlobalGeoTree",
    subtitle = "The top source is large enough to shape the overall dataset.",
    x = NULL,
    y = "Observations",
    caption = "Long source names are shortened for readability."
  )
```

```{r source-pareto, fig.width=9.5, fig.height=5}
pareto_sources <- source_counts |>
  slice_head(n = 50)

ggplot(pareto_sources, aes(x = source_rank)) +
  geom_col(aes(y = share), fill = "#2A7F62", alpha = 0.75, width = 0.85) +
  geom_line(aes(y = cumulative_share), color = "#C23B22", linewidth = 1) +
  geom_point(aes(y = cumulative_share), color = "#C23B22", size = 1.4) +
  scale_x_continuous(breaks = c(1, 5, 10, 25, 50)) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(
    title = "A small number of sources explain most observations",
    subtitle = "Bars show each source's individual share; the red line shows cumulative share.",
    x = "Source rank",
    y = "Share of all observations"
  )
```

```{r source-group-bar, fig.width=8.5, fig.height=4.8}
ggplot(source_group_counts, aes(x = reorder(source_group, observations), y = observations, fill = source_group)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_text(
    aes(label = percent(share, accuracy = 0.1)),
    hjust = -0.08,
    size = 3.3
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  source_fill_scale() +
  labs(
    title = "Observation share by broad source group",
    subtitle = "Citizen-science platforms dominate the grouped source structure.",
    x = NULL,
    y = "Observations"
  )
```

```{r source-table}
source_table <- top_sources |>
  transmute(
    Source = source_raw,
    Observations = comma(observations),
    Share = percent(share, accuracy = 0.1),
    `Cumulative share` = percent(cumulative_share, accuracy = 0.1)
  )

knitr::kable(source_table, align = "lrrr")
```

**Interpretation.** The distribution is not balanced across sources. A few large platforms dominate the evidence base, so global summaries are strongly affected by the behaviour, geography, and taxonomy covered by those platforms.

# Citizen-Science Dominance

```{r citizen-science-data}
citizen_summary <- GlobalGeoTree |>
  mutate(
    source_type = if_else(
      as.character(source_group) %in% citizen_groups,
      "Citizen-science source",
      "Other or unclear source"
    )
  ) |>
  count(source_type, name = "observations") |>
  mutate(
    share = observations / sum(observations),
    source_type = factor(source_type, levels = c("Other or unclear source", "Citizen-science source"))
  )
```

```{r citizen-science-bar, fig.width=7.5, fig.height=4.2}
ggplot(citizen_summary, aes(x = source_type, y = observations, fill = source_type)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = percent(share, accuracy = 0.1)), vjust = -0.45, size = 4) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.14))) +
  source_fill_scale() +
  labs(
    title = "Citizen-science platforms dominate the observations",
    subtitle = "Citizen-science is defined here as iNaturalist plus Pl@ntNet.",
    x = NULL,
    y = "Observations"
  )
```

**Interpretation.** Citizen-science dominance does not make the data unusable, but it changes how the data should be interpreted. Places and species that are popular with users or easy to photograph may be overrepresented.

# Countries and Regional Coverage

```{r country-source-data}
top_country_values <- GlobalGeoTree |>
  count(country_value, sort = TRUE, name = "observations") |>
  slice_head(n = top_n_countries) |>
  pull(country_value)

top_source_groups <- source_group_counts |>
  arrange(desc(observations)) |>
  slice_head(n = top_n_sources) |>
  pull(source_group) |>
  as.character()

country_source <- GlobalGeoTree |>
  filter(country_value %in% top_country_values) |>
  mutate(source_group_plot = if_else(as.character(source_group) %in% top_source_groups, as.character(source_group), "Other source")) |>
  count(country_value, source_group_plot, name = "observations") |>
  group_by(country_value) |>
  mutate(
    country_total = sum(observations),
    share = observations / country_total
  ) |>
  ungroup()

country_order <- country_source |>
  distinct(country_value, country_total) |>
  arrange(country_total) |>
  pull(country_value)

country_source <- country_source |>
  mutate(
    country_value = factor(country_value, levels = country_order),
    source_group_plot = factor(source_group_plot, levels = source_group_levels)
  )

country_dominant <- GlobalGeoTree |>
  count(country_value, source_group, name = "observations") |>
  group_by(country_value) |>
  mutate(country_total = sum(observations), share = observations / country_total) |>
  slice_max(observations, n = 1, with_ties = FALSE) |>
  ungroup()
```

```{r country-source-stacked-share, fig.width=9.5, fig.height=6}
ggplot(country_source, aes(x = country_value, y = share, fill = source_group_plot)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  source_fill_scale() +
  labs(
    title = "Source composition differs by country",
    subtitle = paste0("Top ", top_n_countries, " countries by observations; bars sum to 100% within each country."),
    x = NULL,
    y = "Share of country observations",
    fill = "Source group"
  )
```

```{r country-source-heatmap, fig.width=9.5, fig.height=6}
ggplot(country_source, aes(x = source_group_plot, y = country_value, fill = share)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = if_else(share >= 0.08, percent(share, accuracy = 1), "")), size = 3) +
  scale_fill_viridis_c(labels = percent, option = "C", limits = c(0, 1)) +
  labs(
    title = "Country-by-source heatmap",
    subtitle = "Darker cells show countries that rely heavily on one source group.",
    x = NULL,
    y = NULL,
    fill = "Country share"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r country-source-stacked-count, fig.width=9.5, fig.height=6}
ggplot(country_source, aes(x = country_value, y = observations, fill = source_group_plot)) +
  geom_col(width = 0.75) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.08))) +
  source_fill_scale() +
  labs(
    title = "Observation volume by country and source group",
    subtitle = "Country comparisons combine source composition and very unequal record volumes.",
    x = NULL,
    y = "Observations",
    fill = "Source group"
  )
```

**Interpretation.** Countries differ both in the number of observations and in the source mix. This means that geographic comparisons can be biased by platform adoption. A country dominated by iNaturalist is not directly comparable to a country dominated by local inventories or unknown sources.

## Source Diversity by Country

```{r country-diversity-data}
country_diversity <- GlobalGeoTree |>
  count(country_value, source_group, name = "observations") |>
  group_by(country_value) |>
  summarise(
    total_observations = sum(observations),
    n_source_groups = n_distinct(source_group),
    shannon_source_diversity = shannon_diversity(observations),
    effective_sources = exp(shannon_source_diversity),
    dominant_source = as.character(source_group[which.max(observations)]),
    dominant_source_share = max(observations) / sum(observations),
    .groups = "drop"
  ) |>
  arrange(desc(total_observations))

country_diversity_top <- country_diversity |>
  slice_head(n = top_n_countries) |>
  mutate(country_value = factor(country_value, levels = rev(country_value)))
```

```{r source-diversity-country, fig.width=9.5, fig.height=5.6}
ggplot(country_diversity_top, aes(x = country_value, y = effective_sources, fill = dominant_source)) +
  geom_col(width = 0.72) +
  coord_flip() +
  source_fill_scale() +
  labs(
    title = "How source-diverse are the largest country samples?",
    subtitle = "Effective number of source groups is low when one source dominates a country.",
    x = NULL,
    y = "Effective number of source groups",
    fill = "Dominant source"
  )
```

```{r inat-share-country, fig.width=9.5, fig.height=5.5}
inat_country <- GlobalGeoTree |>
  mutate(is_inaturalist = as.character(source_group) == "iNaturalist") |>
  group_by(country_value) |>
  summarise(
    observations = n(),
    inaturalist_share = mean(is_inaturalist),
    .groups = "drop"
  ) |>
  arrange(desc(observations)) |>
  slice_head(n = top_n_countries) |>
  mutate(country_value = factor(country_value, levels = rev(country_value)))

ggplot(inat_country, aes(x = country_value, y = inaturalist_share)) +
  geom_col(fill = "#8ECAE6", width = 0.72) +
  geom_text(aes(label = percent(inaturalist_share, accuracy = 1)), hjust = -0.08, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent, limits = c(0, 1), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "iNaturalist share by country",
    subtitle = paste0("Top ", top_n_countries, " countries by observation count."),
    x = NULL,
    y = "Share from iNaturalist"
  )
```

**Interpretation.** Source diversity provides a compact way to see which countries depend on one platform and which countries combine several source groups. Low diversity means higher risk that country-level patterns are platform-specific.

## Optional Map: Dominant Source by Country

```{r dominant-country-map, fig.width=9.5, fig.height=5.4}
if (requireNamespace("maps", quietly = TRUE)) {
  world_map <- ggplot2::map_data("world") |>
    mutate(country_join = str_to_upper(region))

  country_name_lookup <- c(
    "US" = "USA",
    "GB" = "UK",
    "UK" = "UK",
    "DE" = "GERMANY",
    "FR" = "FRANCE",
    "ES" = "SPAIN",
    "IT" = "ITALY",
    "AU" = "AUSTRALIA",
    "CA" = "CANADA",
    "BR" = "BRAZIL",
    "CO" = "COLOMBIA",
    "RU" = "RUSSIA",
    "SE" = "SWEDEN",
    "NO" = "NORWAY",
    "BE" = "BELGIUM",
    "NZ" = "NEW ZEALAND",
    "AT" = "AUSTRIA",
    "ZA" = "SOUTH AFRICA",
    "TW" = "TAIWAN"
  )

  map_country_source <- country_dominant |>
    mutate(country_join = if_else(country_value %in% names(country_name_lookup), country_name_lookup[country_value], country_value)) |>
    select(country_join, source_group, share, country_total)

  world_source_map <- world_map |>
    left_join(map_country_source, by = "country_join")

  ggplot(world_source_map, aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = source_group), color = "white", linewidth = 0.08) +
    coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    source_fill_scale(na.value = "gray92") +
    labs(
      title = "Dominant source group by country",
      subtitle = "Countries are colored by the source group contributing the most observations.",
      x = NULL,
      y = NULL,
      fill = "Dominant source",
      caption = "This optional map uses the maps package and a small built-in country-name lookup."
    ) +
    theme(axis.text = element_blank(), panel.grid = element_blank())
} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "Install the optional maps package to render the dominant-source country map.",
      size = 4
    ) +
    theme_void()
}
```

# Sources Over Time

```{r source-year-data}
if (!is.na(year_col)) {
  source_year <- GlobalGeoTree |>
    filter(!is.na(year_value), year_value >= 1900, year_value <= as.integer(format(Sys.Date(), "%Y"))) |>
    mutate(source_group_plot = if_else(as.character(source_group) %in% top_source_groups, as.character(source_group), "Other source")) |>
    count(year_value, source_group_plot, name = "observations") |>
    group_by(year_value) |>
    mutate(
      year_total = sum(observations),
      share = observations / year_total
    ) |>
    ungroup() |>
    mutate(source_group_plot = factor(source_group_plot, levels = source_group_levels))

  if (nrow(source_year) > 0) {
    year_breaks <- seq(min(source_year$year_value, na.rm = TRUE), max(source_year$year_value, na.rm = TRUE), by = 1)
  }
}
```

```{r source-year-stacked, fig.width=9.5, fig.height=5}
if (!is.na(year_col) && exists("source_year") && nrow(source_year) > 0) {
  ggplot(source_year, aes(x = year_value, y = observations, fill = source_group_plot)) +
    geom_col(width = 0.9) +
    scale_x_continuous(breaks = year_breaks) +
    scale_y_continuous(labels = comma) +
    source_fill_scale() +
    labs(
      title = "Observation sources by year",
      subtitle = "The rise of a platform can look like a temporal trend in tree observations.",
      x = "Year",
      y = "Observations",
      fill = "Source group"
    ) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable year column was detected.", size = 4) +
    theme_void()
}
```

```{r source-year-share, fig.width=9.5, fig.height=5}
if (!is.na(year_col) && exists("source_year") && nrow(source_year) > 0) {
  ggplot(source_year, aes(x = year_value, y = share, fill = source_group_plot)) +
    geom_col(width = 0.9) +
    scale_x_continuous(breaks = year_breaks) +
    scale_y_continuous(labels = percent) +
    source_fill_scale() +
    labs(
      title = "Source composition by year",
      subtitle = "Bars sum to 100% within each year, revealing shifts in source dominance.",
      x = "Year",
      y = "Share of observations",
      fill = "Source group",
      caption = "The most recent year may be incomplete depending on the data export date."
    ) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable year column was detected.", size = 4) +
    theme_void()
}
```

**Interpretation.** Temporal patterns are not source-neutral. If iNaturalist or Pl@ntNet grows in recent years, the dataset may show an apparent rise in observations even if actual tree occurrence is unchanged.

# Sources and Taxonomic Groups

This section checks whether source groups disproportionately record particular families, genera, or species. Family is used first when available because it is more readable than hundreds of species names.

```{r taxonomic-source-data}
taxon_level <- case_when(
  !is.na(family_col) ~ "family",
  !is.na(genus_col) ~ "genus",
  TRUE ~ "species"
)

GlobalGeoTree <- GlobalGeoTree |>
  mutate(
    taxon_group = case_when(
      taxon_level == "family" ~ family_value,
      taxon_level == "genus" ~ genus_value,
      TRUE ~ species_value
    ),
    taxon_group = if_else(is.na(taxon_group) | taxon_group == "", "Unknown taxon", taxon_group)
  )

top_taxa <- GlobalGeoTree |>
  count(taxon_group, sort = TRUE, name = "observations") |>
  slice_head(n = top_n_taxa) |>
  pull(taxon_group)

taxon_source <- GlobalGeoTree |>
  filter(taxon_group %in% top_taxa) |>
  mutate(source_group_plot = if_else(as.character(source_group) %in% top_source_groups, as.character(source_group), "Other source")) |>
  count(taxon_group, source_group_plot, name = "observations") |>
  group_by(taxon_group) |>
  mutate(
    taxon_total = sum(observations),
    share = observations / taxon_total
  ) |>
  ungroup()

taxon_order <- taxon_source |>
  distinct(taxon_group, taxon_total) |>
  arrange(taxon_total) |>
  pull(taxon_group)

taxon_source <- taxon_source |>
  mutate(
    taxon_group = factor(taxon_group, levels = taxon_order),
    source_group_plot = factor(source_group_plot, levels = source_group_levels)
  )
```

```{r taxon-source-share, fig.width=9.5, fig.height=6}
ggplot(taxon_source, aes(x = taxon_group, y = share, fill = source_group_plot)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  source_fill_scale() +
  labs(
    title = paste("Source composition by", taxon_level),
    subtitle = "Bars sum to 100% within each taxonomic group.",
    x = NULL,
    y = "Share of observations",
    fill = "Source group"
  )
```

```{r taxon-source-heatmap, fig.width=9.5, fig.height=5.8}
ggplot(taxon_source, aes(x = source_group_plot, y = taxon_group, fill = share)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = if_else(share >= 0.10, percent(share, accuracy = 1), "")), size = 3) +
  scale_fill_viridis_c(labels = percent, option = "C", limits = c(0, 1)) +
  labs(
    title = paste("Taxonomic groups are not equally represented across sources"),
    subtitle = paste("Heatmap by", taxon_level, "and source group."),
    x = NULL,
    y = NULL,
    fill = "Taxon share"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

**Interpretation.** If a family or genus is mostly recorded by one platform, its geographic pattern may partly reflect that platform's user base and coverage.

# Geographic Footprints of Major Sources

```{r source-map-data}
has_coordinates <- !is.na(lat_col) && !is.na(lon_col)

if (has_coordinates) {
  map_sources <- source_group_counts |>
    arrange(desc(observations)) |>
    filter(!source_group %in% c("Unknown source", "Other source")) |>
    slice_head(n = top_n_sources_for_maps) |>
    pull(source_group) |>
    as.character()

  map_data <- GlobalGeoTree |>
    filter(
      as.character(source_group) %in% map_sources,
      !is.na(latitude_value),
      !is.na(longitude_value),
      between(latitude_value, -90, 90),
      between(longitude_value, -180, 180)
    )

  if (nrow(map_data) > map_point_sample_n) {
    map_data <- map_data |>
      slice_sample(n = map_point_sample_n)
  }

  world_outline <- NULL
  if (requireNamespace("maps", quietly = TRUE)) {
    world_outline <- ggplot2::map_data("world")
  }
}
```

```{r source-faceted-map, fig.width=9.5, fig.height=6.8}
if (has_coordinates && exists("map_data") && nrow(map_data) > 0) {
  p <- ggplot()

  if (exists("world_outline") && !is.null(world_outline)) {
    p <- p +
      geom_polygon(
        data = world_outline,
        aes(x = long, y = lat, group = group),
        fill = "gray96",
        color = "gray82",
        linewidth = 0.12,
        inherit.aes = FALSE
      )
  }

  p +
    geom_point(
      data = map_data,
      aes(x = longitude_value, y = latitude_value),
      alpha = 0.12,
      size = 0.22,
      color = "#1F5B74"
    ) +
    facet_wrap(~ source_group, ncol = 2) +
    coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    labs(
      title = "Geographic coverage of major source groups",
      subtitle = "Different platforms leave visibly different spatial footprints.",
      x = "Longitude",
      y = "Latitude",
      caption = "Points may be sampled for readability. Install maps for country outlines."
    )
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable latitude/longitude columns were detected.", size = 4) +
    theme_void()
}
```

```{r source-density-map, fig.width=9.5, fig.height=6.8}
if (has_coordinates && exists("map_data") && nrow(map_data) > 0) {
  ggplot(map_data, aes(x = longitude_value, y = latitude_value)) +
    stat_bin_2d(bins = 85) +
    facet_wrap(~ source_group, ncol = 2) +
    coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    scale_fill_viridis_c(trans = "sqrt", labels = comma, option = "C") +
    labs(
      title = "Observation density by source group",
      subtitle = "Binned counts reveal dense recording regions more clearly than raw points.",
      x = "Longitude",
      y = "Latitude",
      fill = "Records"
    ) +
    theme(legend.position = "right")
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable latitude/longitude columns were detected.", size = 4) +
    theme_void()
}
```

**Interpretation.** The source maps show that platform coverage is geographically uneven. Some regions are visible mainly because they are covered by a particular platform or source pipeline.

# Unknown Source Diagnostics

```{r unknown-source-data}
unknown_country <- GlobalGeoTree |>
  mutate(is_unknown_source = as.character(source_group) == "Unknown source") |>
  group_by(country_value) |>
  summarise(
    observations = n(),
    unknown_observations = sum(is_unknown_source),
    unknown_share = unknown_observations / observations,
    .groups = "drop"
  ) |>
  filter(observations >= 1000) |>
  arrange(desc(unknown_observations)) |>
  slice_head(n = top_n_countries) |>
  mutate(country_value = factor(country_value, levels = rev(country_value)))
```

```{r unknown-source-plot, fig.width=9.5, fig.height=5.4}
ggplot(unknown_country, aes(x = country_value, y = unknown_observations, fill = unknown_share)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = percent(unknown_share, accuracy = 1)), hjust = -0.08, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.14))) +
  scale_fill_viridis_c(labels = percent, option = "C") +
  labs(
    title = "Where are unknown-source observations concentrated?",
    subtitle = "Countries with many unknown-source records need cautious interpretation.",
    x = NULL,
    y = "Unknown-source observations",
    fill = "Unknown share"
  )
```

**Interpretation.** Unknown source records are a metadata-quality issue. If they are concentrated in particular countries, those countries have weaker evidence for source-specific interpretation.

# Rare, Common, and Frequent Categories

This section is shown only when category labels are available either inside **GlobalGeoTree** or through the optional eval lookup file.

```{r category-source-data}
has_category <- category_known_observations > 0

if (has_category) {
  category_source <- GlobalGeoTree |>
    filter(!is.na(category_value), category_value != "") |>
    mutate(source_group_plot = if_else(as.character(source_group) %in% top_source_groups, as.character(source_group), "Other source")) |>
    count(category_value, source_group_plot, name = "observations") |>
    group_by(category_value) |>
    mutate(share = observations / sum(observations)) |>
    ungroup() |>
    mutate(source_group_plot = factor(source_group_plot, levels = source_group_levels))

  category_coverage <- tibble(
    Measure = c("Category source", "Observations with known category", "Species with known category"),
    Value = c(category_source_label, comma(category_known_observations), comma(category_known_species))
  )
}
```

```{r category-coverage-table}
if (has_category) {
  knitr::kable(category_coverage, align = "ll")
}
```

```{r category-source-plot, fig.width=8.5, fig.height=4.8}
if (has_category && exists("category_source") && nrow(category_source) > 0) {
  ggplot(category_source, aes(x = category_value, y = share, fill = source_group_plot)) +
    geom_col(width = 0.75) +
    scale_y_continuous(labels = percent) +
    source_fill_scale() +
    labs(
      title = "Source composition by species frequency category",
      subtitle = "This checks whether Rare/Common/Frequent records come from different source mixtures.",
      x = "Category",
      y = "Share of observations",
      fill = "Source group"
    )
}
```

```{r no-category-note, results='asis'}
if (!has_category) {
  cat(
    "No Rare/Common/Frequent category information was available. ",
    "To enable this section, put an eval file such as `GlobalGeoTree-10kEval-900.csv` ",
    "in the same folder or join a category lookup before rendering."
  )
}
```

# Overall Interpretation

GlobalGeoTree is not source-neutral. The dataset contains millions of observations from thousands of named sources, but the observation structure is dominated by a small number of platforms, especially iNaturalist and Pl@ntNet. Source composition also varies by country, year, and taxonomic group. Therefore, observed geographic and taxonomic patterns should be interpreted as a combination of tree occurrence, recording effort, platform popularity, and source-specific data integration.

For downstream analyses, consider reporting source composition next to any global, country-level, temporal, or taxonomic result. Where possible, repeat key analyses with and without the largest citizen-science source to test robustness.
