---
title: "RQ3 - Temporal Development in GlobalGeoTree"
subtitle: "Changes in tree observation volume, source composition, and geographic coverage from 2015 to 2024"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# packages used in this report
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)

# a simple shared look for all plots, so I don't have to repeat
# the same theme() settings in every single chunk
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "gray35"),
      plot.caption = element_text(color = "gray45", size = 9),
      legend.position = "bottom"
    )
)

# one color per source group, used consistently in every chart so the
# same source always has the same color across the whole report.
# GBIF and Institutional collections were originally two similar blues
# and were easy to mix up next to iNaturalist, so I gave them their own colors.
source_colors <- c(
  "iNaturalist"                = "#8ECAE6",
  "Pl@ntNet"                   = "#FF9DA7",
  "GBIF"                       = "#264653",
  "Wikidata/Wikipedia"         = "#B07AA1",
  "Institutional collections"  = "#9C6644",
  "Inventory/survey"           = "#B2DF8A",
  "Other source"               = "#2CA02C",
  "Unknown source"             = "#E41A1C"
)
source_group_levels <- names(source_colors)
```

# Research Question

**RQ3 - Temporal Development**  
How has the number and distribution of tree observations changed from **2015 to 2024**?

This report answers four subquestions:

1. Has the number of observations increased over time?
2. Which years show the strongest growth or decline?
3. Did geographic coverage expand over time?
4. Are temporal patterns mainly driven by specific sources?

The key idea is that time trends in GlobalGeoTree may reflect both real changes in observation coverage and the growth of data platforms such as iNaturalist and Pl@ntNet.

# Data Loading

```{r config}
# the time window this whole report focuses on
start_year <- 2015
end_year <- 2024

# how many source groups / countries to show in the "top increases" charts
top_n_sources <- 7
top_n_countries <- 15

# which years to show side-by-side in the small map grid later
snapshot_years <- c(2015, 2018, 2021, 2024)
max_points_per_snapshot_year <- 25000
```

```{r read-raw-data}
raw <- read.csv("GlobalGeoTree.csv", stringsAsFactors = FALSE)
```

```{r column-mapping}
# -------------------------------------------------------------------
# IMPORTANT - edit this block if rendering fails with
# "Objekt 'xyz' nicht gefunden" / "object 'xyz' not found".
#
# That error means the column name on the right does not exist in
# your CSV. Run names(raw) in the console, look at the real column
# names, and put the correct ones here. This is the ONLY chunk you
# should need to touch - everything below uses these variables
# instead of typing column names directly, so one fix here is enough
# for the whole report.
# -------------------------------------------------------------------
col_source  <- "source"
col_country <- "country_code"
col_species <- "level3_species"
col_year    <- "year"
col_lat     <- "latitude"
col_lon     <- "longitude"

# quick check so the report fails with a clear message instead of a
# confusing dplyr error, in case a name above is still wrong
needed_cols <- c(col_source, col_country, col_species, col_year)
missing_cols <- setdiff(needed_cols, names(raw))
if (length(missing_cols) > 0) {
  stop(
    "These column names were not found in GlobalGeoTree.csv: ",
    paste(missing_cols, collapse = ", "),
    "\nActual column names in the file are: ",
    paste(names(raw), collapse = ", "),
    "\n--> fix the col_... lines in the column-mapping chunk above."
  )
}

has_coordinates <- col_lat %in% names(raw) && col_lon %in% names(raw)
```

```{r load-data}
df <- raw %>%
  mutate(
    # raw source text, with missing/blank entries labelled explicitly
    source_raw = ifelse(
      is.na(.data[[col_source]]) | str_trim(.data[[col_source]]) == "",
      "Unknown source",
      str_trim(.data[[col_source]])
    ),
    source_lower = str_to_lower(source_raw),

    # group similar sources together (e.g. "iNaturalist.org" -> "iNaturalist")
    # so the charts later don't have hundreds of tiny near-duplicate categories
    source_group = case_when(
      str_detect(source_lower, "inaturalist|inat")                   ~ "iNaturalist",
      str_detect(source_lower, "plantnet|pl@ntnet")                  ~ "Pl@ntNet",
      str_detect(source_lower, "gbif")                                ~ "GBIF",
      str_detect(source_lower, "wikidata|wikipedia")                  ~ "Wikidata/Wikipedia",
      str_detect(source_lower, "herbar|museum|botanic|arboretum")     ~ "Institutional collections",
      str_detect(source_lower, "inventory|census|survey|monitoring")  ~ "Inventory/survey",
      str_detect(source_lower, "unknown|missing")                     ~ "Unknown source",
      TRUE                                                            ~ "Other source"
    ),
    source_group = factor(source_group, levels = source_group_levels),

    country_value = str_to_upper(str_trim(.data[[col_country]])),
    species_value = str_trim(.data[[col_species]]),
    year_value    = as.integer(.data[[col_year]]),
    latitude_value  = if (has_coordinates) suppressWarnings(as.numeric(.data[[col_lat]])) else NA_real_,
    longitude_value = if (has_coordinates) suppressWarnings(as.numeric(.data[[col_lon]])) else NA_real_
  ) %>%
  filter(
    # keep only rows we can actually use: a valid year inside our window,
    # a non-empty country, and a non-empty species
    !is.na(year_value), year_value >= start_year, year_value <= end_year,
    !is.na(country_value), country_value != "",
    !is.na(species_value), species_value != ""
  )
```

```{r data-summary}
# a small overview table so the reader knows how big the cleaned dataset is
data_summary <- tibble(
  Measure = c(
    "Observation file", "Period analysed", "Rows analysed",
    "Distinct sources", "Distinct source groups", "Distinct countries",
    "Distinct species", "Coordinate columns available"
  ),
  Value = c(
    "GlobalGeoTree.csv",
    paste(start_year, end_year, sep = "-"),
    comma(nrow(df)),
    comma(n_distinct(df$source_raw)),
    comma(n_distinct(df$source_group)),
    comma(n_distinct(df$country_value)),
    comma(n_distinct(df$species_value)),
    ifelse(has_coordinates, "Yes", "No")
  )
)

knitr::kable(data_summary, align = "ll")
```

This report uses `r comma(nrow(df))` observations from `r start_year` to `r end_year`.

# Observation Growth Over Time

*Why we look at this:* before asking anything more detailed, we first need to know the basic shape of
the data over time - is the number of observations going up, down, or staying flat?

```{r yearly-counts}
yearly_counts <- df %>%
  count(year_value, name = "observations") %>%
  # right_join against the full year range so that a year with zero
  # observations still shows up as 0, instead of just disappearing
  right_join(tibble(year_value = start_year:end_year), by = "year_value") %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations)) %>%
  arrange(year_value) %>%
  mutate(
    previous_observations = lag(observations),
    absolute_change = observations - previous_observations
  )

growth_summary <- yearly_counts %>% filter(!is.na(absolute_change))

strongest_growth  <- growth_summary %>% slice_max(absolute_change, n = 1, with_ties = FALSE)
strongest_decline <- growth_summary %>% slice_min(absolute_change, n = 1, with_ties = FALSE)

first_year_count <- yearly_counts$observations[yearly_counts$year_value == start_year]
last_year_count  <- yearly_counts$observations[yearly_counts$year_value == end_year]
overall_change   <- (last_year_count - first_year_count) / first_year_count
```

Between `r start_year` and `r end_year`, observations changed from **`r comma(first_year_count)`** to **`r comma(last_year_count)`**, an overall change of **`r percent(overall_change, accuracy = 0.1)`**.

```{r line-observations-by-year, fig.width=9, fig.height=4.8}
# simple line chart: total observations per year.
# this answers subquestion 1 ("has the number of observations increased?")
ggplot(yearly_counts, aes(x = year_value, y = observations)) +
  geom_line(color = "#2A7F62", linewidth = 1.1) +
  geom_point(color = "#2A7F62", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Number of tree observations by year",
    subtitle = paste0("Observation volume in GlobalGeoTree from ", start_year, " to ", end_year, "."),
    x = "Year", y = "Observations"
  )
```

```{r yearly-growth-plot, fig.width=9, fig.height=4.8}
# bar chart of year-over-year change. this is more useful than the line
# chart above for spotting *which* years grew or shrank the most
# (subquestion 2), because small differences are hard to see on a line.
ggplot(growth_summary, aes(x = year_value, y = absolute_change, fill = absolute_change >= 0)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "gray35", linewidth = 0.4) +
  geom_text(
    aes(label = comma(absolute_change)),
    vjust = ifelse(growth_summary$absolute_change >= 0, -0.35, 1.15),
    size = 3.1
  ) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.12, 0.16))) +
  scale_fill_manual(values = c("TRUE" = "#2A7F62", "FALSE" = "#C23B22")) +
  labs(
    title = "Year-to-year change in observations",
    subtitle = "Positive bars indicate growth compared with the previous year.",
    x = "Year", y = "Change in observations",
    caption = paste0(
      "Note: ", end_year, " is the last year in the export and may not be complete yet, so part of\n",
      "the apparent decline could be a data-cutoff effect rather than a real drop in recording."
    )
  )
```

**Interpretation.** The strongest growth occurs in **`r strongest_growth$year_value`**, with **`r comma(strongest_growth$absolute_change)`** more observations than the previous year. The strongest decline occurs in **`r strongest_decline$year_value`**, with **`r comma(strongest_decline$absolute_change)`** fewer observations than the previous year.

Why this caveat matters: if the `r end_year` data was exported partway through the year, the dataset simply hasn't had time to accumulate as many records as a full year would produce. That alone could explain most or all of the apparent "decline" - it doesn't necessarily mean fewer trees are being recorded. This should be re-checked once a complete year of `r end_year` data is available, and it should also be read together with the source composition below, since a platform slowing down its exports would create the exact same pattern.

# Are Temporal Patterns Driven by Specific Sources?

*Why we look at this:* a rise in total observations could either mean "more tree recording is
genuinely happening everywhere", or it could simply mean "one citizen-science app (like iNaturalist)
grew a lot and is dragging the total up with it". The charts below try to tell these two explanations apart.

```{r source-year-data}
# keep the biggest source groups separate, and lump everything else
# together as "Other source" so the legend doesn't get too long
top_source_groups <- df %>%
  count(source_group, sort = TRUE) %>%
  slice_head(n = top_n_sources) %>%
  pull(source_group) %>%
  as.character()

source_year <- df %>%
  mutate(source_group_plot = ifelse(as.character(source_group) %in% top_source_groups,
                                     as.character(source_group), "Other source")) %>%
  count(year_value, source_group_plot, name = "observations") %>%
  right_join(
    expand.grid(year_value = start_year:end_year, source_group_plot = source_group_levels),
    by = c("year_value", "source_group_plot")
  ) %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations)) %>%
  group_by(year_value) %>%
  mutate(share = observations / sum(observations)) %>%   # each source's % of that year's total
  ungroup() %>%
  mutate(source_group_plot = factor(source_group_plot, levels = source_group_levels))
```

```{r stacked-area-source-year, fig.width=9, fig.height=5.2}
# stacked area chart: shows absolute volume per source group over time.
# useful for seeing *which* source is responsible for the overall growth.
ggplot(source_year, aes(x = year_value, y = observations, fill = source_group_plot)) +
  geom_area(alpha = 0.9, linewidth = 0.15, color = "white") +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(
    title = "Observation volume by source group and year",
    subtitle = "A rising source group can drive the apparent growth of the full dataset.",
    x = "Year", y = "Observations", fill = "Source group"
  )
```

```{r source-share-year, fig.width=9, fig.height=5.2}
# 100% stacked bar chart: shows *relative* composition per year, with the
# volume differences removed. This complements the chart above - it answers
# "did the mix of sources change?" separately from "did the total grow?".
ggplot(source_year, aes(x = year_value, y = share, fill = source_group_plot)) +
  geom_col(width = 0.9) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(
    title = "Source composition by year",
    subtitle = "Bars sum to 100% within each year.",
    x = "Year", y = "Share of observations", fill = "Source group",
    caption = "The most recent year may be incomplete depending on the data export date."
  )
```

```{r source-growth-plot-data}
source_growth <- source_year %>%
  arrange(source_group_plot, year_value) %>%
  group_by(source_group_plot) %>%
  mutate(source_change = observations - lag(observations)) %>%
  ungroup()

largest_source_growth <- source_growth %>%
  filter(!is.na(source_change)) %>%
  slice_max(source_change, n = 8, with_ties = FALSE) %>%
  mutate(label = paste0(year_value, ": ", source_group_plot))
```

```{r source-growth-plot, fig.width=9, fig.height=4.8}
# ranks the single biggest year-over-year jumps, split by source group.
# this makes it very concrete which (year, source) combination caused
# the biggest swings seen in the very first growth chart.
ggplot(largest_source_growth, aes(x = reorder(label, source_change), y = source_change, fill = source_group_plot)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(
    title = "Largest source-specific yearly increases",
    subtitle = "This identifies which source groups contributed most to single-year growth.",
    x = NULL, y = "Increase from previous year"
  )
```

**Interpretation.** If most growth is concentrated in one source group, the temporal trend is source-driven. In that case, the line chart should not be interpreted as a neutral increase in global tree recording.

One limitation to flag: this chart ranks *absolute* increases, and the dataset as a whole gets bigger every year, so later years have a built-in advantage simply because there's more data overall to grow from. A version using *percentage* growth per source (instead of raw counts) would separate "this source genuinely exploded" from "this source is just large in a year that was large for everyone" - worth doing as a follow-up if this distinction matters for the conclusions.

# Geographic Coverage Over Time

*Why we look at this:* more observations is not the same as *broader* observations. A dataset
could double in size just by recording the same few countries more intensively. The charts below
check whether new places are actually being added over time (subquestion 3).

```{r country-coverage-data}
country_year <- df %>%
  count(year_value, country_value, name = "observations")

country_year_complete <- expand.grid(
    year_value = start_year:end_year,
    country_value = sort(unique(df$country_value))
  ) %>%
  left_join(country_year, by = c("year_value", "country_value")) %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations))

countries_per_year <- country_year %>%
  group_by(year_value) %>%
  summarise(countries_observed = n_distinct(country_value), .groups = "drop")

# the year a country first shows up anywhere in the data
country_first_year <- country_year %>%
  group_by(country_value) %>%
  summarise(first_observed_year = min(year_value), .groups = "drop")

# running total of "how many countries have we ever seen by year X"
cumulative_countries <- tibble(year_value = start_year:end_year) %>%
  rowwise() %>%
  mutate(cumulative_countries = sum(country_first_year$first_observed_year <= year_value)) %>%
  ungroup()
```

```{r countries-per-year, fig.width=9, fig.height=4.8}
# how many countries had at least one observation in each individual year
# (this can go up AND down - e.g. a country with one observation in 2017
# and none afterwards would only count in 2017)
#
# y-axis is forced to start at 0 on purpose: an axis that starts near 170
# instead of 0 makes normal year-to-year wiggle look like a dramatic swing
ggplot(countries_per_year, aes(x = year_value, y = countries_observed)) +
  geom_line(color = "#4E79A7", linewidth = 1.1) +
  geom_point(color = "#4E79A7", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    title = "Countries represented each year",
    subtitle = "Annual country coverage shows whether observations became geographically broader.",
    x = "Year", y = "Countries with observations"
  )
```

```{r cumulative-country-coverage, fig.width=9, fig.height=4.8}
# unlike the chart above, this one can only ever go up - it counts every
# country that has EVER appeared by a given year, so it shows the
# long-term geographic reach of the dataset rather than year-to-year noise
ggplot(cumulative_countries, aes(x = year_value, y = cumulative_countries)) +
  geom_area(fill = "#8ECAE6", alpha = 0.55) +
  geom_line(color = "#1F5B74", linewidth = 1.1) +
  geom_point(color = "#1F5B74", size = 2.2) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Cumulative country coverage over time",
    subtitle = "The curve rises when a country appears in the analysed period for the first time.",
    x = "Year", y = "Cumulative countries"
  )
```

```{r new-countries-by-year, fig.width=9, fig.height=4.6}
# IMPORTANT: the first year of the window (start_year = 2015) is left out
# of this chart on purpose. Every country with any observation in 2015
# would automatically count as "new" purely because 2015 is where our
# analysis window starts - not because tree recording really began there
# for the first time. Including it would create one huge, misleading bar.
# So this chart only counts countries that are genuinely new from 2016 on.
new_countries <- country_first_year %>%
  filter(first_observed_year > start_year) %>%
  count(first_observed_year, name = "new_countries") %>%
  right_join(tibble(first_observed_year = (start_year + 1):end_year), by = "first_observed_year") %>%
  mutate(new_countries = ifelse(is.na(new_countries), 0L, new_countries)) %>%
  arrange(first_observed_year)

countries_at_start <- sum(country_first_year$first_observed_year == start_year)
```

```{r new-countries-plot, fig.width=9, fig.height=4.6}
ggplot(new_countries, aes(x = first_observed_year, y = new_countries)) +
  geom_col(fill = "#4E79A7", width = 0.72) +
  scale_x_continuous(breaks = (start_year + 1):end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "New countries entering the dataset by year",
    subtitle = "New country entries indicate expansion of geographic coverage within the analysed period.",
    x = "First observed year in period", y = "New countries",
    caption = paste0(
      start_year, " is excluded from this chart: ", countries_at_start,
      " countries already had observations in the very first year of the\n",
      "window, which is not the same as those countries being newly added to the dataset."
    )
  )
```

**Interpretation.** If the number of countries and cumulative country coverage increase over time, GlobalGeoTree is expanding geographically. If observation counts increase without much country expansion, growth may be concentrated in already well-covered countries.

## Which Countries Drive Temporal Growth?

*Why we look at this:* the charts above tell us *how many* countries are covered, but not *which*
countries are responsible for the swings we saw in the very first growth chart. This breaks growth
down by country instead of by source, as a cross-check.

```{r country-growth-data}
country_growth <- country_year_complete %>%
  arrange(country_value, year_value) %>%
  group_by(country_value) %>%
  mutate(country_change = observations - lag(observations)) %>%
  ungroup()

largest_country_growth <- country_growth %>%
  filter(!is.na(country_change)) %>%
  slice_max(country_change, n = top_n_countries, with_ties = FALSE) %>%
  mutate(label = paste0(year_value, ": ", country_value))
```

```{r country-growth-plot, fig.width=9, fig.height=5}
ggplot(largest_country_growth, aes(x = reorder(label, country_change), y = country_change)) +
  geom_col(fill = "#2A7F62", width = 0.72) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Largest country-specific yearly increases",
    subtitle = "Temporal growth may be concentrated in specific countries rather than global.",
    x = NULL, y = "Increase from previous year"
  )
```

# Map Snapshots for Selected Years

*Why we look at this:* country counts treat a huge country and a tiny one the same way. Plotting
the actual coordinates lets us see *where inside* countries the observations are - whether the
dataset is spreading into new regions or just getting denser in the same spots.

```{r map-snapshot-data}
snapshot_years <- snapshot_years[snapshot_years >= start_year & snapshot_years <= end_year]

if (has_coordinates) {
  map_data_points <- df %>%
    filter(
      year_value %in% snapshot_years,
      !is.na(latitude_value), !is.na(longitude_value),
      between(latitude_value, -90, 90), between(longitude_value, -180, 180)
    ) %>%
    group_by(year_value) %>%
    slice_sample(n = max_points_per_snapshot_year) %>%  # cap points per year so the plot stays readable
    ungroup()

  # country outlines are optional - only add them if the maps package is installed
  world_outline <- if (requireNamespace("maps", quietly = TRUE)) ggplot2::map_data("world") else NULL
}
```

```{r map-snapshots, fig.width=9.5, fig.height=6.8}
if (has_coordinates && nrow(map_data_points) > 0) {
  p <- ggplot()

  if (!is.null(world_outline)) {
    p <- p + geom_polygon(
      data = world_outline, aes(x = long, y = lat, group = group),
      fill = "gray96", color = "gray82", linewidth = 0.12, inherit.aes = FALSE
    )
  }

  p +
    geom_point(
      data = map_data_points, aes(x = longitude_value, y = latitude_value),
      alpha = 0.15, size = 0.22, color = "#1F5B74"
    ) +
    facet_wrap(~ year_value, ncol = 2) +
    coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    labs(
      title = "Map snapshots of observations over time",
      subtitle = "Selected years show whether records spread geographically or intensify in existing regions.",
      x = "Longitude", y = "Latitude",
      caption = "Points are sampled per year for readability. Install maps for country outlines."
    )
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable coordinates were detected, so map snapshots are skipped.", size = 4) +
    theme_void()
}
```

```{r grid-coverage-data}
# split the world into 5x5 degree boxes ("grid cells") so we have a
# more fine-grained measure of spatial spread than country borders give us
if (has_coordinates) {
  grid_coverage <- df %>%
    filter(
      !is.na(latitude_value), !is.na(longitude_value),
      between(latitude_value, -90, 90), between(longitude_value, -180, 180)
    ) %>%
    mutate(
      lat_bin = floor(latitude_value / 5) * 5,
      lon_bin = floor(longitude_value / 5) * 5,
      grid_cell = paste(lat_bin, lon_bin, sep = "_")
    ) %>%
    distinct(year_value, grid_cell)

  grid_first_year <- grid_coverage %>%
    group_by(grid_cell) %>%
    summarise(first_observed_year = min(year_value), .groups = "drop")

  cumulative_grid_cells <- tibble(year_value = start_year:end_year) %>%
    rowwise() %>%
    mutate(cumulative_grid_cells = sum(grid_first_year$first_observed_year <= year_value)) %>%
    ungroup()
}
```

```{r cumulative-grid-coverage, fig.width=9, fig.height=4.8}
# same cumulative logic as the country chart earlier, but using the finer
# 5x5 degree grid instead of whole countries - this can keep rising even
# after country coverage has flattened out, if growth is happening *within*
# already-covered countries (e.g. new regions inside a large country)
if (has_coordinates && nrow(cumulative_grid_cells) > 0) {
  ggplot(cumulative_grid_cells, aes(x = year_value, y = cumulative_grid_cells)) +
    geom_area(fill = "#B2DF8A", alpha = 0.65) +
    geom_line(color = "#2A7F62", linewidth = 1.1) +
    geom_point(color = "#2A7F62", size = 2.2) +
    scale_x_continuous(breaks = start_year:end_year) +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Cumulative spatial grid coverage over time",
      subtitle = "A 5-degree latitude/longitude grid approximates whether spatial coverage expands.",
      x = "Year", y = "Cumulative occupied grid cells"
    )
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable coordinates were detected.", size = 4) +
    theme_void()
}
```

**Interpretation.** Map snapshots and grid coverage distinguish two different processes: more observations in the same places versus expansion into new regions.

# Overall Interpretation

From 2015 to 2024, temporal development in GlobalGeoTree should be interpreted as a combination of:

- changing observation volume,
- expansion or concentration of geographic coverage,
- country-specific growth,
- and source/platform growth.

If the strongest year-to-year increases are driven by citizen-science platforms or a small number of countries, the temporal trend is not purely biological. It is partly a record-production trend. Therefore, temporal analyses should report both total observations and source composition by year.

Two limitations worth keeping in mind before treating the headline numbers as final:

- **`r end_year` is likely an incomplete year** in this export, so the apparent decline at the end of the series may shrink or disappear once a full year of data is available.
- **The "new countries" chart excludes `r start_year`** because countries present at the very start of the analysis window are not the same as countries that newly entered the dataset; counting them as new would overstate geographic expansion.
