---
title: "RQ3 - Temporal Development in GlobalGeoTree"
subtitle: "Changes in tree observation volume, source composition, and geographic coverage from 2015 to 2024"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 3
    number-sections: false
    documentclass: article
    pdf-engine: xelatex
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Required packages:
# install.packages(c("data.table", "dplyr", "stringr", "ggplot2", "scales"))
#
# Optional for map outlines:
# install.packages("maps")

library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)

theme_gtree <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(color = "gray35"),
      plot.caption = element_text(color = "gray45", size = base_size - 2),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      strip.background = element_rect(fill = "gray94", color = NA),
      plot.margin = margin(8, 14, 8, 8)
    )
}

theme_set(theme_gtree())

source_palette <- c(
  "iNaturalist" = "#8ECAE6",
  "Pl@ntNet" = "#FF9DA7",
  "GBIF" = "#4E79A7",
  "Wikidata/Wikipedia" = "#B07AA1",
  "Institutional collections" = "#1F77B4",
  "Inventory/survey" = "#B2DF8A",
  "Other source" = "#2CA02C",
  "Unknown source" = "#E41A1C"
)

source_group_levels <- c(
  "iNaturalist",
  "Pl@ntNet",
  "GBIF",
  "Wikidata/Wikipedia",
  "Institutional collections",
  "Inventory/survey",
  "Other source",
  "Unknown source"
)

source_fill_scale <- function(..., na.value = "gray75") {
  scale_fill_manual(values = source_palette, drop = TRUE, na.value = na.value, ...)
}
```

# Research Question

**RQ3 - Temporal Development**  
How has the number and distribution of tree observations changed from **2015 to 2024**?

This report answers four subquestions:

1. Has the number of observations increased over time?
2. Which years show the strongest growth or decline?
3. Did geographic coverage expand over time?
4. Are temporal patterns mainly driven by specific sources?

The key idea is that time trends in GlobalGeoTree may reflect both real changes in observation coverage and the growth of data platforms such as iNaturalist and Pl@ntNet.

# Data Loading

```{r user-config}
# -------------------------------------------------------------------------
# User configuration
# -------------------------------------------------------------------------

# Put this .qmd in the same folder as the full dataset.
# The full dataset should be called GlobalGeoTree, without "_full".
observations_path <- "GlobalGeoTree.csv"

start_year <- 2015
end_year <- 2024

# For faster draft rendering, set use_subsample <- TRUE.
# Use the full data for final counts.
use_subsample <- FALSE
sample_n <- 500000
set.seed(2026)

top_n_sources <- 7
top_n_countries <- 15

# Years shown in map snapshots. These should stay inside start_year:end_year.
snapshot_years <- c(2015, 2018, 2021, 2024)

# Maximum plotted points per snapshot year.
max_points_per_snapshot_year <- 25000
```

```{r helper-functions}
normalize_names <- function(x) {
  x |>
    str_replace_all("([a-z0-9])([A-Z])", "\\1_\\2") |>
    str_replace_all("[^A-Za-z0-9]+", "_") |>
    str_replace_all("_+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()
}

first_existing <- function(data, candidates, label, required = TRUE) {
  found <- candidates[candidates %in% names(data)]

  if (length(found) > 0) {
    return(found[[1]])
  }

  if (required) {
    stop(
      paste0(
        "Could not find a column for ", label, ". Tried: ",
        paste(candidates, collapse = ", "),
        "\nAvailable columns are:\n",
        paste(names(data), collapse = ", ")
      ),
      call. = FALSE
    )
  }

  NA_character_
}

clean_source_label <- function(x) {
  x <- as.character(x)
  x <- ifelse(is.na(x) | str_trim(x) == "", "Unknown source", x)
  str_squish(x)
}

source_group_label <- function(x) {
  source_clean <- str_to_lower(clean_source_label(x))

  case_when(
    str_detect(source_clean, "inaturalist|i-naturalist|inat") ~ "iNaturalist",
    str_detect(source_clean, "plantnet|pl@ntnet") ~ "Pl@ntNet",
    str_detect(source_clean, "gbif") ~ "GBIF",
    str_detect(source_clean, "wikidata|wikipedia") ~ "Wikidata/Wikipedia",
    str_detect(source_clean, "herbar|museum|botanic|arboretum") ~ "Institutional collections",
    str_detect(source_clean, "inventory|census|survey|monitoring") ~ "Inventory/survey",
    str_detect(source_clean, "unknown|missing|na") ~ "Unknown source",
    TRUE ~ "Other source"
  )
}

sample_up_to_grouped <- function(data, n_per_group) {
  data |>
    group_modify(~ slice_sample(.x, n = min(nrow(.x), n_per_group))) |>
    ungroup()
}
```

```{r load-data}
if (!file.exists(observations_path)) {
  stop(
    paste0(
      "Could not find ", observations_path, ". Put this QMD in the same folder ",
      "as GlobalGeoTree.csv, or edit observations_path in the configuration chunk."
    ),
    call. = FALSE
  )
}

GlobalGeoTree <- fread(observations_path, showProgress = FALSE)
names(GlobalGeoTree) <- normalize_names(names(GlobalGeoTree))

if (use_subsample && nrow(GlobalGeoTree) > sample_n) {
  GlobalGeoTree <- GlobalGeoTree |>
    slice_sample(n = sample_n)
}

source_col <- first_existing(
  GlobalGeoTree,
  c(
    "source", "sources", "data_source", "dataset_source", "platform",
    "publisher", "publisher_title", "institution_code", "dataset_name",
    "basis_of_record", "recorded_by"
  ),
  "observation source"
)

country_col <- first_existing(
  GlobalGeoTree,
  c("country_code", "countrycode", "iso_a2", "iso2", "country", "level0_country", "country_name"),
  "country"
)

species_col <- first_existing(
  GlobalGeoTree,
  c("level3_species", "species", "scientific_name", "scientificname", "accepted_scientific_name", "taxon_name"),
  "species"
)

year_col <- first_existing(
  GlobalGeoTree,
  c("year", "observation_year", "event_year", "recorded_year"),
  "year",
  required = FALSE
)

date_col <- first_existing(
  GlobalGeoTree,
  c("event_date", "eventdate", "date", "observed_on", "observation_date", "recorded_date"),
  "date",
  required = FALSE
)

lat_col <- first_existing(
  GlobalGeoTree,
  c("latitude", "decimal_latitude", "decimallatitude", "lat", "y"),
  "latitude",
  required = FALSE
)

lon_col <- first_existing(
  GlobalGeoTree,
  c("longitude", "decimal_longitude", "decimallongitude", "lon", "lng", "x"),
  "longitude",
  required = FALSE
)

if (is.na(year_col) && !is.na(date_col)) {
  GlobalGeoTree <- GlobalGeoTree |>
    mutate(year_from_date = suppressWarnings(as.integer(str_sub(as.character(.data[[date_col]]), 1, 4))))
  year_col <- "year_from_date"
}

if (is.na(year_col)) {
  stop(
    "No usable year or date column was found. RQ3 needs a year column or a date column.",
    call. = FALSE
  )
}

GlobalGeoTree <- GlobalGeoTree |>
  mutate(
    source_raw = clean_source_label(.data[[source_col]]),
    source_group = factor(source_group_label(source_raw), levels = source_group_levels),
    country_value = str_to_upper(str_squish(as.character(.data[[country_col]]))),
    species_value = str_squish(as.character(.data[[species_col]])),
    year_value = suppressWarnings(as.integer(.data[[year_col]])),
    latitude_value = if (!is.na(lat_col)) suppressWarnings(as.numeric(.data[[lat_col]])) else NA_real_,
    longitude_value = if (!is.na(lon_col)) suppressWarnings(as.numeric(.data[[lon_col]])) else NA_real_
  ) |>
  filter(
    !is.na(year_value),
    year_value >= start_year,
    year_value <= end_year,
    !is.na(source_raw),
    !is.na(country_value),
    !is.na(species_value)
  )

if (nrow(GlobalGeoTree) == 0) {
  stop(
    paste0("No observations were found between ", start_year, " and ", end_year, "."),
    call. = FALSE
  )
}

has_coordinates <- !is.na(lat_col) && !is.na(lon_col)

report_note <- if (use_subsample) {
  paste0("This report uses a random subsample of ", comma(nrow(GlobalGeoTree)), " observations from ", start_year, " to ", end_year, ".")
} else {
  paste0("This report uses ", comma(nrow(GlobalGeoTree)), " observations from ", start_year, " to ", end_year, ".")
}
```

```{r data-summary}
data_summary <- tibble(
  Measure = c(
    "Observation file",
    "Period analysed",
    "Rows analysed",
    "Distinct sources",
    "Distinct source groups",
    "Distinct countries",
    "Distinct species",
    "Coordinate columns available"
  ),
  Value = c(
    observations_path,
    paste(start_year, end_year, sep = "-"),
    comma(nrow(GlobalGeoTree)),
    comma(n_distinct(GlobalGeoTree$source_raw)),
    comma(n_distinct(GlobalGeoTree$source_group)),
    comma(n_distinct(GlobalGeoTree$country_value)),
    comma(n_distinct(GlobalGeoTree$species_value)),
    ifelse(has_coordinates, "Yes", "No")
  )
)

knitr::kable(data_summary, align = "ll")
```

`r report_note`

# Observation Growth Over Time

```{r yearly-counts}
yearly_counts <- GlobalGeoTree |>
  count(year_value, name = "observations") |>
  right_join(tibble(year_value = start_year:end_year), by = "year_value") |>
  mutate(observations = if_else(is.na(observations), 0L, observations)) |>
  arrange(year_value) |>
  mutate(
    previous_observations = lag(observations),
    absolute_change = observations - previous_observations,
    relative_change = if_else(previous_observations > 0, absolute_change / previous_observations, NA_real_)
  )

growth_summary <- yearly_counts |>
  filter(!is.na(absolute_change))

strongest_growth <- growth_summary |>
  slice_max(absolute_change, n = 1, with_ties = FALSE)

strongest_decline <- growth_summary |>
  slice_min(absolute_change, n = 1, with_ties = FALSE)

first_year_count <- yearly_counts |>
  filter(year_value == min(year_value)) |>
  pull(observations)

last_year_count <- yearly_counts |>
  filter(year_value == max(year_value)) |>
  pull(observations)

overall_change <- if (first_year_count > 0) {
  (last_year_count - first_year_count) / first_year_count
} else {
  NA_real_
}

overall_change_text <- if (!is.na(overall_change)) {
  percent(overall_change, accuracy = 0.1)
} else {
  "not defined because the first year has zero observations"
}
```

Between `r min(yearly_counts$year_value)` and `r max(yearly_counts$year_value)`, observations changed from **`r comma(first_year_count)`** to **`r comma(last_year_count)`**, an overall change of **`r overall_change_text`**.

```{r line-observations-by-year, fig.width=9, fig.height=4.8}
ggplot(yearly_counts, aes(x = year_value, y = observations)) +
  geom_line(color = "#2A7F62", linewidth = 1.1) +
  geom_point(color = "#2A7F62", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Number of tree observations by year",
    subtitle = paste0("Observation volume in GlobalGeoTree from ", start_year, " to ", end_year, "."),
    x = "Year",
    y = "Observations"
  )
```

```{r yearly-growth-plot, fig.width=9, fig.height=4.8}
ggplot(growth_summary, aes(x = year_value, y = absolute_change, fill = absolute_change >= 0)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "gray35", linewidth = 0.4) +
  geom_text(
    aes(label = comma(absolute_change)),
    vjust = if_else(growth_summary$absolute_change >= 0, -0.35, 1.15),
    size = 3.1
  ) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.12, 0.14))) +
  scale_fill_manual(values = c("TRUE" = "#2A7F62", "FALSE" = "#C23B22")) +
  labs(
    title = "Year-to-year change in observations",
    subtitle = "Positive bars indicate growth compared with the previous year.",
    x = "Year",
    y = "Change in observations"
  )
```

**Interpretation.** The strongest growth occurs in **`r strongest_growth$year_value`**, with **`r comma(strongest_growth$absolute_change)`** more observations than the previous year. The strongest decline occurs in **`r strongest_decline$year_value`**, with **`r comma(strongest_decline$absolute_change)`** fewer observations than the previous year. These changes should be interpreted together with source composition, because platform growth can create strong temporal trends.

# Are Temporal Patterns Driven by Specific Sources?

```{r source-year-data}
source_year_counts <- GlobalGeoTree |>
  count(year_value, source_group, name = "observations") |>
  right_join(
    expand.grid(
      year_value = start_year:end_year,
      source_group = factor(source_group_levels, levels = source_group_levels)
    ) |>
      as_tibble(),
    by = c("year_value", "source_group")
  ) |>
  mutate(observations = if_else(is.na(observations), 0L, observations)) |>
  group_by(year_value) |>
  mutate(
    year_total = sum(observations),
    share = if_else(year_total > 0, observations / year_total, 0)
  ) |>
  ungroup()

top_source_groups <- GlobalGeoTree |>
  count(source_group, sort = TRUE, name = "observations") |>
  slice_head(n = top_n_sources) |>
  pull(source_group) |>
  as.character()

source_year_plot <- GlobalGeoTree |>
  mutate(source_group_plot = if_else(as.character(source_group) %in% top_source_groups, as.character(source_group), "Other source")) |>
  count(year_value, source_group_plot, name = "observations") |>
  right_join(
    expand.grid(
      year_value = start_year:end_year,
      source_group_plot = source_group_levels
    ) |>
      as_tibble(),
    by = c("year_value", "source_group_plot")
  ) |>
  mutate(observations = if_else(is.na(observations), 0L, observations)) |>
  group_by(year_value) |>
  mutate(
    year_total = sum(observations),
    share = if_else(year_total > 0, observations / year_total, 0)
  ) |>
  ungroup() |>
  mutate(source_group_plot = factor(source_group_plot, levels = source_group_levels))
```

```{r stacked-area-source-year, fig.width=9, fig.height=5.2}
ggplot(source_year_plot, aes(x = year_value, y = observations, fill = source_group_plot)) +
  geom_area(alpha = 0.9, linewidth = 0.15, color = "white") +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  source_fill_scale() +
  labs(
    title = "Observation volume by source group and year",
    subtitle = "A rising source group can drive the apparent growth of the full dataset.",
    x = "Year",
    y = "Observations",
    fill = "Source group"
  )
```

```{r source-share-year, fig.width=9, fig.height=5.2}
ggplot(source_year_plot, aes(x = year_value, y = share, fill = source_group_plot)) +
  geom_col(width = 0.9) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = percent) +
  source_fill_scale() +
  labs(
    title = "Source composition by year",
    subtitle = "Bars sum to 100% within each year.",
    x = "Year",
    y = "Share of observations",
    fill = "Source group",
    caption = "The most recent year may be incomplete depending on the data export date."
  )
```

```{r source-growth-contribution}
source_growth <- source_year_plot |>
  arrange(source_group_plot, year_value) |>
  group_by(source_group_plot) |>
  mutate(previous_observations = lag(observations), source_change = observations - previous_observations) |>
  ungroup()

largest_source_growth <- source_growth |>
  filter(!is.na(source_change)) |>
  slice_max(source_change, n = 8, with_ties = FALSE) |>
  mutate(label = paste0(year_value, ": ", source_group_plot))
```

```{r source-growth-plot, fig.width=9, fig.height=4.8}
ggplot(largest_source_growth, aes(x = reorder(label, source_change), y = source_change, fill = source_group_plot)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  source_fill_scale() +
  labs(
    title = "Largest source-specific yearly increases",
    subtitle = "This identifies which source groups contributed most to single-year growth.",
    x = NULL,
    y = "Increase from previous year"
  )
```

**Interpretation.** If most growth is concentrated in one source group, the temporal trend is source-driven. In that case, the line chart should not be interpreted as a neutral increase in global tree recording.

# Geographic Coverage Over Time

```{r country-coverage-data}
country_year <- GlobalGeoTree |>
  count(year_value, country_value, name = "observations")

country_year_complete <- country_year |>
  right_join(
    expand.grid(
      year_value = start_year:end_year,
      country_value = sort(unique(GlobalGeoTree$country_value))
    ) |>
      as_tibble(),
    by = c("year_value", "country_value")
  ) |>
  mutate(observations = if_else(is.na(observations), 0L, observations))

countries_per_year <- country_year |>
  group_by(year_value) |>
  summarise(
    countries_observed = n_distinct(country_value),
    observations = sum(observations),
    .groups = "drop"
  )

country_first_year <- country_year |>
  group_by(country_value) |>
  summarise(first_observed_year = min(year_value), .groups = "drop")

cumulative_countries <- tibble(year_value = start_year:end_year) |>
  rowwise() |>
  mutate(cumulative_countries = sum(country_first_year$first_observed_year <= year_value)) |>
  ungroup()

new_countries <- country_first_year |>
  count(first_observed_year, name = "new_countries") |>
  right_join(tibble(first_observed_year = start_year:end_year), by = "first_observed_year") |>
  mutate(new_countries = if_else(is.na(new_countries), 0L, new_countries)) |>
  arrange(first_observed_year)
```

```{r countries-per-year, fig.width=9, fig.height=4.8}
ggplot(countries_per_year, aes(x = year_value, y = countries_observed)) +
  geom_line(color = "#4E79A7", linewidth = 1.1) +
  geom_point(color = "#4E79A7", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Countries represented each year",
    subtitle = "Annual country coverage shows whether observations became geographically broader.",
    x = "Year",
    y = "Countries with observations"
  )
```

```{r cumulative-country-coverage, fig.width=9, fig.height=4.8}
ggplot(cumulative_countries, aes(x = year_value, y = cumulative_countries)) +
  geom_area(fill = "#8ECAE6", alpha = 0.55) +
  geom_line(color = "#1F5B74", linewidth = 1.1) +
  geom_point(color = "#1F5B74", size = 2.2) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Cumulative country coverage over time",
    subtitle = "The curve rises when a country appears in the analysed period for the first time.",
    x = "Year",
    y = "Cumulative countries"
  )
```

```{r new-countries-by-year, fig.width=9, fig.height=4.6}
ggplot(new_countries, aes(x = first_observed_year, y = new_countries)) +
  geom_col(fill = "#4E79A7", width = 0.72) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "New countries entering the dataset by year",
    subtitle = "New country entries indicate expansion of geographic coverage within the analysed period.",
    x = "First observed year in period",
    y = "New countries"
  )
```

**Interpretation.** If the number of countries and cumulative country coverage increase over time, GlobalGeoTree is expanding geographically. If observation counts increase without much country expansion, growth may be concentrated in already well-covered countries.

## Which Countries Drive Temporal Growth?

```{r country-growth-data}
country_growth <- country_year |>
  right_join(country_year_complete, by = c("year_value", "country_value"), suffix = c("_observed", "")) |>
  select(year_value, country_value, observations) |>
  arrange(country_value, year_value) |>
  group_by(country_value) |>
  mutate(previous_observations = lag(observations), country_change = observations - previous_observations) |>
  ungroup()

largest_country_growth <- country_growth |>
  filter(!is.na(country_change)) |>
  slice_max(country_change, n = top_n_countries, with_ties = FALSE) |>
  mutate(label = paste0(year_value, ": ", country_value))
```

```{r country-growth-plot, fig.width=9, fig.height=5}
ggplot(largest_country_growth, aes(x = reorder(label, country_change), y = country_change)) +
  geom_col(fill = "#2A7F62", width = 0.72) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Largest country-specific yearly increases",
    subtitle = "Temporal growth may be concentrated in specific countries rather than global.",
    x = NULL,
    y = "Increase from previous year"
  )
```

# Map Snapshots for Selected Years

```{r map-snapshot-data}
snapshot_years <- snapshot_years[snapshot_years >= start_year & snapshot_years <= end_year]

if (has_coordinates && length(snapshot_years) > 0) {
  map_snapshot_data <- GlobalGeoTree |>
    filter(
      year_value %in% snapshot_years,
      !is.na(latitude_value),
      !is.na(longitude_value),
      between(latitude_value, -90, 90),
      between(longitude_value, -180, 180)
    ) |>
    group_by(year_value) |>
    sample_up_to_grouped(max_points_per_snapshot_year)

  world_outline <- NULL
  if (requireNamespace("maps", quietly = TRUE)) {
    world_outline <- ggplot2::map_data("world")
  }
}
```

```{r map-snapshots, fig.width=9.5, fig.height=6.8}
if (has_coordinates && exists("map_snapshot_data") && nrow(map_snapshot_data) > 0) {
  p <- ggplot()

  if (exists("world_outline") && !is.null(world_outline)) {
    p <- p +
      geom_polygon(
        data = world_outline,
        aes(x = long, y = lat, group = group),
        fill = "gray96",
        color = "gray82",
        linewidth = 0.12,
        inherit.aes = FALSE
      )
  }

  p +
    geom_point(
      data = map_snapshot_data,
      aes(x = longitude_value, y = latitude_value),
      alpha = 0.15,
      size = 0.22,
      color = "#1F5B74"
    ) +
    facet_wrap(~ year_value, ncol = 2) +
    coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    labs(
      title = "Map snapshots of observations over time",
      subtitle = "Selected years show whether records spread geographically or intensify in existing regions.",
      x = "Longitude",
      y = "Latitude",
      caption = "Points are sampled per year for readability. Install maps for country outlines."
    )
} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "No usable coordinates were detected, so map snapshots are skipped.",
      size = 4
    ) +
    theme_void()
}
```

```{r grid-coverage-data}
if (has_coordinates) {
  grid_coverage <- GlobalGeoTree |>
    filter(
      !is.na(latitude_value),
      !is.na(longitude_value),
      between(latitude_value, -90, 90),
      between(longitude_value, -180, 180)
    ) |>
    mutate(
      lat_bin = floor(latitude_value / 5) * 5,
      lon_bin = floor(longitude_value / 5) * 5,
      grid_cell = paste(lat_bin, lon_bin, sep = "_")
    ) |>
    distinct(year_value, grid_cell)

  grid_first_year <- grid_coverage |>
    group_by(grid_cell) |>
    summarise(first_observed_year = min(year_value), .groups = "drop")

  cumulative_grid_cells <- tibble(year_value = start_year:end_year) |>
    rowwise() |>
    mutate(cumulative_grid_cells = sum(grid_first_year$first_observed_year <= year_value)) |>
    ungroup()
}
```

```{r cumulative-grid-coverage, fig.width=9, fig.height=4.8}
if (has_coordinates && exists("cumulative_grid_cells") && nrow(cumulative_grid_cells) > 0) {
  ggplot(cumulative_grid_cells, aes(x = year_value, y = cumulative_grid_cells)) +
    geom_area(fill = "#B2DF8A", alpha = 0.65) +
    geom_line(color = "#2A7F62", linewidth = 1.1) +
    geom_point(color = "#2A7F62", size = 2.2) +
    scale_x_continuous(breaks = start_year:end_year) +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Cumulative spatial grid coverage over time",
      subtitle = "A 5-degree latitude/longitude grid approximates whether spatial coverage expands.",
      x = "Year",
      y = "Cumulative occupied grid cells"
    )
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable coordinates were detected.", size = 4) +
    theme_void()
}
```

**Interpretation.** Map snapshots and grid coverage distinguish two different processes: more observations in the same places versus expansion into new regions.

# Overall Interpretation

From 2015 to 2024, temporal development in GlobalGeoTree should be interpreted as a combination of:

- changing observation volume,
- expansion or concentration of geographic coverage,
- country-specific growth,
- and source/platform growth.

If the strongest year-to-year increases are driven by citizen-science platforms or a small number of countries, the temporal trend is not purely biological. It is partly a record-production trend. Therefore, temporal analyses should report both total observations and source composition by year.
