---
title: "RQ3 - Temporal Development in GlobalGeoTree"
subtitle: "Changes in tree observation volume, source composition, and geographic coverage from 2015 to 2024"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "gray35"),
      plot.caption = element_text(color = "gray45", size = 9),
      legend.position = "bottom"
    )
)

# colors per source group
# (GBIF and Institutional collections used to be two similar blues -
# changed so they don't get confused with iNaturalist in the stacked charts)
source_colors <- c(
  "iNaturalist"                = "#8ECAE6",
  "Pl@ntNet"                   = "#FF9DA7",
  "GBIF"                       = "#264653",
  "Wikidata/Wikipedia"         = "#B07AA1",
  "Institutional collections"  = "#9C6644",
  "Inventory/survey"           = "#B2DF8A",
  "Other source"                = "#2CA02C",
  "Unknown source"             = "#E41A1C"
)

source_group_levels <- names(source_colors)
```

# Research Question

**RQ3 - Temporal Development**  
How has the number and distribution of tree observations changed from **2015 to 2024**?

This report answers four subquestions:

1. Has the number of observations increased over time?
2. Which years show the strongest growth or decline?
3. Did geographic coverage expand over time?
4. Are temporal patterns mainly driven by specific sources?

The key idea is that time trends in GlobalGeoTree may reflect both real changes in observation coverage and the growth of data platforms such as iNaturalist and Pl@ntNet.

# Data Loading

```{r config}
# settings for this report
start_year <- 2015
end_year <- 2024

top_n_sources <- 7
top_n_countries <- 15

# years used for the small-multiple maps
snapshot_years <- c(2015, 2018, 2021, 2024)
max_points_per_snapshot_year <- 25000
```

```{r load-data}
# I checked names(read.csv("GlobalGeoTree.csv")) once and the columns I need
# are called source, country, species, year, latitude, longitude.
raw <- read.csv("GlobalGeoTree.csv", stringsAsFactors = FALSE)

df <- raw %>%
  mutate(
    source_raw = ifelse(is.na(source) | str_trim(source) == "", "Unknown source", str_trim(source)),
    country_value = str_to_upper(str_trim(country)),
    species_value = str_trim(species),
    year_value = as.integer(year),
    latitude_value = suppressWarnings(as.numeric(latitude)),
    longitude_value = suppressWarnings(as.numeric(longitude))
  ) %>%
  filter(
    !is.na(year_value), year_value >= start_year, year_value <= end_year,
    !is.na(country_value), country_value != "",
    !is.na(species_value), species_value != ""
  )

has_coordinates <- "latitude" %in% names(raw) & "longitude" %in% names(raw)
```

```{r data-summary}
data_summary <- tibble(
  Measure = c(
    "Observation file", "Period analysed", "Rows analysed",
    "Distinct sources", "Distinct source groups", "Distinct countries",
    "Distinct species", "Coordinate columns available"
  ),
  Value = c(
    "GlobalGeoTree.csv",
    paste(start_year, end_year, sep = "-"),
    comma(nrow(df)),
    comma(n_distinct(df$source_raw)),
    comma(n_distinct(df$source_group)),
    comma(n_distinct(df$country_value)),
    comma(n_distinct(df$species_value)),
    ifelse(has_coordinates, "Yes", "No")
  )
)

knitr::kable(data_summary, align = "ll")
```

This report uses `r comma(nrow(df))` observations from `r start_year` to `r end_year`.

# Observation Growth Over Time

```{r yearly-counts}
yearly_counts <- df %>%
  count(year_value, name = "observations") %>%
  right_join(tibble(year_value = start_year:end_year), by = "year_value") %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations)) %>%
  arrange(year_value) %>%
  mutate(
    previous_observations = lag(observations),
    absolute_change = observations - previous_observations
  )

growth_summary <- yearly_counts %>% filter(!is.na(absolute_change))

strongest_growth  <- growth_summary %>% slice_max(absolute_change, n = 1, with_ties = FALSE)
strongest_decline <- growth_summary %>% slice_min(absolute_change, n = 1, with_ties = FALSE)

first_year_count <- yearly_counts$observations[yearly_counts$year_value == start_year]
last_year_count  <- yearly_counts$observations[yearly_counts$year_value == end_year]
overall_change   <- (last_year_count - first_year_count) / first_year_count
```

Between `r start_year` and `r end_year`, observations changed from **`r comma(first_year_count)`** to **`r comma(last_year_count)`**, an overall change of **`r percent(overall_change, accuracy = 0.1)`**.

```{r line-observations-by-year, fig.width=9, fig.height=4.8}
ggplot(yearly_counts, aes(x = year_value, y = observations)) +
  geom_line(color = "#2A7F62", linewidth = 1.1) +
  geom_point(color = "#2A7F62", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Number of tree observations by year",
    subtitle = paste0("Observation volume in GlobalGeoTree from ", start_year, " to ", end_year, "."),
    x = "Year", y = "Observations"
  )
```

```{r yearly-growth-plot, fig.width=9, fig.height=4.8}
ggplot(growth_summary, aes(x = year_value, y = absolute_change, fill = absolute_change >= 0)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "gray35", linewidth = 0.4) +
  geom_text(
    aes(label = comma(absolute_change)),
    vjust = ifelse(growth_summary$absolute_change >= 0, -0.35, 1.15),
    size = 3.1
  ) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.12, 0.16))) +
  scale_fill_manual(values = c("TRUE" = "#2A7F62", "FALSE" = "#C23B22")) +
  labs(
    title = "Year-to-year change in observations",
    subtitle = "Positive bars indicate growth compared with the previous year.",
    x = "Year", y = "Change in observations",
    caption = "Note: 2024 is the last year in the export and may not be complete yet, so part of\nthe apparent decline could be a data-cutoff effect rather than a real drop in recording."
  )
```

**Interpretation.** The strongest growth occurs in **`r strongest_growth$year_value`**, with **`r comma(strongest_growth$absolute_change)`** more observations than the previous year. The strongest decline occurs in **`r strongest_decline$year_value`**, with **`r comma(strongest_decline$absolute_change)`** fewer observations than the previous year. Because the 2024 data may have been exported before the year was finished, this "decline" should not automatically be read as a real drop in tree recording &mdash; it needs to be checked again once a full year of 2024 data is available. These changes should also be interpreted together with source composition, because platform growth can create strong temporal trends on its own.

# Are Temporal Patterns Driven by Specific Sources?

```{r source-year-data}
top_source_groups <- df %>%
  count(source_group, sort = TRUE) %>%
  slice_head(n = top_n_sources) %>%
  pull(source_group) %>%
  as.character()

source_year <- df %>%
  mutate(source_group_plot = ifelse(as.character(source_group) %in% top_source_groups,
                                     as.character(source_group), "Other source")) %>%
  count(year_value, source_group_plot, name = "observations") %>%
  right_join(
    expand.grid(year_value = start_year:end_year, source_group_plot = source_group_levels),
    by = c("year_value", "source_group_plot")
  ) %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations)) %>%
  group_by(year_value) %>%
  mutate(share = observations / sum(observations)) %>%
  ungroup() %>%
  mutate(source_group_plot = factor(source_group_plot, levels = source_group_levels))
```

```{r stacked-area-source-year, fig.width=9, fig.height=5.2}
ggplot(source_year, aes(x = year_value, y = observations, fill = source_group_plot)) +
  geom_area(alpha = 0.9, linewidth = 0.15, color = "white") +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(
    title = "Observation volume by source group and year",
    subtitle = "A rising source group can drive the apparent growth of the full dataset.",
    x = "Year", y = "Observations", fill = "Source group"
  )
```

```{r source-share-year, fig.width=9, fig.height=5.2}
ggplot(source_year, aes(x = year_value, y = share, fill = source_group_plot)) +
  geom_col(width = 0.9) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(
    title = "Source composition by year",
    subtitle = "Bars sum to 100% within each year.",
    x = "Year", y = "Share of observations", fill = "Source group",
    caption = "The most recent year may be incomplete depending on the data export date."
  )
```

```{r source-growth-plot-data}
source_growth <- source_year %>%
  arrange(source_group_plot, year_value) %>%
  group_by(source_group_plot) %>%
  mutate(source_change = observations - lag(observations)) %>%
  ungroup()

largest_source_growth <- source_growth %>%
  filter(!is.na(source_change)) %>%
  slice_max(source_change, n = 8, with_ties = FALSE) %>%
  mutate(label = paste0(year_value, ": ", source_group_plot))
```

```{r source-growth-plot, fig.width=9, fig.height=4.8}
ggplot(largest_source_growth, aes(x = reorder(label, source_change), y = source_change, fill = source_group_plot)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = source_colors, drop = TRUE) +
  labs(
    title = "Largest source-specific yearly increases",
    subtitle = "This identifies which source groups contributed most to single-year growth.",
    x = NULL, y = "Increase from previous year"
  )
```

**Interpretation.** If most growth is concentrated in one source group, the temporal trend is source-driven. In that case, the line chart should not be interpreted as a neutral increase in global tree recording. Note also that this chart ranks *absolute* increases, so later years are more likely to appear simply because the dataset as a whole is bigger by then &mdash; a version using *percentage* growth per source would be a useful follow-up to separate "this source grew a lot" from "this source is large in a large year."

# Geographic Coverage Over Time

```{r country-coverage-data}
country_year <- df %>%
  count(year_value, country_value, name = "observations")

country_year_complete <- expand.grid(
    year_value = start_year:end_year,
    country_value = sort(unique(df$country_value))
  ) %>%
  left_join(country_year, by = c("year_value", "country_value")) %>%
  mutate(observations = ifelse(is.na(observations), 0L, observations))

countries_per_year <- country_year %>%
  group_by(year_value) %>%
  summarise(countries_observed = n_distinct(country_value), .groups = "drop")

country_first_year <- country_year %>%
  group_by(country_value) %>%
  summarise(first_observed_year = min(year_value), .groups = "drop")

cumulative_countries <- tibble(year_value = start_year:end_year) %>%
  rowwise() %>%
  mutate(cumulative_countries = sum(country_first_year$first_observed_year <= year_value)) %>%
  ungroup()
```

```{r countries-per-year, fig.width=9, fig.height=4.8}
# y-axis starts at 0 here on purpose: with only the top part of the range
# shown, the year-to-year wiggle looks much bigger than it actually is
ggplot(countries_per_year, aes(x = year_value, y = countries_observed)) +
  geom_line(color = "#4E79A7", linewidth = 1.1) +
  geom_point(color = "#4E79A7", size = 2.4) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    title = "Countries represented each year",
    subtitle = "Annual country coverage shows whether observations became geographically broader.",
    x = "Year", y = "Countries with observations"
  )
```

```{r cumulative-country-coverage, fig.width=9, fig.height=4.8}
ggplot(cumulative_countries, aes(x = year_value, y = cumulative_countries)) +
  geom_area(fill = "#8ECAE6", alpha = 0.55) +
  geom_line(color = "#1F5B74", linewidth = 1.1) +
  geom_point(color = "#1F5B74", size = 2.2) +
  scale_x_continuous(breaks = start_year:end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Cumulative country coverage over time",
    subtitle = "The curve rises when a country appears in the analysed period for the first time.",
    x = "Year", y = "Cumulative countries"
  )
```

```{r new-countries-by-year, fig.width=9, fig.height=4.6}
# the first year of the window (2015) is left out of this chart on purpose:
# every country that has any observation in 2015 automatically counts as
# "new" just because 2015 is where our data starts, not because it is
# really a new country for tree recording. Including it would make the
# 2015 bar huge and misleading, so this chart only shows real new entries
# from 2016 onward.
new_countries <- country_first_year %>%
  filter(first_observed_year > start_year) %>%
  count(first_observed_year, name = "new_countries") %>%
  right_join(tibble(first_observed_year = (start_year + 1):end_year), by = "first_observed_year") %>%
  mutate(new_countries = ifelse(is.na(new_countries), 0L, new_countries)) %>%
  arrange(first_observed_year)

countries_at_start <- sum(country_first_year$first_observed_year == start_year)
```

```{r new-countries-plot, fig.width=9, fig.height=4.6}
ggplot(new_countries, aes(x = first_observed_year, y = new_countries)) +
  geom_col(fill = "#4E79A7", width = 0.72) +
  scale_x_continuous(breaks = (start_year + 1):end_year) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "New countries entering the dataset by year",
    subtitle = "New country entries indicate expansion of geographic coverage within the analysed period.",
    x = "First observed year in period", y = "New countries",
    caption = paste0(start_year, " is excluded: ", countries_at_start,
                      " countries already had observations in the first year of the\nwindow, which is not the same as those countries being newly added.")
  )
```

**Interpretation.** If the number of countries and cumulative country coverage increase over time, GlobalGeoTree is expanding geographically. If observation counts increase without much country expansion, growth may be concentrated in already well-covered countries.

## Which Countries Drive Temporal Growth?

```{r country-growth-data}
country_growth <- country_year_complete %>%
  arrange(country_value, year_value) %>%
  group_by(country_value) %>%
  mutate(country_change = observations - lag(observations)) %>%
  ungroup()

largest_country_growth <- country_growth %>%
  filter(!is.na(country_change)) %>%
  slice_max(country_change, n = top_n_countries, with_ties = FALSE) %>%
  mutate(label = paste0(year_value, ": ", country_value))
```

```{r country-growth-plot, fig.width=9, fig.height=5}
ggplot(largest_country_growth, aes(x = reorder(label, country_change), y = country_change)) +
  geom_col(fill = "#2A7F62", width = 0.72) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Largest country-specific yearly increases",
    subtitle = "Temporal growth may be concentrated in specific countries rather than global.",
    x = NULL, y = "Increase from previous year"
  )
```

# Map Snapshots for Selected Years

```{r map-snapshot-data}
snapshot_years <- snapshot_years[snapshot_years >= start_year & snapshot_years <= end_year]

if (has_coordinates) {
  map_data_points <- df %>%
    filter(
      year_value %in% snapshot_years,
      !is.na(latitude_value), !is.na(longitude_value),
      between(latitude_value, -90, 90), between(longitude_value, -180, 180)
    ) %>%
    group_by(year_value) %>%
    slice_sample(n = max_points_per_snapshot_year) %>%
    ungroup()

  world_outline <- if (requireNamespace("maps", quietly = TRUE)) ggplot2::map_data("world") else NULL
}
```

```{r map-snapshots, fig.width=9.5, fig.height=6.8}
if (has_coordinates && nrow(map_data_points) > 0) {
  p <- ggplot()

  if (!is.null(world_outline)) {
    p <- p + geom_polygon(
      data = world_outline, aes(x = long, y = lat, group = group),
      fill = "gray96", color = "gray82", linewidth = 0.12, inherit.aes = FALSE
    )
  }

  p +
    geom_point(
      data = map_data_points, aes(x = longitude_value, y = latitude_value),
      alpha = 0.15, size = 0.22, color = "#1F5B74"
    ) +
    facet_wrap(~ year_value, ncol = 2) +
    coord_quickmap(xlim = c(-180, 180), ylim = c(-60, 85), expand = FALSE) +
    labs(
      title = "Map snapshots of observations over time",
      subtitle = "Selected years show whether records spread geographically or intensify in existing regions.",
      x = "Longitude", y = "Latitude",
      caption = "Points are sampled per year for readability. Install maps for country outlines."
    )
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable coordinates were detected, so map snapshots are skipped.", size = 4) +
    theme_void()
}
```

```{r grid-coverage-data}
if (has_coordinates) {
  grid_coverage <- df %>%
    filter(
      !is.na(latitude_value), !is.na(longitude_value),
      between(latitude_value, -90, 90), between(longitude_value, -180, 180)
    ) %>%
    mutate(
      lat_bin = floor(latitude_value / 5) * 5,
      lon_bin = floor(longitude_value / 5) * 5,
      grid_cell = paste(lat_bin, lon_bin, sep = "_")
    ) %>%
    distinct(year_value, grid_cell)

  grid_first_year <- grid_coverage %>%
    group_by(grid_cell) %>%
    summarise(first_observed_year = min(year_value), .groups = "drop")

  cumulative_grid_cells <- tibble(year_value = start_year:end_year) %>%
    rowwise() %>%
    mutate(cumulative_grid_cells = sum(grid_first_year$first_observed_year <= year_value)) %>%
    ungroup()
}
```

```{r cumulative-grid-coverage, fig.width=9, fig.height=4.8}
if (has_coordinates && nrow(cumulative_grid_cells) > 0) {
  ggplot(cumulative_grid_cells, aes(x = year_value, y = cumulative_grid_cells)) +
    geom_area(fill = "#B2DF8A", alpha = 0.65) +
    geom_line(color = "#2A7F62", linewidth = 1.1) +
    geom_point(color = "#2A7F62", size = 2.2) +
    scale_x_continuous(breaks = start_year:end_year) +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Cumulative spatial grid coverage over time",
      subtitle = "A 5-degree latitude/longitude grid approximates whether spatial coverage expands.",
      x = "Year", y = "Cumulative occupied grid cells"
    )
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No usable coordinates were detected.", size = 4) +
    theme_void()
}
```

**Interpretation.** Map snapshots and grid coverage distinguish two different processes: more observations in the same places versus expansion into new regions.

# Overall Interpretation

From 2015 to 2024, temporal development in GlobalGeoTree should be interpreted as a combination of:

- changing observation volume,
- expansion or concentration of geographic coverage,
- country-specific growth,
- and source/platform growth.

If the strongest year-to-year increases are driven by citizen-science platforms or a small number of countries, the temporal trend is not purely biological. It is partly a record-production trend. Therefore, temporal analyses should report both total observations and source composition by year.

Two specific things to keep in mind before treating the headline numbers as final:

- **2024 is likely an incomplete year** in this export, so the apparent decline at the end of the series may shrink or disappear once a full year of data is available.
- **The "new countries" chart excludes 2015** because countries present at the very start of the window are not the same as countries that newly entered the dataset; counting them as new would overstate geographic expansion.
