---
title: "GlobalGeoTree — Climate–Family Dependencies"
subtitle: "Direct exploration: which families prefer which climates, and where (GGT-10kEval-90)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## 0. Why this document is separate from the PCA/t-SNE one

The earlier exploration leaned on PCA and t-SNE to find structure
automatically. On this dataset that approach hit a ceiling: `level0`
(deciduous/evergreen broadleaf/evergreen needleleaf) doesn't separate
cleanly in unsupervised projections, partly because it's a coarse 3-class
label and partly because compressed components (PC1, PC2, t-SNE/UMAP axes)
are hard to map back to "this family likes cold places."

This document asks the question directly instead: **for specific families,
do climate and elevation values differ in the way ecological intuition
suggests** (e.g. conifers skewing colder/higher, certain families skewing
warmer/drier)? Every section below uses real, interpretable variables
(°C, mm/yr, meters) rather than abstract components, and ends in either a
table or a map you can sanity-check against known geography.

```{r setup}
library(jsonlite)
library(dplyr)
library(purrr)
library(tibble)
library(stringr)
library(ggplot2)
library(tidyr)
library(forcats)

theme_set(theme_minimal())
```

## 1. Loading the data

Same robust-by-name loader as before — set `root` to your local extracted
`GGT-10kEval-90` folder and switch the chunk to `eval: true` (or just run
it interactively) once you're working with the real data. A synthetic
stand-in further below lets this whole document render without it.

```{r load-data}
#| eval: false

root <- "/Users/nataliiaremezova/Documents/Workspace/BHT/DataViz/GGT-10kEval-90"

aux_files <- list.files(
  root,
  pattern = "\\.auxiliary\\.json$",
  recursive = TRUE,
  full.names = TRUE
)

aux_vars <- c(
  "longitude", "latitude", "elevation", "slope", "aspect",
  "soil_moisture_0_5cm", "soil_moisture_5_15cm", "soil_moisture_15_30cm",
  paste0("bio", sprintf("%02d", 1:19))
)

read_one <- function(aux_path) {
  key <- basename(aux_path) |> str_remove("\\.auxiliary\\.json$")
  text_path <- file.path(dirname(aux_path), paste0(key, ".text.json"))

  aux   <- fromJSON(aux_path)
  label <- fromJSON(text_path)

  aux_row <- aux[aux_vars]
  aux_row[sapply(aux_row, is.null)] <- NA

  tibble(
    key     = key,
    level0  = label$level0,
    family  = label$level1_family,
    genus   = label$level2_genus,
    species = label$level3_species
  ) |>
    bind_cols(as_tibble(aux_row))
}

df <- map_dfr(aux_files, read_one)

df <- df |>
  rename(
    annual_mean_temp = bio01, mean_diurnal_temp_range = bio02, isothermality = bio03,
    temp_seasonality = bio04, max_temp_warmest_month = bio05, min_temp_coldest_month = bio06,
    temp_annual_range = bio07, mean_temp_wettest_quarter = bio08, mean_temp_driest_quarter = bio09,
    mean_temp_warmest_quarter = bio10, mean_temp_coldest_quarter = bio11,
    annual_precipitation = bio12, precip_wettest_month = bio13, precip_driest_month = bio14,
    precip_seasonality = bio15, precip_wettest_quarter = bio16, precip_driest_quarter = bio17,
    precip_warmest_quarter = bio18, precip_coldest_quarter = bio19
  )
```

```{r simulate-data}
#| echo: false
# Synthetic stand-in with deliberately built-in family-climate relationships
# (e.g. Pinaceae skewed cold/high-elevation, Fabaceae skewed warm/dry) so
# the demonstration plots below show what a real signal looks like.
# DELETE this chunk once you're rendering against the real loader above.
set.seed(7)
n <- 4000

family_profiles <- tibble(
  family = c("Pinaceae", "Betulaceae", "Salicaceae", "Fagaceae",
             "Rosaceae", "Fabaceae", "Proteaceae", "Araliaceae"),
  temp_mean_C = c(2, 6, 8, 11, 13, 22, 19, 24),
  temp_sd_C   = c(5, 5, 6, 6, 7, 5, 6, 4),
  precip_mean = c(700, 850, 750, 1000, 900, 450, 600, 1800),
  precip_sd   = c(250, 300, 300, 350, 350, 250, 300, 500),
  elev_mean   = c(1400, 600, 500, 450, 500, 350, 700, 200),
  elev_sd     = c(700, 400, 350, 350, 400, 300, 500, 250),
  level0      = c("Evergreen Needleleaf", "Deciduous Broadleaf", "Deciduous Broadleaf",
                   "Deciduous Broadleaf", "Deciduous Broadleaf", "Deciduous Broadleaf",
                   "Evergreen Broadleaf", "Evergreen Broadleaf")
)

sampled_family <- sample(family_profiles$family, n, replace = TRUE)
fp <- family_profiles[match(sampled_family, family_profiles$family), ]

temp_C  <- rnorm(n, fp$temp_mean_C, fp$temp_sd_C)
precip  <- pmax(50, rnorm(n, fp$precip_mean, fp$precip_sd))
elev    <- pmax(0, rnorm(n, fp$elev_mean, fp$elev_sd))

# Rough lat/lon so cold/high-elevation families land at higher latitudes,
# loosely centered on the continental US for the map demo in Section 5.
lat <- pmin(70, pmax(20, 50 - temp_C * 0.9 + rnorm(n, 0, 6)))
lon <- runif(n, -125, -66)

df <- tibble(
  key = sprintf("%06d", seq_len(n)),
  level0 = fp$level0,
  family = sampled_family,
  genus = paste0(substr(sampled_family, 1, 4), sample(1:6, n, replace = TRUE)),
  species = paste(genus, sample(letters[1:10], n, replace = TRUE)),
  longitude = lon, latitude = lat,
  elevation = elev,
  slope = pmax(0, rnorm(n, 8, 6)),
  aspect = sample(0:359, n, replace = TRUE),
  soil_moisture_0_5cm = pmax(0, rnorm(n, 25, 8)),
  soil_moisture_5_15cm = pmax(0, rnorm(n, 27, 8)),
  soil_moisture_15_30cm = pmax(0, rnorm(n, 29, 8)),
  annual_mean_temp = temp_C * 10,
  temp_seasonality = pmax(500, rnorm(n, 2500, 800)),
  annual_precipitation = precip
)
```

## 1.5 Common names for families

Latin family names are precise but not friendly to a non-biologist
audience. This builds a lookup table mapping the most common tree families
to their everyday English names and joins it onto the data — used in
tables/legends from here on instead of raw Latin where it helps
readability. Families not in this list (there are 275 in the full
dataset) fall back to "Latin name only" rather than `NA`.

```{r common-names}
family_common_names <- tribble(
  ~family,          ~common_name,
  "Pinaceae",       "Pine family",
  "Betulaceae",     "Birch family",
  "Salicaceae",     "Willow family",
  "Fagaceae",       "Beech/Oak family",
  "Rosaceae",       "Rose family",
  "Fabaceae",       "Legume/Pea family",
  "Proteaceae",     "Protea family",
  "Araliaceae",     "Ginseng/Ivy family",
  "Ulmaceae",       "Elm family",
  "Viburnaceae",    "Viburnum family",
  "Aceraceae",      "Maple family",
  "Sapindaceae",    "Soapberry/Maple family",
  "Oleaceae",       "Olive/Ash family",
  "Juglandaceae",   "Walnut family",
  "Cupressaceae",   "Cypress family",
  "Taxaceae",       "Yew family",
  "Myrtaceae",      "Myrtle/Eucalyptus family",
  "Lauraceae",      "Laurel family",
  "Magnoliaceae",   "Magnolia family",
  "Moraceae",       "Mulberry/Fig family",
  "Anacardiaceae",  "Cashew/Sumac family",
  "Caprifoliaceae", "Honeysuckle family",
  "Ericaceae",      "Heath family",
  "Malvaceae",      "Mallow family",
  "Bignoniaceae",   "Trumpet vine family",
  "Cornaceae",      "Dogwood family",
  "Hamamelidaceae", "Witch-hazel family",
  "Platanaceae",    "Plane/Sycamore family",
  "Cactaceae",      "Cactus family",
  "Asteraceae",     "Daisy/Sunflower family",
  "Myricaceae",     "Bayberry family",
  "Ebenaceae",      "Ebony family",
  "Theaceae",       "Tea family",
  "Meliaceae",      "Mahogany family",
  "Combretaceae",   "Combretum family",
  "Arecaceae",      "Palm family"
)

df <- df |>
  left_join(family_common_names, by = "family") |>
  mutate(
    family_label = if_else(is.na(common_name),
                            family,
                            paste0(common_name, " (", family, ")"))
  )

# Quick check: which of your top families aren't in the lookup yet?
df |> distinct(family, common_name) |> filter(is.na(common_name))
```

📝 The last line is meant to be checked every time you rerun on real data
— if it lists families you actually care about, just add a row to
`family_common_names` above. The full GlobalGeoTree dataset spans 275
families, so this lookup deliberately only covers commonly known ones
rather than trying to be exhaustive.

The most direct answer to "does family X like it cold/hot/dry/wet/high":
group by family and summarize. No projection, no dimensionality reduction
— just the numbers.

```{r family-niche-table}
top_families <- df |> count(family, sort = TRUE) |> slice_head(n = 10) |> pull(family)

family_niche <- df |>
  filter(family %in% top_families) |>
  group_by(family, family_label) |>
  summarise(
    n = n(),
    mean_temp_C = mean(annual_mean_temp / 10, na.rm = TRUE),
    sd_temp_C = sd(annual_mean_temp / 10, na.rm = TRUE),
    mean_precip_mm = mean(annual_precipitation, na.rm = TRUE),
    sd_precip_mm = sd(annual_precipitation, na.rm = TRUE),
    mean_elevation_m = mean(elevation, na.rm = TRUE),
    sd_elevation_m = sd(elevation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  relocate(family_label, .after = family) |>
  arrange(mean_temp_C)

family_niche
```

```{r family-temp-ranked-plot}
family_niche |>
  mutate(family_label = fct_reorder(family_label, mean_temp_C)) |>
  ggplot(aes(family_label, mean_temp_C)) +
  geom_pointrange(aes(ymin = mean_temp_C - sd_temp_C, ymax = mean_temp_C + sd_temp_C)) +
  coord_flip() +
  labs(title = "Mean annual temperature by family (± 1 SD)",
       x = NULL, y = "°C")
```

📝 **Reading this plot:** families near the top have the coldest typical
climate (point position) and the narrowest/widest range of climates they
tolerate (error bar length). A family with a tight error bar around a cold
mean is a real candidate for "genuinely cold-adapted family"; a family
with a huge error bar spanning both cold and warm temperatures is more
likely just widely distributed, not climate-specialized.

## 3. Climate niche scatter — temperature vs. precipitation

This is the direct, interpretable version of what the PCA biplot was
trying to show, except both axes are real units instead of abstract
components — closer to a classic Whittaker biome diagram than to a PCA
plot.

```{r niche-scatter}
df |>
  filter(family %in% top_families) |>
  ggplot(aes(annual_mean_temp / 10, annual_precipitation, color = family_label)) +
  geom_point(alpha = 0.25, size = 0.8) +
  stat_ellipse(level = 0.68, linewidth = 0.8) +
  labs(title = "Climate niche by family — top 10 families",
       x = "Annual mean temperature (°C)", y = "Annual precipitation (mm/yr)",
       color = "Family")
```

📝 The ellipses (68% confidence region per family) are doing the same job
the PCA arrows tried to do, but per-family and in real units. Families
whose ellipses barely overlap have a genuinely distinguishable climate
niche; families whose ellipses sit on top of each other are climate
generalists, or this particular curated sample doesn't capture their niche
distinctly. This plot is also where you'd visually spot something like
"family X occupies the cold/wet corner" — directly, without inferring it
from PC loadings.

## 4. Elevation by family

Tests the "pines like high/cold places" intuition directly.

```{r elevation-boxplot}
df |>
  filter(family %in% top_families) |>
  mutate(family_label = fct_reorder(family_label, elevation, .fun = median)) |>
  ggplot(aes(family_label, elevation, fill = family_label)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Elevation distribution by family", x = NULL, y = "Elevation (m)")
```

```{r elevation-temp-joint}
df |>
  filter(family %in% top_families) |>
  ggplot(aes(elevation, annual_mean_temp / 10, color = family_label)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(title = "Elevation vs. temperature by family",
       x = "Elevation (m)", y = "Annual mean temperature (°C)", color = "Family")
```

📝 The second plot also serves as a sanity check on the data itself: all
families should show a *negative* elevation–temperature slope (the
physical lapse rate, ~6.5°C per 1000m) regardless of family identity. If
some family's line goes the wrong way, that's worth investigating as a
possible data or join issue before reading anything ecological into it.

## 5. Regional map — continental US

Filtering to a bounding box and plotting real coordinates is far more
checkable than an abstract embedding — you can directly compare against
known geography (e.g. conifers in the Rockies/Pacific Northwest vs.
broadleaf in the East).

🛠 **Fix — clip to actual land, not a rectangle.** A plain lat/lon box
(`-125 to -66, 24 to 50`) also catches a strip of southern Canada, northern
Mexico, and ocean near the coasts. `maps::map.where()` does a real
point-in-polygon test against US state outlines, so points outside the
country's actual land area get dropped instead of just clipped to a
rectangle. This matters more once you're on real data — with the
synthetic placeholder, longitude is pure `runif()` with no relationship to
land at all, so even this fix won't make the placeholder maps look
geographically meaningful; it'll matter once real coordinates are in.

```{r us-map}
# install.packages("maps")  # ships US state borders, no API key needed
library(maps)

us_bbox <- df |>
  filter(longitude > -125, longitude < -66, latitude > 24, latitude < 50) |>
  filter(family %in% top_families) |>
  mutate(us_state = map.where("state", longitude, latitude)) |>
  filter(!is.na(us_state))   # drops ocean / Canada / Mexico points

ggplot(us_bbox, aes(longitude, latitude)) +
  borders("state", fill = "grey95", colour = "grey70") +
  geom_point(aes(color = family_label, size = elevation), alpha = 0.6) +
  scale_size_continuous(range = c(0.5, 3), name = "Elevation (m)") +
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  labs(title = "Continental US occurrences by family",
       subtitle = "Point size = elevation",
       x = NULL, y = NULL, color = "Family") +
  theme(panel.grid = element_blank())
```

```{r us-map-temp}
ggplot(us_bbox, aes(longitude, latitude)) +
  borders("state", fill = "grey95", colour = "grey70") +
  geom_point(aes(color = annual_mean_temp / 10), alpha = 0.7, size = 1) +
  scale_color_viridis_c(option = "inferno", name = "°C") +
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  facet_wrap(~family_label) +
  labs(title = "Annual mean temperature at occurrence points, by family",
       x = NULL, y = NULL) +
  theme(panel.grid = element_blank())
```

📝 **Reading these maps.** The first map is the most intuitive check —
does a cold-climate family actually cluster in the Rockies/Pacific
Northwest/New England rather than the Gulf Coast? The faceted second map
controls for the fact that "where points are" and "how warm those points
are" can both vary — a family could be regionally clustered for reasons
unrelated to temperature (sampling effort, data source), so check whether
the color (temperature) pattern within each facet looks geographically
sensible (warmer toward the south/coasts, cooler at elevation/north) before
attributing a pattern to the family's climate preference specifically.

⚠️ **If you're still seeing points scattered uniformly with no
geographic structure (e.g. in the ocean, or no visible north/south
temperature gradient) after rerunning this**, that's a sign the
`load-data` chunk is still set to `eval: false` and you're looking at the
synthetic placeholder rather than real GlobalGeoTree coordinates — flip
that chunk on and delete the `simulate-data` chunk below it.

⚠️ Note on cacti: GlobalGeoTree is **tree occurrences only**, so true
cacti (Cactaceae) are unlikely to appear — most cacti aren't trees. If
you're looking for the "loves it hot and dry" analogue in this dataset,
check which genera/families actually show up in the arid Southwest in the
map above (candidates: *Prosopis* (mesquite), *Acacia*-type Fabaceae
genera, or *Yucca* if classified as tree-form here) rather than assuming
cacti specifically will be present.

## 6. A direct, supervised check: does climate predict family?

Rather than reading tea leaves from a 2D projection, fit a small model and
look at which variables it actually uses. This converts "I think I see
clustering" into a number.

```{r rf-importance}
#| eval: false
# install.packages("randomForest")
library(randomForest)

rf_data <- df |>
  filter(family %in% top_families) |>
  select(family, elevation, slope, annual_mean_temp, temp_seasonality,
         annual_precipitation) |>
  mutate(family = factor(family)) |>
  na.omit()

set.seed(1)
rf <- randomForest(family ~ ., data = rf_data, ntree = 300, importance = TRUE)

# class.error keyed by Latin name (model's factor levels) — join back to
# common names for a readable readout
tibble(family = rownames(rf$confusion), class_error = rf$confusion[, "class.error"]) |>
  left_join(family_common_names, by = "family") |>
  arrange(class_error)

varImpPlot(rf, main = "Variable importance for predicting family from climate")
```

📝 **How to read this.** `class.error` per family tells you, concretely,
which families *are* climate-distinguishable in this data (low error) and
which aren't (high error, near the ~90% chance level for 10 balanced
classes) — a much more direct readout than squinting at cluster overlap in
a scatter plot. The importance plot tells you which of elevation,
temperature, seasonality, or precipitation is actually doing the
distinguishing work, which directly answers "is it the cold, the dryness,
or the altitude that matters for this family" rather than leaving you to
guess from a PCA loading.

## 7. Summary / where to go next

- Skip PCA/t-SNE as the entry point for this question — they compress
  variance, not climate-family relationships specifically, and on this
  dataset the variance they compress doesn't line up cleanly with family.
- The niche table (Section 2) and scatter (Section 3) directly answer
  "does family X prefer cold/hot/wet/dry," in real units, per family.
- The elevation analysis (Section 4) is the cleanest test of a "pines like
  high/cold places"-style hypothesis, and includes a built-in sanity check
  (the lapse rate) for catching data issues.
- The US map (Section 5) is the most interpretable visual — it can be
  directly compared against known biogeography (conifers in the
  mountains/north, broadleaf in the east, etc.), which an abstract
  embedding can't offer.
- The random forest check (Section 6) turns "I think I see structure" into
  a per-family error rate and a ranked list of which climate variables
  matter — worth running before writing up any claims about which families
  are climate-specialized.
- Natural extension: repeat Sections 2–5 on the 300- or 900-species eval
  splits for more samples per family, and restrict the US map to a single
  well-known contrast pair (e.g. Pinaceae vs. a warm-climate broadleaf
  family) for the clearest possible illustration.
