# ---
title: "Germany Tree Observations: Frequency, Sources, Bundesländer, and Climate Zones"
author: "GlobalGeoTree team"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

# Purpose

This report checks and visualizes Germany observations from the larger GlobalGeoTree dataset. It focuses on five questions:

1. How are dataset-defined tree frequency categories (`Rare`, `Common`, `Frequent`) distributed across the German map?
2. How many observations are recorded by year?
3. How are observations from different sources distributed across Germany?
4. How are observations mapped to German Bundesländer, and how do observation frequencies differ by Bundesland?
5. Can observations be mapped to external climate-zone polygons?

> Important: the full GlobalGeoTree dataset may not contain the `category` variable. If it does not, this report joins `Rare`, `Common`, and `Frequent` from the larger eval dataset `GlobalGeoTree-10kEval-900.csv`. In that case, category-based visualizations only describe species that can be matched to that eval file.

# Setup

```{r setup}
# Required packages only.
# data.table: fast CSV reading
# dplyr/stringr: data cleaning and summaries
# ggplot2/scales: visualizations
# sf: spatial joins and maps
required_packages <- c("data.table", "dplyr", "stringr", "ggplot2", "scales", "sf")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Please install the following packages before rendering this report: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(sf)

set.seed(42)

# -------------------------------------------------------------------------
# User configuration
# -------------------------------------------------------------------------

# Main observations file. If the full file is present, it is used.
# Otherwise, the 900-species eval file is used as fallback.
observations_path <- if (file.exists("GlobalGeoTree.csv")) {
  "GlobalGeoTree.csv"
} else if (file.exists("GlobalGeoTree-10kEval-900.csv")) {
  "GlobalGeoTree-10kEval-900.csv"
} else {
  "GlobalGeoTree-10kEval-90.csv"
}

# Preferred category lookup. This should contain the Category column.
eval_lookup_path <- if (file.exists("GlobalGeoTree-10kEval-900.csv")) {
  "GlobalGeoTree-10kEval-900.csv"
} else if (file.exists("GlobalGeoTree-10kEval-90.csv")) {
  "GlobalGeoTree-10kEval-90.csv"
} else {
  NA_character_
}

# Germany country code.
focus_country_code <- "DE"

# Maximum number of points displayed in point-based maps.
# Counts and summaries still use all Germany rows.
max_points_per_source_map <- 20000
max_points_per_category_map <- 20000

# Bundesländer boundaries.
# The report will try to download/cache GISCO NUTS1 boundaries if this file is absent.
bundeslaender_path <- "data/external/germany_nuts1_2021.geojson"

# Optional climate-zone polygon file.
# This must be an sf-readable file, for example GeoJSON or Shapefile.
# The tree dataset itself does not contain climate zones.
climate_zones_path <- "data/external/climate_zones_germany.geojson"

# If your climate file uses a known column for the climate-zone label, put it here.
# If NULL, the report tries to detect a reasonable label column.
climate_zone_column <- NULL
```

```{r helper-functions}
clean_colnames <- function(x) {
  x <- tolower(gsub("[^A-Za-z0-9]+", "_", trimws(x)))
  x <- gsub("^_+|_+$", "", x)
  make.unique(x, sep = "_")
}

read_csv_clean <- function(path, cols = NULL) {
  raw_names <- names(data.table::fread(path, nrows = 0, showProgress = FALSE))
  clean_names <- clean_colnames(raw_names)
  
  if (is.null(cols)) {
    dt <- data.table::fread(path, showProgress = FALSE)
    names(dt) <- clean_names
    return(as_tibble(dt))
  }
  
  keep_clean <- intersect(cols, clean_names)
  if (length(keep_clean) == 0) {
    stop("None of the requested columns were found in ", path)
  }
  
  keep_indices <- match(keep_clean, clean_names)
  dt <- data.table::fread(path, select = raw_names[keep_indices], showProgress = FALSE)
  names(dt) <- clean_names[keep_indices]
  as_tibble(dt)
}

mode_value <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return(NA_character_)
  names(which.max(table(x)))
}

clean_category <- function(x) {
  x <- stringr::str_squish(as.character(x))
  x <- stringr::str_to_title(x)
  x[!(x %in% c("Rare", "Common", "Frequent"))] <- NA_character_
  x
}

sample_up_to <- function(data, n_max) {
  if (nrow(data) <= n_max) {
    data
  } else {
    dplyr::slice_sample(data, n = n_max)
  }
}

theme_gtree <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(size = base_size, color = "gray35"),
      plot.caption = element_text(size = base_size - 2, color = "gray45"),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")
    )
}

theme_map_gtree <- function(base_size = 12) {
  theme_void(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3),
      plot.subtitle = element_text(size = base_size, color = "gray35"),
      plot.caption = element_text(size = base_size - 2, color = "gray45"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

format_number <- function(x) scales::comma(x, accuracy = 1)
```

# Load Germany observations

```{r load-data}
# First read only the columns needed for this analysis.
# Category is optional because the full dataset may not contain it.
requested_cols <- c(
  "sample_id", "country_code", "level0", "level1_family", "level2_genus",
  "level3_species", "location", "source", "species_key", "year",
  "longitude", "latitude", "category"
)

obs <- read_csv_clean(observations_path, cols = requested_cols)

required_cols <- c("country_code", "longitude", "latitude", "level3_species")
missing_required <- setdiff(required_cols, names(obs))
if (length(missing_required) > 0) {
  stop("The observations file is missing required columns: ", paste(missing_required, collapse = ", "))
}

obs <- obs |>
  mutate(
    country_code = str_to_upper(str_squish(as.character(country_code))),
    level3_species = str_squish(as.character(level3_species)),
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude),
    year = if ("year" %in% names(obs)) as.integer(year) else NA_integer_,
    source = if ("source" %in% names(obs)) str_squish(as.character(source)) else "Unknown source",
    species_key = if ("species_key" %in% names(obs)) as.character(species_key) else NA_character_
  ) |>
  filter(
    country_code == focus_country_code,
    !is.na(longitude), !is.na(latitude),
    between(longitude, -180, 180),
    between(latitude, -90, 90),
    !is.na(level3_species), level3_species != ""
  )

if (!("category" %in% names(obs))) {
  obs$category <- NA_character_
} else {
  obs$category <- clean_category(obs$category)
}

germany_raw <- obs
```

## Data preview

```{r table-head-full}
head_table <- germany_raw |>
  select(any_of(c(
    "sample_id", "country_code", "level3_species", "species_key",
    "category", "source", "year", "longitude", "latitude"
  ))) |>
  head(10)

knitr::kable(
  head_table,
  caption = "First 10 Germany rows after filtering and basic cleaning."
)
```

# Add frequency categories from the eval dataset

```{r join-categories}
category_source_note <- "Category was read directly from the main observations file."

if (any(is.na(germany_raw$category)) && !is.na(eval_lookup_path) && file.exists(eval_lookup_path)) {
  eval_lookup <- read_csv_clean(
    eval_lookup_path,
    cols = c("species_key", "level3_species", "category")
  )
  
  if (!("species_key" %in% names(eval_lookup))) eval_lookup$species_key <- NA_character_
  if (!("level3_species" %in% names(eval_lookup))) eval_lookup$level3_species <- NA_character_
  if (!("category" %in% names(eval_lookup))) eval_lookup$category <- NA_character_
  
  eval_lookup <- eval_lookup |>
    mutate(
      species_key = as.character(species_key),
      level3_species = str_squish(as.character(level3_species)),
      category = clean_category(category)
    ) |>
    filter(!is.na(category))
  
  # Join by species_key first where possible.
  if ("species_key" %in% names(germany_raw) && "species_key" %in% names(eval_lookup)) {
    lookup_by_key <- eval_lookup |>
      filter(!is.na(species_key), species_key != "") |>
      group_by(species_key) |>
      summarise(category_lookup = mode_value(category), .groups = "drop")
    
    germany_raw <- germany_raw |>
      left_join(lookup_by_key, by = "species_key") |>
      mutate(category = coalesce(category, category_lookup)) |>
      select(-category_lookup)
  }
  
  # For still-unmatched rows, join by species name.
  lookup_by_species <- eval_lookup |>
    filter(!is.na(level3_species), level3_species != "") |>
    group_by(level3_species) |>
    summarise(category_lookup = mode_value(category), .groups = "drop")
  
  germany_raw <- germany_raw |>
    left_join(lookup_by_species, by = "level3_species") |>
    mutate(category = coalesce(category, category_lookup)) |>
    select(-category_lookup)
  
  category_source_note <- paste0(
    "Missing Rare/Common/Frequent values were filled from ",
    eval_lookup_path,
    " by species_key first and species name second."
  )
}

germany <- germany_raw |>
  mutate(
    category = if_else(is.na(category), "Unknown", category),
    category = factor(category, levels = c("Rare", "Common", "Frequent", "Unknown")),
    source = if_else(is.na(source) | source == "", "Unknown source", source)
  )

category_source_note
```

```{r overview-metrics}
overview_metrics <- tibble::tibble(
  metric = c(
    "Germany observations",
    "Germany recorded species",
    "Germany observations with known category",
    "Germany recorded species with known category",
    "Germany observations with unknown category",
    "Germany recorded species with unknown category"
  ),
  value = c(
    nrow(germany),
    n_distinct(germany$level3_species),
    sum(germany$category != "Unknown"),
    n_distinct(germany$level3_species[germany$category != "Unknown"]),
    sum(germany$category == "Unknown"),
    n_distinct(germany$level3_species[germany$category == "Unknown"])
  )
)

knitr::kable(
  overview_metrics |>
    mutate(value = format_number(value)),
  caption = "Germany category coverage summary."
)
```

# Load Bundesländer boundaries

```{r load-bundeslaender}
dir.create(dirname(bundeslaender_path), recursive = TRUE, showWarnings = FALSE)

gisco_nuts1_url <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_4326_LEVL_1.geojson"

if (!file.exists(bundeslaender_path)) {
  try(
    utils::download.file(gisco_nuts1_url, bundeslaender_path, mode = "wb", quiet = TRUE),
    silent = TRUE
  )
}

bundeslaender <- tryCatch(
  {
    path_to_read <- if (file.exists(bundeslaender_path)) bundeslaender_path else gisco_nuts1_url
    x <- sf::read_sf(path_to_read)
    names(x) <- clean_colnames(names(x))
    x <- sf::st_transform(x, 4326)
    x <- sf::st_make_valid(x)
    
    if ("cntr_code" %in% names(x)) x <- filter(x, cntr_code == "DE")
    if ("levl_code" %in% names(x)) x <- filter(x, levl_code == 1)
    
    name_col <- intersect(c("name_latn", "nuts_name", "name", "nuts_id"), names(x))[1]
    if (is.na(name_col)) stop("Could not identify Bundesland name column.")
    
    x |>
      mutate(bundesland = .data[[name_col]]) |>
      select(bundesland, geometry)
  },
  error = function(e) NULL
)

has_bundeslaender <- !is.null(bundeslaender) && nrow(bundeslaender) > 0

if (has_bundeslaender) {
  germany_boundary <- sf::st_union(bundeslaender)
  germany_bbox <- sf::st_bbox(bundeslaender)
} else {
  germany_boundary <- NULL
  germany_bbox <- c(
    xmin = min(germany$longitude, na.rm = TRUE),
    xmax = max(germany$longitude, na.rm = TRUE),
    ymin = min(germany$latitude, na.rm = TRUE),
    ymax = max(germany$latitude, na.rm = TRUE)
  )
}

if (!has_bundeslaender) {
  cat("Bundesländer boundaries could not be loaded. Maps will render without Bundesländer borders. Check your internet connection or place a boundary file at `", bundeslaender_path, "`.", sep = "")
}
```

# 1. Frequency categories across the German map

This map shows where observations of species with known `Rare`, `Common`, or `Frequent` category labels occur. Unknown category observations are omitted from this figure because they cannot be assigned to one of the three frequency categories.

```{r map-category-frequency, fig.width=9, fig.height=6.5}
germany_known_category <- germany |>
  filter(category %in% c("Rare", "Common", "Frequent")) |>
  droplevels()

if (nrow(germany_known_category) == 0) {
  plot.new()
  text(0.5, 0.5, "No Germany observations have known Rare/Common/Frequent categories.")
} else {
  p <- ggplot()
  
  if (has_bundeslaender) {
    p <- p +
      geom_sf(data = germany_boundary, fill = "gray97", color = "gray55", linewidth = 0.3)
  }
  
  p +
    stat_bin_2d(
      data = germany_known_category,
      aes(x = longitude, y = latitude, fill = after_stat(count)),
      bins = 45,
      alpha = 0.85
    ) +
    facet_wrap(~ category) +
    scale_fill_gradient(low = "#deebf7", high = "#08519c", labels = comma_format()) +
    coord_sf(
      xlim = c(germany_bbox["xmin"], germany_bbox["xmax"]),
      ylim = c(germany_bbox["ymin"], germany_bbox["ymax"]),
      expand = FALSE
    ) +
    labs(
      title = "Spatial distribution of known frequency categories in Germany",
      subtitle = "Each panel bins observations belonging to species labeled Rare, Common, or Frequent in the eval lookup.",
      fill = "Observations",
      caption = "Unknown categories are excluded. If category was joined from the eval file, this map covers only matched species."
    ) +
    theme_map_gtree()
}
```

```{r category-counts-plot, fig.width=8, fig.height=4.5}
category_counts <- germany |>
  count(category, name = "observations") |>
  mutate(share = observations / sum(observations))

ggplot(category_counts, aes(x = category, y = observations)) +
  geom_col(width = 0.65, fill = "#3182bd") +
  geom_text(aes(label = paste0(format_number(observations), "\n", percent(share, accuracy = 0.1))),
            vjust = -0.2, size = 3.5) +
  scale_y_continuous(labels = comma_format(), expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Germany observations by category coverage",
    subtitle = "The Unknown group represents observations whose species could not be matched to a Rare/Common/Frequent label.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree() +
  theme(legend.position = "none")
```

# 2. Observations by year

```{r observations-by-year, fig.width=8, fig.height=4.7}
year_counts <- germany |>
  filter(!is.na(year)) |>
  count(year, name = "observations") |>
  arrange(year)

if (nrow(year_counts) == 0) {
  plot.new()
  text(0.5, 0.5, "No usable year column was found.")
} else {
  ggplot(year_counts, aes(x = year, y = observations)) +
    geom_col(fill = "#31a354", width = 0.75) +
    geom_line(linewidth = 0.6, color = "gray25") +
    geom_point(size = 1.8, color = "gray25") +
    scale_x_continuous(breaks = pretty_breaks()) +
    scale_y_continuous(labels = comma_format()) +
    labs(
      title = "Germany observations by year",
      subtitle = "Annual counts show when the Germany subset was most strongly represented in the dataset.",
      x = "Year",
      y = "Observations"
    ) +
    theme_gtree()
}
```

# 3. Observations by source on the country map

This map shows the main data sources spatially. To keep the map readable, only the largest sources are shown separately; the rest are grouped into `Other`.

```{r source-map-data}
top_sources <- germany |>
  count(source, name = "observations", sort = TRUE) |>
  slice_head(n = 5) |>
  pull(source)

germany_sources <- germany |>
  mutate(source_group = if_else(source %in% top_sources, source, "Other"))

source_map_data <- germany_sources |>
  group_by(source_group) |>
  group_modify(~ sample_up_to(.x, max_points_per_source_map)) |>
  ungroup()
```

```{r source-map, fig.width=10, fig.height=7}
p <- ggplot()

if (has_bundeslaender) {
  p <- p +
    geom_sf(data = germany_boundary, fill = "gray97", color = "gray60", linewidth = 0.3)
}

p +
  geom_point(
    data = source_map_data,
    aes(x = longitude, y = latitude),
    alpha = 0.18,
    size = 0.45
  ) +
  facet_wrap(~ source_group) +
  coord_sf(
    xlim = c(germany_bbox["xmin"], germany_bbox["xmax"]),
    ylim = c(germany_bbox["ymin"], germany_bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "Spatial footprint of major observation sources in Germany",
    subtitle = "Each panel shows sampled observation points from one source group. Counts are computed from all rows, but points are sampled for readability.",
    caption = paste("Maximum displayed points per source group:", format_number(max_points_per_source_map))
  ) +
  theme_map_gtree()
```

```{r source-bar, fig.width=8, fig.height=5}
source_counts <- germany_sources |>
  count(source_group, name = "observations", sort = TRUE)

ggplot(source_counts, aes(x = reorder(source_group, observations), y = observations)) +
  geom_col(fill = "#756bb1") +
  coord_flip() +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "Germany observations by source group",
    subtitle = "The same source grouping is used in the source map above.",
    x = NULL,
    y = "Observations"
  ) +
  theme_gtree()
```

# 4. Observations mapped to Bundesländer

```{r bundesland-spatial-join}
if (has_bundeslaender) {
  germany_points <- sf::st_as_sf(
    germany,
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )
  
  germany_bundes_sf <- sf::st_join(
    germany_points,
    bundeslaender,
    left = TRUE
  )
  
  germany_bundes <- germany_bundes_sf |>
    sf::st_drop_geometry()
  
  bundesland_counts <- germany_bundes |>
    filter(!is.na(bundesland)) |>
    count(bundesland, name = "observations", sort = TRUE)
  
  bundesland_unknown_count <- sum(is.na(germany_bundes$bundesland))
  
  bundesland_map <- bundeslaender |>
    left_join(bundesland_counts, by = "bundesland") |>
    mutate(observations = if_else(is.na(observations), 0L, observations))
} else {
  germany_bundes <- NULL
  bundesland_counts <- NULL
  bundesland_map <- NULL
  bundesland_unknown_count <- NA_integer_
}
```

```{r bundesland-join-summary}
if (has_bundeslaender) {
  bundes_summary <- tibble::tibble(
    metric = c(
      "Germany observations joined to a Bundesland",
      "Germany observations not joined to a Bundesland",
      "Bundesländer with at least one observation"
    ),
    value = c(
      sum(!is.na(germany_bundes$bundesland)),
      bundesland_unknown_count,
      nrow(bundesland_counts)
    )
  )
  
  knitr::kable(
    bundes_summary |>
      mutate(value = format_number(value)),
    caption = "Spatial join summary for Germany observations and Bundesländer."
  )
} else {
  cat("Bundesländer boundaries were unavailable, so the spatial join could not be performed.")
}
```

## Bundesländer bar chart

```{r bundesland-bar, fig.width=8, fig.height=6}
if (!has_bundeslaender || nrow(bundesland_counts) == 0) {
  plot.new()
  text(0.5, 0.5, "Bundesländer counts are unavailable.")
} else {
  ggplot(bundesland_counts, aes(x = reorder(bundesland, observations), y = observations)) +
    geom_col(fill = "#fd8d3c") +
    coord_flip() +
    scale_y_continuous(labels = comma_format()) +
    labs(
      title = "Germany observations by Bundesland",
      subtitle = "Observation counts are based on spatially joining coordinates to GISCO NUTS1 Bundesländer boundaries.",
      x = NULL,
      y = "Observations"
    ) +
    theme_gtree()
}
```

## Bundesländer map

```{r bundesland-map, fig.width=8, fig.height=7}
if (!has_bundeslaender || is.null(bundesland_map)) {
  plot.new()
  text(0.5, 0.5, "Bundesländer map is unavailable.")
} else {
  ggplot(bundesland_map) +
    geom_sf(aes(fill = observations), color = "white", linewidth = 0.35) +
    scale_fill_gradient(low = "#fee5d9", high = "#a50f15", labels = comma_format()) +
    labs(
      title = "Observation frequency by Bundesland",
      subtitle = "Darker Bundesländer contain more Germany observations in the dataset.",
      fill = "Observations",
      caption = "Boundaries: GISCO NUTS1 regions. Counts: GlobalGeoTree Germany subset."
    ) +
    theme_map_gtree()
}
```

## Known frequency categories by Bundesland

This stacked chart uses only observations with a known `Rare`, `Common`, or `Frequent` category.

```{r bundesland-category-bar, fig.width=9, fig.height=6}
if (!has_bundeslaender || is.null(germany_bundes)) {
  plot.new()
  text(0.5, 0.5, "Bundesländer category chart is unavailable.")
} else {
  bundesland_category_counts <- germany_bundes |>
    filter(!is.na(bundesland), category %in% c("Rare", "Common", "Frequent")) |>
    count(bundesland, category, name = "observations") |>
    left_join(bundesland_counts, by = "bundesland", suffix = c("_category", "_total")) |>
    mutate(bundesland = factor(bundesland, levels = rev(bundesland_counts$bundesland)))
  
  if (nrow(bundesland_category_counts) == 0) {
    plot.new()
    text(0.5, 0.5, "No known categories are available after Bundesland joining.")
  } else {
    ggplot(bundesland_category_counts, aes(x = bundesland, y = observations_category, fill = category)) +
      geom_col() +
      coord_flip() +
      scale_y_continuous(labels = comma_format()) +
      labs(
        title = "Known frequency categories by Bundesland",
        subtitle = "Only observations matched to Rare/Common/Frequent labels are included.",
        x = NULL,
        y = "Observations with known category",
        fill = "Category"
      ) +
      theme_gtree()
  }
}
```

# 5. Observations mapped to climate zones

GlobalGeoTree does not contain climate-zone labels. To map observations to known climate zones, an external climate-zone polygon layer is required. Put a GeoJSON or Shapefile at the path configured in the setup chunk:

```text
`r climate_zones_path`
```

The file should contain polygon geometries and one column identifying the climate zone, for example `climate_zone`, `zone`, `class`, `koppen`, or `kg_class`.

```{r climate-load-and-join}
read_climate_zones <- function(path, zone_column = NULL) {
  if (!file.exists(path)) return(NULL)
  
  climate <- sf::read_sf(path)
  names(climate) <- clean_colnames(names(climate))
  climate <- sf::st_transform(climate, 4326)
  climate <- sf::st_make_valid(climate)
  
  if (!is.null(zone_column)) {
    zone_column <- clean_colnames(zone_column)
    if (!(zone_column %in% names(climate))) {
      stop("Configured climate_zone_column not found in climate file.")
    }
  } else {
    candidates <- c("climate_zone", "zone", "class", "koppen", "kg_class", "climate", "name", "label")
    zone_column <- intersect(candidates, names(climate))[1]
    
    if (is.na(zone_column)) {
      non_geometry_cols <- setdiff(names(climate), attr(climate, "sf_column"))
      zone_column <- non_geometry_cols[1]
    }
  }
  
  climate |>
    mutate(climate_zone = as.character(.data[[zone_column]])) |>
    select(climate_zone, geometry)
}

climate_zones <- tryCatch(
  read_climate_zones(climate_zones_path, climate_zone_column),
  error = function(e) NULL
)

has_climate_zones <- !is.null(climate_zones) && nrow(climate_zones) > 0

if (has_climate_zones) {
  germany_points_for_climate <- sf::st_as_sf(
    germany,
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )
  
  germany_climate_sf <- sf::st_join(germany_points_for_climate, climate_zones, left = TRUE)
  germany_climate <- sf::st_drop_geometry(germany_climate_sf)
  
  climate_counts <- germany_climate |>
    filter(!is.na(climate_zone), climate_zone != "") |>
    count(climate_zone, name = "observations", sort = TRUE)
} else {
  germany_climate <- NULL
  climate_counts <- NULL
}
```

```{r climate-status, results='asis'}
if (!has_climate_zones) {
  cat(
    "**Climate-zone layer not found.** The climate-zone section is ready, but it needs an external polygon layer at `",
    climate_zones_path,
    "`. Without this external file, the report cannot honestly assign observations to known climate zones.\n",
    sep = ""
  )
}
```

## Climate-zone distribution

```{r climate-bar, fig.width=8, fig.height=5}
if (!has_climate_zones || nrow(climate_counts) == 0) {
  plot.new()
  text(0.5, 0.5, "Climate-zone results unavailable. Provide a climate-zone polygon file to enable this section.")
} else {
  ggplot(climate_counts, aes(x = reorder(climate_zone, observations), y = observations)) +
    geom_col(fill = "#2b8cbe") +
    coord_flip() +
    scale_y_continuous(labels = comma_format()) +
    labs(
      title = "Germany observations by climate zone",
      subtitle = "Observations were spatially joined to the external climate-zone layer.",
      x = NULL,
      y = "Observations"
    ) +
    theme_gtree()
}
```

## Climate-zone map

```{r climate-map, fig.width=8, fig.height=7}
if (!has_climate_zones) {
  plot.new()
  text(0.5, 0.5, "Climate-zone map unavailable. Provide a climate-zone polygon file to enable this section.")
} else {
  climate_points <- germany |>
    sample_up_to(25000)
  
  p <- ggplot() +
    geom_sf(data = climate_zones, aes(fill = climate_zone), alpha = 0.55, color = "white", linewidth = 0.25)
  
  if (has_bundeslaender) {
    p <- p + geom_sf(data = germany_boundary, fill = NA, color = "gray25", linewidth = 0.35)
  }
  
  p +
    geom_point(
      data = climate_points,
      aes(x = longitude, y = latitude),
      alpha = 0.18,
      size = 0.4
    ) +
    coord_sf(
      xlim = c(germany_bbox["xmin"], germany_bbox["xmax"]),
      ylim = c(germany_bbox["ymin"], germany_bbox["ymax"]),
      expand = FALSE
    ) +
    labs(
      title = "Tree observations over climate zones",
      subtitle = "Climate zones come from an external polygon layer; points show sampled Germany observations.",
      fill = "Climate zone"
    ) +
    theme_map_gtree()
}
```

# Interpretation notes

The category-based visualizations must be interpreted with care. If the full dataset did not contain `category`, then `Rare`, `Common`, and `Frequent` were joined from the eval dataset. In that case, the category maps and category-by-Bundesland chart only describe the matched subset of species, not the full Germany species pool.

The Bundesländer analysis is more complete because it uses all Germany observations with valid coordinates, regardless of whether category is known.

The climate-zone analysis requires external climate-zone polygons. Without such a file, the report cannot assign observations to known climate zones from the GlobalGeoTree data alone.
